### 2014-06-03 Response to rreview.tex
updates

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Response to Referee’s report of SIM submission 13-0382 entitled "Comparing and
combing biomarkers as principle surrogates for time-to-event clinical
endpoints"}
\author{Erin E. Gabriel, Michael C. Sachs, Peter B. Gilbert}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle
We are happy to have the opportunity to revise our manuscript. We are very pleased that reviewer 1 felt that the ideas presented in our manuscript are of potential interest and value and that our theoretical developments seem solid. We apologize for the slow turn-around on this response; we waited until our cited other work, (Gabriel and Gilbert 2013), citation number [10] in the revised manuscript, had been accepted for publication. We have however attempted to make this manuscript as stand-alone as possible. We have reorganized the sections to improve readability and reduce the technical feel of the manuscript.

\section{Response to Reviewer 1}
We felt that Reviewer one had two main points and two minor points, and we hope we have addressed all in the revised manuscript.\\
\begin{itemize}
\item Main point
\begin{enumerate}
 \item In its present form, the paper is hard to appreciate as a stand alone. Moreover, the application is hard going on impossible to understand for me at least without the details apparently provided in [10]. Much of the intuition and rationale behind the ideas presented is omitted with reference to another paper [10] by two of the co-authors. This paper is to my knowledge presently not available for a general reader. In conclusion, for the paper to be readable, the authors need to provide a much more self-contained presentation of their ideas and how they in effect differ from the developments presented in [10]. In particular, we find this to be very problematic in the crucial section 2 where an in detail motivation for using F1 and F0 as the cornerstone of evaluating surrogates lacks.\\
 \textbf{We have added slightly more detail about Gabriel and Gilbert 2013,citation [10], that has now been accepted for publication in Biostatistics. This manuscript in its accepted form has been included in the revisions submission materials. In addition, we have taken care to make sure the explanation of the methods is as stand-alone as possible throughout the manuscript. We have more clearly pointed out the novel methods outlined in the manuscript in the introduction. We have also reorganized the sub-sections of the methods section to reduce the technical feel of the manuscript and make the methods easier to follow. We believe the manuscript is now much more self-contained and the advancements are easier to see and less weighed down in technical detail.\\
 We feel that the manuscript introduces a set of tools that can be used in practice to estimate the causal risk estimands and evaluate and compare univariate and multivariate candidate principle surrogates for continuous time-to-event clinical endpoints. Our proposed estimation method for the risks was previously introduced in the literature, but has only been explored for use with binary clinical endpoints. What is more, similar estimation methods have been investigated for continuous time-to-event endpoints, but only for univariate candidate surrogates. Many of the suggested summary statistics were introduced in previous works to evaluating diagnostic tests in a time-to-event setting, however they have not been suggested previously for evaluating principle surrogates. The standardized total gain was suggested in the PS literature as a means of comparison of candidate PS, but only in binary-clinical endpoint settings.}
 \item A more in detail discussion of the assumptions A1-A5 and how they provide a link between the fundamental causal quantities sought and the observational data at hand is lacking.
 \textbf{We have added what we hope to be a more clear statement about what the assumptions A1-A4 imply, directly following their outline in the revised text.}
\end{enumerate}
\item  few minor details
\begin{enumerate}
 \item page 4, line 13: perfect, \textbf{corrected, thank you}
 \item page 4, line 56: we think you’re lacking an 1 before the sum. \textbf{I have changed this equation for clarity} 
\end{enumerate}
\end{itemize}

\section{Response Reviewer 2}
\begin{enumerate}
\item The presentation is highly technical, with numerous references to your own work, some of which is not published yet. We had a very difficult time reading the manuscript without consulting the references. The paper should be written such that it can mostly be understood on its own. This should include a clarification of which parts are novel, as surrogate combinations for binary endpoints and surrogate endpoints in a time-to-event setting already seem to have been addressed in other manuscripts.\\
 \textbf{We have taken care to make sure the explanation of the methods is as stand-alone as possible throughout the manuscript. We have reorganized the sub-sections of the methods section reduce the technical feel of the manuscript and make the methods easier to follow. We believe the manuscript is now much more self-contained and less weighed down in technical detail.}

\item The application example is somewhat disappointing, as no real differences between the surrogates (and their combination) can be seen. This might either mean that there are no differences, or that the proposed tools are not adequate. You should at least present one real application, where the proposed tools are useful for identifying a promising surrogate combination.\\
\textbf{We found an error in the analysis dataset we were using. Now that we have corrected this issue the two candidate surrogates do not perform similarly, however the combination surrogate is still not an improvement over HBA1C alone. We selected these data to use as the example because HBA1C had been identified in our previous work as having surrogate value and it was hypothesized that EGFR might have some surrogate value and add to a combination surrogate of the two. EGFR was also selected because a baseline predictor could be found that was adequately correlated with both it and HBA1C. As with any real data analysis we did not know the results a priori and we do not know the true values of the coefficients. The simulations are used to show the adequacy of our proposed tools under known scenarios, we believe the example is of clinical interest and not run simply as an illustration of our methods.}

\item One specific aspect of the application example that needs to be clarified is why a combination of the surrogates can perform worse than its components. A reasonable estimation procedure might set the parameter for one of the components to a value close to zero in such a situation. Why does this not happen here? Is this potentially an instance where the Weibull model does not fit very well?\\
\textbf{The combination surrogate being estimated to have less value as a surrogate than either of the components in our example was an error that has been corrected; the same error that caused HBA1C and EGFR to perform similarly caused this issue. We should have caught this error and we apologize and thank you for your question that allowed us to find the error. Although it is no longer the case in our example, we do not rule out the that in practice one could find that the estimated STG or other summaries might be non-significantly lower for a combination surrogate than for one of its components. This may be caused by the addition of biomarkers to the risk model that are not adequately correlated with the joint BIP used or the addition of useless surrogates that only add noise to the model. As suggested, it could also be the case that the Weibull model does not fit the data well. All of these cases reflect a reduction in the precision of the estimated risks and not a true reduction in PS value. If it is believed that the Weibull model is a poor fit for the data other parametric models can be considered for the clinical outcome and can be easily accommodated by the suggested semi-parametric EML method. The reasons for a reduced estimated PS value of the summaries for a combination surrogate over its components should be investigated, but can be considered reasonable evidence of no improvement given the data available and the risk model assumed.}

\item While a naive analysis, such as separately evaluating the prognostic potential of surrogate variables separately in the treatment arms (using regression models or even Kaplan-Meier estimators) and then picking promising candidates by hand, will be obviously biased, but might be the first approach considered by practitioners. Therefore, the proposed approach should be contrasted with such a naive (biased) approach, e.g. to highlight the severity of the bias. In particular, it would be interesting to see the size of the bias relative to problems potentially introduced by the more restrictive model required by your proposal.\\
\textbf{We have performed a naive analysis of both the candidate surrogates separately in the arms of the trial using a Cox model. We find that although the ranking of the candidate surrogates would be the same based on the magnitude of the HRs in the treatment arm alone, we didn't have the power to show significance for either candidate's HR. This may be due to our use of the Cox model. In the control arm the ranking of the candidates would be the opposite, showing that HBA1C under control is less associated with outcome than is EGFR under control. The naive analysis addresses a different question than our analysis, mainly the association of the observed candidate and observed outcome in either arm of the trial. Our method evaluates the association of treatment-efficacy and the candidate surrogate under treatment. Therefore, we cannot discuss the bias of the naive analysis versus our method. Although for one of the naive analyses we find some evidence that HBA1C is a better candidate surrogate than is EGFR, this will not always be the case. The literature is full of many strong correlates of risk, biomarkers that are associated with outcome in the treatment arm alone, that are not associated with treatment efficacy and are therefore not useful PSs (Fleming and Demets, 1996). We summarize these analyses in the revised manuscript.\\
With regard to the Weibull model used, the detection of time-variation in the CEP curve is not the focus of this work, as it was in Gabriel and Gilbert (2013), and therefore the restricted model does not impede the illustration of our methods. The restricted model is used to reduce the complexity of the technical detail needed to understand the more general concepts that are proposed in the manuscript. We use the restricted Weibull model, which is actually a standard Weibull model in most settings, to help focus the manuscript on the evaluation and comparison of multivariate candidate surrogates in the time-to-event setting and our proposed summary statistics. We believe we have now made this point clear in the revised manuscript.}

\item Time-to-event endpoints sometimes are analyzed with respect to a certain time horizon, such as 1-year survival. The proposed criteria for judging surrogates reflect such a fixed horizon reasoning. However, time-to-event endpoints will often be judged over the whole course of time, e.g. using logrank tests. To reflect this, criteria for judging surrogate endpoints should not be limited to a certain time horizon. Therefore, you should extend your proposal accordingly, or at least present results for several time horizons.\\
\textbf{We looked at several different time points for the example outcomes and figures. We have added an additional figure to the supplementary materials with these new figures at 3, 4 and 9 year time points. Also, although we illustrate our methods at particular fixed time points $t$, average risk over time based on the CDF can be estimated using the same methods. We define average risk by, 
\[
\mbox{avergrisk}_{z}(s_{1,1}, \ldots, s_{1,J}) = 1/k \sum_{t=t_1}^{t_k}{risk_{z}(t|s_{1,1}, \ldots, s_{1,J})}.
\]
Average risk can then be used in any of the suggested summary curves or statistics; those statistics are then time-independent but still account for a time-to-event clinical outcome with censoring. We have added a similar statement to the manuscript.}
\end{enumerate}

\end{document}

### 2014-06-03 Compare_053014_statsmed_revisions.tex
updates

\documentclass[times, doublespace]{simauth}

\usepackage{moreverb}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
 \usepackage[numbers, square, sort&compress]{natbib}

\begin{document}
% The following line would print the thesis in 
%opening

\runninghead{EE Gabriel and MC Sachs and PB Gilbert}
\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints}
\author{Erin E. Gabriel\affil{a}\corrauth\
Michael C. Sachs\affil{b}, Peter B. Gilbert\affil{a,c}}
\address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
\affilnum{b}Kidney Research Institute, University of Washington\\
\affilnum{c}Department of Biostatistics, University of Washington}
\corraddr{egabriel@fhcrc.org}
\begin{abstract}
Principal surrogate endpoints are useful as targets for Phase I and II trials. In many recent trials, multiple post-randomization biomarkers are measured. However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information. We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate principal surrogates. We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates. These include the time-dependent and surrogate-dependent true and false positive fraction, the time-dependent and the integrated standardized total gain and the cumulative distribution function of the risk difference. We illustrate the operating characteristics of our proposed methods in simulations and outline how these statistics can be used to evaluate and compare candidate principal surrogates. We use these methods to investigate candidate surrogates in the Diabetes Control and Complications Trial.
\end{abstract}
\keywords{Causal inference; Comparison of candidate principal stratification; Survival analysis; Accuracy measures; Multivariate principal stratification}
\maketitle

\section{Introduction} 
A valid surrogate endpoint can be used as a primary outcome for evaluating treatments in phase I-II trials, by virtue of providing reliable prediction of treatment effects on a clinical endpoint of interest. In 2002, \citet{Frangakis02} introduced the principal stratification framework and a definition of a principal surrogate (PS). Several methods for evaluation of principal surrogates have been developed under this framework, eg. \citep{Taylor05, Follmann06, Gilbert08, Wolfson10, Li10, Huang11, Huang12, Zigler12, Gabriel13}.  We extend this literature in several ways.

We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate principal surrogates of continuous time-to-event clinical endpoints. Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint. \citet{Gabriel13} extended methods for evaluating univariate biomarker as a PS of continuous time-to-event clinical endpoints. \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.  Our method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.

We propose the use of several curves for displaying surrogate quality which consider treatment efficacy within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves. Evaluation of multivariate principal surrogates is complicated by the increased dimension.  Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect. The strength of this association can be displayed by the CEP curve \citep{Gilbert08}. The CEP is a contrast of risk over the treatment arms within subgroups defined by the candidate surrogate under active treatment and marginalizing over the candidate surrogate under control. The dimension of the marginal CEP function increases with the dimension of the candidate surrogate, making the marginal CEP difficult to display and interpret for multivariate candidate PS. This makes the curves two-dimensional regardless of the number of biomarkers included in the risk model. Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define. We also consider the cumulative distribution function (CDF) of the marginal CEP function. Time-dependent TPF and FPF curves based on the levels of diagnostic tests and time-independent curves based on levels of causal contrasts, \citep{Huang12b}, have been proposed in previous works. However, no version of the TPF or FPF curves have been proposed for the evaluation or comparison of candidate PSs, nor has the causal-contrast-based and time-dependent versions of the curves been defined.

We develop summary statistics for the comparison of candidate PS for a continuous time-to-event end point. Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS . Often evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality. We develop two extensions of the standard total gain \citet{Huang11}, both of which provide a single values measure of surrogate quality that can be used to compare candidate surrogates for the same clinical output in the same trial. The time-dependent standard total gain, ($STG(t)$), allows for a single value comparison of marginal CEP models as well as inference for comparison of candidate PS that summaries all information and properly accounts for censoring at and given time point $t$. Over a range of time points, a display of $STG(t)$ versus $t$ curve provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time. The time integrated  standard total gain, ($widetilde{STG}$), which is the time-dependent standard total gain integrated over the estimated survive times, provides a single value summary of surrogate quality independent of time.

In Section 2, we outline our notation. In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PSs or the comparison of candidate PSs.  In Section 3.1, we describe the time-dependent risk estimands and outline the assumptions that are helpful for identifying these estimands.  We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.  We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison. In Section 3.7 we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}. In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PSs. In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT). We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combination of the two candidates is not an improvement over the higher quality surrogate alone. In Section 6, we discuss potential limitations of our methods and future avenues for research.

\section{Notation and Criterion for Principal Surrogacy} \label{DEF}
Let $Z$ be the treatment indicator, 0 for control/non-active treatment and 1 for treatment. Let $W$ be a vector of baseline measurements taken prior to randomization. In the principal stratification framework of \citet{Frangakis02} we use potential outcomes, where all post-randomization measures are considered under assignment to either treatment arm for each individual. Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z=\{0,1\}$. Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time. Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z=\{0,1\}$. Let $Y_{i}^{t}(z)$ be the indicator that $T_i(z)\leq t$. Let $D_i(t)=Y_{i}^{t}(0)-Y_{i}^{t}(1)$ be the individual treatment effect for subject $i$ on the clinical endpoint at or before time $t$. Let $\{S_{1,i}\ldots,S_{J,i}, T_i, C_i, Y_i, Y_{i}^{t} \}$ denote the observed values of the potential measures for subject $i$.

All $S_j$ are measured at the same fixed time point $\tau$ after randomization, and this time point, post-randomization but prior to clinical-event, and z define the surrogate measurement. If $T_{j,i}(z)$ is less than $\tau$ for a given subject $i$, all $S_{j,i}(z)$ are undefined for that subject, as observation of the post-randomization pre-clinical-event measures $S_{j}(z)$ are no longer possible under either treatment condition. For this reason, subjects with min$(T,C) < \tau$ are excluded from the evaluation cohort. Let $M_j$ be the indicator that $S_{j}(1)$ is observed, and let $s_{1,j,i}$ be the realization of $S_{j}(1)$ for subject $i$. We assume the full set of observed and counterfactual outcomes are independently and identically distributed over the trial subjects $i= 1, \ldots n$. Let $F_{S_1(1), \ldots, S_J(1)}(\cdot,\dots,\cdot)$ and $F_{S_1(1), \ldots, S_J(1)|W}(\cdot|\cdot)$ be the joint CDF of the $S_1(1), \ldots, S_J(1)$ and joint CDF of the $S_1(1), \ldots, S_J(1)$ conditional on $W$, respectively. Define $\rho_z(t)=Pr(Y^t(z)=1)=E(Y^t(z))=Pr(T(z)\leq t)$ to be the marginal prevalence of the potential clinical outcome at or before time $t$ for treatment arm $z$. Our risk estimands of interest, $risk_z(s)$, are functions of the candidate surrogate that give some measure of risk under the treatment $z$.

\citet{Frangakis02} developed the concept of principal surrogacy and defined a PS by a criterion based on joint risk estimands, which conditions on the candidate surrogate under active treatment and under control, $risk_z(s_1,s_0)$. If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN). Subsequent work added a second criterion that a good PS should modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, what \citet{Gilbert08} called average causal sufficiency ACS. This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.  ACN is a strict but sensible necessary condition, which guarantees that there is no effect on the clinical out when there is no effect on the PS. However, it is clearly not a sufficient condition, as any biomarker in a trial with no clinical treatment effect is likely to satisfy ACN.  The combination of ACS and ACN are often looked evaluated in the current literature, with some works relaxing strict ACN.  ACS also suggests a possibly ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate. Strict ACN alone classify candidates as either PS or not PS, with no other indication of quality.

In \citet{Gilbert08}  and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate; marginal risk estimands condition on $S(1)$ and marginalizing over the distribution of $S(0)$. The concept of ACS is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$. This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect. Although in some special cases when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN can not be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}.  Often the concept of ACN is relaxed in practice, such that if the CEP of interest is equal to zero when $s_1=0$, ACN is said to be likely. Although this is similar to ANC, this is not the same as ANC unless CB holds and $S(0)=0$. \citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in general clinical trial where biomarkers under placebo will not tend to be constant.

For a univariate candidate PS strict or relaxed ACN can be displayed using the joint CEP or the marginal CEP.  Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically. As well, the definition does not seem as sensible in the multi-dimensional case. Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates. However, it seems like you would want less clinical treatment effect if there was no treatment effect on any of the biomarkers.  We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional of a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$. For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections 3.3 and 3.4. Like ACN, the classification accuracy measures allow for the quantification of the agreement between treatment effect on the clinical outcome and predictive model when there is no treatment effect on any of the biomarkers in the model.  Unlike ANC, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.  These accuracy measures can also be displayed as curves vs. the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model. We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary for a set of biomarkers to have value as a PS, but that this is not sufficient. Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.

ACS for a multivariate PS is wide variation in the CEP over subgroups defined by the set of biomarkers. This is difficult to display graphically due to the increased dimension of the surrogate. We propose the use of functions based on the CEP rather than the set of biomarker values for display of ACS. The propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS. We also suggest a summary statistic standardized total gain that quantifies both our suggested accuracy concept and ACS in a single measurement. Details for standard total gain are given in Section \ref{STG}.  As well, if one assumes a parametric a model for risk conditional on a set of biomarkers a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.

After a set of candidate multivariate or univariate candidate surrogates is determine to have some value as PS, it is of interest to rank them and test to determine if any one candidate is significantly better than others. ANC can only classify candidates into two groups. Our suggested accuracy measures and the CDF of the risk model CEP can help more finely group candidates by predictive quality and amount of variation in CEP, but they do not provide inference for comparing candidates. We propose the use of standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for comparison inference. These are outlined in Sections \ref{STG} and \ref{ISTG}.

\section{Methods}
\subsection{Time-Dependent Causal Risk Estimands} \label{CRE}
In the time-to-event setting there are many ways to define risk. The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF. The conditional CDF-based risk estimands can by defined as:
\begin{eqnarray*}
risk_{1}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_1(t|s_{1,1}, \ldots, s_{1,J})\\
&\equiv& P\{T(1) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{1}(1)=s_{1,J},T(1)> \tau, T(0)> \tau\},\\
risk_{0}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_0(t|s_{1,1}, \ldots, s_{1,J}) \\
&\equiv& P\{T(0) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{J}(1)=s_{1,J},T(1)>\tau, T(0)>\tau\}.
\end{eqnarray*}
The subscript of a function indicates the value of $z$. The comparison between the risk estimands over the treatment arms is the causal comparison of interest for principal surrogate evaluation. Our contrast of interest is defined as $\Delta(t|s_{1,1}, \dots, s_{1,J}) \equiv risk_0(t|s_{1,1}, \ldots, s_{1,J})-risk_1(t|s_{1,1}, \ldots, s_{1,J})$, the risk difference. Using the CDF based risk estimands, this contrast is causal because both risks condition on the same sets. This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}. For this reason, we focus on the CDF based marginal risks, which provide well defined  interpretations for our suggested summary statistics.

In order to link these risk estimands to the observable data we must make some assumptions; Assumptions A1-A4 reduce the number of missing potential outcomes and help identify the risk estimands.
\begin{itemize}
\item A1: Stable Unit Treatment Value Assumption (SUTVA) and Consistency 
\item A2: Ignorable Treatment Assignment
\item A3: Equal individual clinical risk up to time $\tau$: $T(1)\leq\tau$ if and only if $T(0)\leq\tau$
\item A4: Random censoring: $T(z) \perp C(z)$ for $z=\{0,1\}$.
\end{itemize} 
Assumption A1 implies that subject's potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency. For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$. SUTVA may not hold is some situations, such as cluster randomized vaccine trials where herd immunity may change the treatment effect on all outcomes in the cluster.  Assumption A2 implies that the distribution of the potential outcomes S(z) and Y(z) are independent of the observed treatment assignment. Assumption A2, is generally true in randomized, blinded trials.  Assumption A3 aides in the identification of subjects with min$(T,C) < \tau$ to be excluded from the evaluation cohort without causing bias. Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced events or drop-out over the trial arms prior to $\tau$. Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used. Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z=\{0,1\}$.

Given assumptions A1-A4 we can link the time-dependent potential outcome risk estimands of interest to observed data clinical risk estimands and the partially observed candidate surrogates by assuming a parametric survival model. We will assume a Weibull model, the probability density function (pdf) for which we will denote as $g(\cdot|\cdot,\cdot)$. We parameterize the scale term of this model, $\gamma=(\gamma_{00}, \gamma_{10},\gamma_{0j}, \gamma_{1j})$, and allow for a constant shape term, $\beta_{0}$; this parameterization is equivalent to a classical Weibull survival model. The density is given by:
\begin{eqnarray*}
g(t|\gamma, \beta, z, s_{1,1}, \ldots,s_{1,J}, y)=\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})^{y} Q_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J}),
\end{eqnarray*}
where $Q_z(t|\gamma, \beta, s_{1,j})$ is the conditional Weibull survivor function for treatment arm $Z=z$ and $\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})$ is the conditional Weibull hazard function. We assume the risk estimands are of the form:
\begin{itemize}
\item A5: $risk_{z}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) \equiv 1-\exp(-(t/\exp(\gamma_{z0}+\sum_j{\gamma_{zj}s_{1,j})})^{(\exp(\beta_{0}))}$
\end{itemize}
for $z=\{0,1\}$. We focus on this classical model in the main text for clarity; more complex time-dependent models are outlined in Appendix B of the supplementary materials. In order for both assumptions A3 and A5 to hold simultaneously, we can only considered censoring or event times greater than $\tau$. This can be accomplished in practice by rescaling all follow-up times from $\tau$ rather than from enrollment.

One can test the null hypothesis of no surrogate value or no ACS, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$. If any of the $S_{J}(1)$ associated terms significantly differ from zero there is evidence of variation in the risk difference over the candidate PS.

\subsection{Trial Augmentation and Estimation} \label{EST}
The assumed parametric form of the risk estimands, A5, condition on the $S_{j}(1)$, which are missing for all subjects assigned non-active treatment $z=0$. Several papers have suggested the use of baseline covariates to account for the missing $S_{j}(1)$ values, a strategy called the ``baseline immunogenicity predictor'' (BIP) augmented trial design \citep{Follmann06, Gilbert08}. A useful BIP, $W$, is highly correlated with the $S_{j}(1)$ which allows for good prediction of the missing $S_i(1)$. Although other trial augmentations have been proposed in the PS literature, the only augmentation available in our motivating data set is BIP, so it will be our focus. Using the BIP trial augmentation the observed likelihood given Assumption A5 is:
\begin{eqnarray*}
&&\mbox{L}(\beta,\gamma,\nu) \equiv \prod^{n}_i\left\{g(T_i|\gamma, \beta, Z_i, s_{1,1,i},\ldots,s_{1,J,i}, Y_i)\right\}^{\prod_{j=1}^{J}{M_{j,i}}}\\ 
&&\left\{\int g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, s_{1,l+1},\ldots, s_{1,J}, Y_i)dF_{S_{l+1}(1),\ldots,S_{J}(1)|W}(s_{1,l+1},\ldots,s_{1,J})\right\}^{\{\prod_{j=1}^{l}{M_{j,i}}*\prod_{j=l+1}^{J}{(1-M_{j,i})}\}}\\ 
&&\left\{\int g(T_i|\gamma, \beta, Z_i,s_{1,1}, \ldots, s_{1,J}, Y_i)dF_{S_{1}(1),\ldots, S_{J}(1)|W}(s_{1,1},\ldots, s_{1,J})\right\}^{\{\prod_{j=1}^{J}{(1-M_{j,i})}\}}.
\end{eqnarray*}

This likelihood could be solved by direct optimization if we assumed a parametric model for $S_j(1)|W$. However, as proposed in \citet{Pepe91}, consistent estimation of outcome model parameters can be achieved using an estimated distribution of $S_j(1)|W$; this is called estimated maximum likelihood. Following \citet{Huang11}, we assume the semi-parametric location-scale model of \citet{Heagerty99} for $F_{S_{j}(1)|W}$. Considering first a univariate candidate surrogate, the semi-parametric model is defined as $F_{S_{j}(1)|W} \sim F_j[\{s_{1,j} -\mu_j(w)\} / \sigma_j(w)]= F_j(\varsigma_j)$, where $F_j$ is the CDF of the univariate residuals $\varsigma_j$ from fitting the location and scale parameters $\mu_j(w)$ and $\sigma_j(w)$, respectively.  We define $\mu_j(w)$ and $\sigma_j(w)$ as parametric functions of the baseline variable(s) $W$, $\mu_j(w) = \gamma_j'w$ and $\log\{\sigma_j(w)\} = \eta_j'w$. We can estimate these parameters by solving the equations given in \citet{Heagerty99}:
\begin{eqnarray}
&&\sum_{k=1}^{n_V} {w_{k} (s_{(1,j,k)} - \gamma_j' w_{k} )\over \{\exp(\eta_j'w_k)\}^2} =0\\ \label{stuff}
&&\sum_{k=1}^{n_V} {w_{k}\{(s_{(1,j,k)} - \gamma_j' w_{k})^2 - \{\exp(\eta_j'w_k)\}^2\} \over \{\exp(\eta_j'w_k)\}^2}=0,\label{stuff2}
\end{eqnarray}
among those treated subjects with all $S(1)_{(j,k)}$ and $W$ measured. We denote the number of subjects in this validation group by $n_V$.

Fitting the set of parameter equations for the validation subjects yields $n_V$ residuals, $\varsigma_{j,k} \equiv S_{j,k}(1) - \hat{\gamma_j}'w_{k} + \exp(\hat{\eta_j}'w_{k})$. Missing counterfactual $S_{j}(1)$ values can be imputed using the parameter estimates and the $n_V$ residuals by, $S_{j,i,k}^*(1) = \hat{\gamma_j}'w_{i} + \exp(\hat{\eta_j}'w_{i}) \varsigma_{j,k}$ for $n_V$ total imputations for each missing $S_{j}(1)$ value. We can then use these imputed values $S_{j,i,k}^*(1)$ to estimate the EML contribution for subject $i$ by $\int g(t_i|\gamma, \beta, z_i, s_{1,1}, \ldots, s_{1,J}, y_i)dF_{S_{1}(1),\ldots,S_{J}(1)|W}(s_{1,1},\ldots,s_{1,J}|W)$, by the empirical integral ${\left(1/n_{V}\right)}\sum_k^{n_V}{g_z(t_i|\gamma, \beta,z_i, S_{1,k}^*(1), \ldots,S_{J,k}^*(1),y_i)}$. missingness in the treatment arm, such as subsampling, case-cohort sampling and unplanned missing at random variables, can be inputed in the same way. In cases were the validation set is not a random sample of the treatment subjects, the location-scale models can be weighted to accommodate the bias sampling.

To accommodate multivariate surrogates, the same location-scale model is fit for each of the univariate candidate PS separately in the set of the validation subjects. Imputed values for each missing biomarker are produced separately, but provided that the same baseline variables are used in each model empirical integration over the set of all imputations is approximately equivalent to integrating over the joint distribution of biomarkers given the baseline variables used for imputation. This is an important limitation of this method. The BIP need not be based on a single baseline measurement or have the same correlation with each candidate, but the BIP must be the same in each location-scale model for the estimated likelihood to be proper using our procedure. This is not a requirement to compare univariate candidate surrogates not included in the same model. This gives us an estimated log likelihood of:
\begin{eqnarray*}
&&\mathit{l}(\beta,\gamma, \hat{\nu})=\sum_{i} log(g_z(T_i|\gamma, \beta,Z_i, s_{1,1,i}, \ldots,s_{1,j,i},Y_i))*\left\{\prod_{j=1}^{J}{M
_{j,i}}\right\}+\\
&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, S^{*}_{l+1,i,k}(1),\ldots, S^{*}_{J,i,k}(1), Y_i))}\right\}*\\
&&\left\{\prod_{j=1}^{l}{M_{j,i}}*\prod_{j=l+1}^{J}{(1-M_{j,i})}\right\}+\\
&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g_z(T_i|\gamma, \beta,Z_i, S_{1,i,k}^*(1), \ldots,S_{j,i,k}^*(1),Y_i))}\right\}*\left\{\prod_{j=1}^{J}{(1-M
_{j,i})}\right\}\cdot
\end{eqnarray*}

The asymptotic distributional results of \citet{Pepe91} for general EML estimators do not carry over to our setting, due to the zero probability of observing $S_{j}(1)$ in placebo recipients. We suggest the bootstrap for inference. The assumed form of the risk estimates, A5, can be estimated using the parameters from solving the EML and the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, can be estimated directly from the risk estimates.

\subsection{Classification Accuracy Measures} \label{CAM}
The time-dependent sensitivity and specificity for the risk difference conditional on $S_{1}(1), \dots, S_{J}(1)$, $\Delta(t|s_{1,1}, \dots, s_{1,J})$, can be defined by adapting \citet{Heagerty00} as:
\begin{eqnarray*}
\mbox{Sensitivity}(t|c)&=&P(\Delta(t|s_{1,1}, \dots, s_{1,J}) > c|D(t)=1)=TPF(t|c),\\
\mbox{Specificity}(t|c)&=&P(\Delta(t|s_{1,1}, \dots, s_{1,J}) \leq c|D(t)=0)=1-FPF(t|c).
\end{eqnarray*}
Here, $TPF(t|c)$ and $FPF(t|c)$ are the time-dependent true and false positive fractions. Sensitivity and true positive fraction represent the probability that we predict a subject will have benefit using the model and some threshold $c$, $\Delta(t|s_{1,1}, \dots, s_{1,J}) > c$, given that we observe that they did have benefit, $D(t)=1$. Similarly, specificity represents the probability that we predict a subject will have no benefit based on our model and using some threshold $c$, $\Delta(t|s_{1,1}, \dots, s_{1,J}) < c$, given that we observe that they did not have benefit, $D(t)=0$. This is one minus the false positive fraction, which is the probability that we falsely predict that a subject will have benefit, when they truly did not. These probabilities over all the thresholds, demonstrate how well our surrogate based model predicts our outcome of interest, benefit from treatment, and are therefore a useful illustration of the usefulness of the surrogate to predict outcome given the assumed model. Figure 1 of Appendix A of supplementary materials illustrate the meaning of these classification accuracy measures.

In this case, sensitivity and specificity are based on unobservable data and cannot be estimated empirically. However, under assumptions A1-A6, we can estimate them. Assumption A6 is needed to identify the classification accuracy measures, but monotonicity does not have to imply that the treatment is never harmful and in some cases the alternative assumption that the treatment never helps is more likely. Assumption A6 is not fully testable, although it can be rejected. Given Assumptions A1-A5, there are clear testable implications of Assumption A6, such as $P(T\leq t|Z=1)\geq P(T\leq t|Z=0)$ for all $t$ if no harm of treatment is assumed, $P(T \leq t|Z=0) \geq P(T\leq t|Z=1)$ for all $t$ if no benefit of treatment is assumed, and no switch in sign of the model-based estimate of risk difference over the range of the observed values of the $S_j(1)$s.

Assumption A6 may be more plausible in settings where the outcome is a specific illness or infection rather than death from any cause. We assume without loss of generality the no-harm of active treatment form of monotonicity heretofore, $T(0) \leq T(1)$, as this is the form of the assumption that is more likely to apply to our motivating example. Assuming A1-A6, $TPF(t|c)$ can be defined as:
\begin{eqnarray*}
TPF(t|c)&=&{E[\Delta(t|S_{1,1}, \dots, S_{1,J})\}\mathrm{I}_{\{\Delta(t|S_{1,1}, \dots, S_{1,J})>c\}}] \over E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\}}.
\end{eqnarray*}
The derivation of this form is given in Appendix A of the supplementary materials; both the definition and derivation are adaptations of time-independent curves given in \citet{Huang12b}. The function $FPR(t|c)$ can be defined similarly based on the expectation of $1-\Delta(t|s_{1,1}, \dots, s_{1,J})$. One can estimate these accuracy measures using the plug-in estimators:
\begin{eqnarray*}
\widehat{TPF}(t|c)&=& {\int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})} \hspace{2mm}\mbox{and},\\
\widehat{FPF}(t|c)&=& {\int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}.\\
\end{eqnarray*}
Here $\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})$ is a maximum likelihood estimate under A1-A6 and using the estimated coefficients from the assumed model A5.

If the model for risk difference that conditions on $\{s_{1,1}, \ldots, s_{1,J}\}$, indicates there is a high probability of having a risk difference above a given threshold for all subjects that failed on non-active treatment but not on active treatment at or before time $t$, this is evidence of the predictive power of the model. The summary $TPF(t|c)$ is therefore a good measure for the comparison or evaluation of candidate PS. Similarly, $\mbox{Specificity}(t|c)$ or $1-FPF(t|c)$, is of interest as those subjects not helped by treatment should have a high probability of having a small risk difference based on a predictive model. As these measures are threshold dependent, they can be used to compare not just candidate surrogates but thresholds for a given surrogate. Thresholds $c$ found to have high $TPF(t|c)$ and low $FPF(t|c)$ are ideal.

It is of interest to determine the relationship between $TPF(t|c)$ and $FPF(t|c)$. The time-dependent ROC curve gives an illustration of the $TPF(t|c)$ and $FPF(t|c)$ relationship at given levels of $c$. This curve is defined by $ROC^{t}(q)=TPF(t|\{(FPF)^{-1}(t|q)\})$ for a given time point $t$, adapting the binary-clinical endpoint $ROC(q)$ definition of \citet{Huang12b}. Figure \ref{4panel} panels B and D displays the $ROC^{t}(q)$ curves for several different surrogate quality levels. The most left-upper point on the $ROC^{t}(q)$ curve is of interest for comparison of candidate surrogates, as candidates with left-upper points closer to $(0,1)$ will tend to be better PS, as elaborated in Section \ref{STG}.
\begin{sidewaysfigure}[H]
\begin{center}
%\includegraphics[scale=1.75, angle=270, clip=TRUE, trim=0 0 20mm 0]{4Panel_trueTE.pdf}
\includegraphics[scale=0.80]{Figure1.jpeg}
\end{center}
\caption{Panel A displays the CDF of the risk difference based on a useless and an ideal univariate surrogate. Panel B displays the ROC curves for the same ideal and useless surrogate candidates. Panel C displays the CDF of risk difference based on a medium quality univariate partial surrogate, and the CDF of risk difference based on a multivariate surrogate, which is the combination of two medium quality univariate partial surrogates. Panel D displays the ROC curves for the same medium quality univariate partial surrogate and multivariate surrogate. \label{4panel}}
\end{sidewaysfigure}

When A6 is observably violated the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased. However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities. Greater discussion of this point is given the Appendix A of the supplementary materials. When there are obvious violations of A6 the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued. When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation or candidate surrogates.

\subsection{Cumulative Distribution Function of $\Delta$} \label{CDF}
 Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PSs and for the comparison of candidate PS. An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$. If we assume monotonicity of treatment effect, formally stated as
\begin{itemize}
\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z=\{0,1\}$,
\end{itemize}
then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$). If we alternatively assume no benefit, then the line will be at the proportion of subjects with $D(t)=-1$. Assumption A6 is not needed for the function $CDF^{t}_{\Delta}(c)$ to be well-defined. For a useless surrogate, the plot of $CDF^{t}_{\Delta}(c)$ versus $c$ will be a vertical line at $\rho_0(t)-\rho_1(t)$, the difference in prevalence of the potential clinical outcomes between the trial arms. Partially useful surrogates will have $CDF^{t}_{\Delta}(c)$ versus $c$ curves between these two curves with better surrogates having flatter curves. Figure \ref{4panel} displays the CDF of the risk difference for several different surrogate quality levels. We suggest the plug-in estimator of $\widehat{CDF}^{t}_{\Delta}(c)$, $\widehat{CDF}^{t}_{\Delta}(c)=\int{\mathrm{I}_{[\hat{\Delta}(t|s_{1,1,i}, \dots, s_{1,j,i})\leq c]} d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}$. Although we display a useless surrogate in Figure \ref{4panel}, we suggest testing for any surrogate value prior to estimation of $\widehat{CDF}^{t}_{\Delta}(c)$ as it was found in simulations that there is evidence of bias when there is no variation in risk difference; a suggested test is given in Section \ref{CRE}.

Let $R^t(v)$ be the $v$th quantile of $\Delta(t|s_{1,1}, \dots, s_{1,J})$, where $R^t(v)=c$ for $v\in(0,1)$ such that $CDF^{t}_{\Delta}(c)=\{R^t\}^{-1}(v)$. The $R^t(v)$ versus $v$ curve has been called the predictiveness curve for the risk difference model. This curve has been previously suggested as a means of candidate PS comparison and evaluation \citep{Gilbert08}. The range of the curve is restricted by the range of risk difference, which is determined by the prevalence of the potential clinical events of interest in the population. Therefore, we prefer the $CDF^{t}_{\Delta}(c)$ versus $c$ curve for displaying the proportion of those subjects for which the treatment was ineffective at a given level of estimated risk difference. For characterization of the predictiveness of the model we prefer the classification accuracy measures.

\subsection{Time-dependent Standardized Total Gain} \label{STG}
It is easily shown that the area under the predictiveness curve, $R^t(v)$ versus $v$, is equal to $\rho_0(t)-\rho_1(t)$. For this reason, the area between the quantile curve $R^t(v)$ and the line $\rho_0(t)-\rho_1(t)$ on a plot against $v$ is of interest for evaluation and comparison of candidate PS. This area has been called the total gain (TG). \citet{Huang11} suggest a standardized version of TG for the comparison of PS for binary clinical endpoints. We extend the concept of total gain to the time-to-event setting and define the time-dependent total gain as $TG(t) = \int_0^{1}{|R^t(v) -(\rho_0(t) -\rho_1(t))|dv}$ and the time-dependent standardized total gain as $STG(t) \equiv TG(t)/[2(\rho_0(t) - \rho_1(t))\{1-\rho_0(t) + \rho_1(t)\}]$.

The $STG(t)$ is of interest for comparison of candidate surrogates regardless of the validity of assumption A6 as steeper quantile curves, and thus greater area between $R^t(v)$ and $(\rho_0(t) -\rho_1(t))$, correspond to better prediction of clinical treatment effect and a better candidate surrogate. Under A6, $STG(t)$ also has a useful interpretation based on the classification accuracy measures, given by $STG(t)=max_c\{\mbox{Sensitivity}(t|c) + \mbox{Specificity}(t|c)\} -1$. This interpretation follows directly from the proof given in \citet{Huang11} for the binary clinical endpoint setting. Given this interpretation, $STG(t)$ at a fixed time point is a useful single measurement for evaluation of univariate and multivariate candidate surrogates, as this is equivalent to the upper-most left point on the $ROC^{t}(q)$ curve for the same time point $t$.

We suggest plug-in estimators for $TG(t)$ and $STG(t)$ given by, 
\begin{eqnarray}
\widehat{TG}(t)&=&\int{|\widehat{\Delta}(t|s_{1,1}, \dots, s_{1,J}) -\{\hat{\rho}_0(t) -\hat{\rho}_1(t)\}|d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})} \hspace{4mm} \mbox{and} \label{TG} \\
\widehat{STG}(t)&=&\widehat{TG}(t)/[2(\hat{\rho}_0(t) - \hat{\rho}_1(t)]\{1-\hat{\rho}_0(t) + \hat{\rho}_1(t)\}.
\end{eqnarray}
Here $\{\hat{\rho}_0(t) -\hat{\rho}_1(t)\}$ is a maximum likelihood estimate. Even when assumption A6 does not hold, the difference in estimated $STG(t)$ can be used for inference to compare candidate PS in the same trial at a particular time point $t$, by Wald test of the null hypothesis $STG_1(t)-STG_2(t)=0$.

As well, the same estimator we use at a fixed time point $t$ can be used to estimate the $STG(t)$ versus $t$ curve, for all $t$ greater than $\tau$ and less than the time of trial closure. This curve can be used to investigate time-variation in surrogate quality for a particular surrogate or for comparison. Curves, $STG(t)$ versus $t$, that have noticeable downward slopes may indicate that a candidate surrogate modifies the clinical treatment effect less over time, declining surrogate quality, or that the marginal risk difference is declining. To better understand the context of declining or increasing $STG(t)$ over time, the marginal risk differences over time should be graphed simultaneously. Figure \ref{STG} Panel A depicts a time-independent high quality surrogate and useless surrogate in a trial with declining marginal risk difference, Panel B depicts the same surrogates for a trial with increasing marginal risk difference.

\subsection{Integrated Standardized Total Gain} \label{ISTG}
As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times. We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the distribution function of the event times. The estimated marginal distribution of event times $\hat{F}(T)$ is calculated by integrating over $Z$ and $S$ in the conditional CDF defined by the assumption A5. Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
\[
 \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
\]
The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point. By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates. The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.

\subsection{Evaluation and Comparison of Candidate PS} \label{DEF}
Testing null hypothesis $H0_1$,  no surrogate value, using the biomarker associated coefficients from the risk estimation model as outlined in section \ref{CRE} is a first step in evaluation or comparing biomarkers as PS.  This test can also be used to help determine what biomarkers to include in a multivariate surrogate. One should first consider each biomarker as a univariate surrogate, eliminating those from consideration that are found to have no evidence of variation in the CEP-curve by $H0_1$. Then, combinations of the biomarkers can be considered as multivariate candidate surrogates, giving greater consideration to combinations of univariate candidates that were found to have some evidence of surrogate quality. The coefficient test of  null hypothesis $H0_1$ can again be used to reduce the dimension of multivariate surrogates, as a biomarker $j$ for which the coefficients $\gamma_{0j}=\gamma_{1j}=0$ in a multivariate surrogate can be considered to not add value to surrogate. Once the field of consideration has been narrowed to considered only candidates with at least some surrogate quality, PS can be further evaluated based on their $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ and the $STG(t)$ v $t$  curves. These curves can also help to give a visualization of a ranking of the candidates.

The $STG(t)$ or $\widetilde{STG}$ difference between each of the candidates that are evaluated to have similarly high value as PS can be used to determine if there is statistical evidence to support the superiority of one of the candidates. The choice of time-point for comparison should be motivated by the scientific question of interest. For example, If one year treatment efficacy is of primary interest, a $t$ of one year less $\tau$ would be a logical choice from comparison. If there is no time point of particular scientific interest, inference should be based on $\widetilde{STG}$ difference and $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ should be considered at the longest follow-up time less $\tau$. As the CDF based risk is cumulative, a $t$ that is equal to the longest follow-up less $\tau$ includes all the available outcome information and allows for the best investigation of durability both of surrogacy and treatment effect. Comparisons over all observed follow-up times greater than $\tau$ can be investigated using the $STG(t)$ versus $t$ plots.

Ideally there will be a single candidate surrogate that has visually superior $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ that also has statistically significantly higher $\widetilde{STG}$ than all other candidates. This may not always be the case. When assumption A6 holds the candidate surrogate with the highest left-most point of the $ROC^{t}(q)$ curve should also be the candidate surrogate with the highest $STG(t)$. When A6 is violated this may not be true. When A6 does not hold, the candidate surrogate with the highest $\widetilde{STG}$ and a $CDF^{t}_{\Delta}(c)$ curve that is superior to all other candidates for at least some $c$, has the most value as a PS among the candidates considered. If this candidate is found to be statistically significantly superior to some or all the other candidates this provides greater evidence of superiority. However, as with almost all statistical tests, lack of significant reject of the null does not prove the null; lack of significant  $\widetilde{STG}$ difference does not prove lack of superiority.

\section{Simulation Studies}
Suppose the conditional CDF of $T$ given $S_j(1)$ and $Z$ follows a Weibull model and $\{S_j(1),W\}$ follows a multivariate truncated normal model with correlation R$_{S_jW}$ between each of the univariate candidates and the BIP and correlation R$_{S_kS_j}$ between each pair of candidates. Information lost to drop out occurs completely at random and at a rate of approximately 5\% per year. Event times are censored at 3 years after $\tau$, at which time the trials have 50\% cumulative treatment efficacy on average. An average of 103 treatment arm infections are observed by 3 years post $\tau$, corresponding to an annual incidence rate of 0.06, whereas an average of 206 control arm infections are observed in the same period, for an annual incidence rate of 0.12 over the 500 simulations.

We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$. We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $STG(3)$ of 0.23 and 0.51, respectively. We investigate these different levels of candidate surrogates for two different BIP correlations, $R_{s_{1j}w}$, of 0.8 and 0.5. For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0.8), (0.5, 0.5)\}$ with a correlation between the candidate surrogates ranging from 0 to 0.3. Results for the 0.3 correlation are displayed. 500 simulation were run for each scenario; some simulations did not converge and were removed from the simulations used for bias and power calculations, there was a 2.5\% non-convergence rate  All candidate PS in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.

Bias results are displayed in Table \ref{Bias} and suggest that the semi-parametric EML method for the Weibull risk difference model is adequately unbiased and that the suggested estimators for $CDF^{t}_{\Delta}(c)$, $TPF(t|c)$, $FPF(t|c)$, $STG(t)$ and $\widetilde{STG}$ have adequate precision and accuracy in all cases for which they are applicable. As we have suggested above, there is some bias when there is no surrogate value. When there is no surrogate value the estimated model coefficients are still unbiased, with average bias less than 3\% of one Monte Carlo SD (results not displayed). The large standard deviation of the $STG(t)$ estimates over all the simulations may imply that this estimator is not as efficient as possible. A correlation of 0.5 or higher between a candidate PS and a BIP seems adequate for unbiased estimation of the suggested summary statistics and curves. The results for the joint Wald test of any surrogate value, displayed in Table \ref{Bias}, suggest that this test has correct Type 1 error and adequate power in all scenarios. Monte Carlo standard errors are used in all Wald tests due to the computational burden of the bootstrap.

Table \ref{Power} displays the power to discriminate between two surrogate quality levels based on the $STG(t)$ difference Wald test for fixed time point $t=3$. Monte Carlo standard errors are used and only those candidate surrogates with any surrogate quality are displayed. We find that for all null hypothesis scenarios the test has correct type 1 error. We also find that the test has adequate power to detect differences in $STG(3)$, particular when there is good correlation between the BIP and the candidate PS, with the power increasing with the difference in $STG(3)$, as expected. The test is not valid if either of the candidate surrogates has no surrogate value, as type 1 error is inflated when the true value is on the boundary of the parameter space (results not shown).

Table \ref{Ipow} displays the power to discriminate between two surrogate quality levels based on the difference in integrated STG, $\widetilde{STG}$. We again find that for all null hypothesis scenarios the test has nearly a nominal 5\% Type 1 error rate with similar to slightly higher power than the $STG(3)$ difference test for the same scenarios.

\begin{sidewaystable}[ht]
\begin{center}
\caption{Performance of Accuracy Measures and the Joint Wald Test of the Null Hypothesis of No Surrogate Value H0$_1$ \label{Bias}}
\begin{tabular}{lccccccc}
  \hline
&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$&$STG(3)^{*}=0.23$&$STG(3)^{*}=0.51$\\ 
  \hline
Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0 (0.009) & 0.006 (0.011) \\ 
 &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & -0.001 (0.009) & 0.025 (0.031) \\ 
Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058) & -0.024 (0.065) & 0.013 (0.048) & 0.063 (0.139) & 0.023 (0.037) \\ 
 &0.5& 0.56 (0.048) & 0.028 (0.07) & -0.006 (0.075) & 0.011 (0.047) & 0.062 (0.128) & 0.013 (0.072) \\ 
Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002) & 0.001 (0.003) & 0 (0.001) & -0.004 (0.01) & -0.004 (0.006) \\ 
 &0.5& 0.497 (0.002) & -0.001 (0.003) & 0 (0.003) & -0.001 (0.001) & -0.004 (0.009) & -0.012 (0.018) \\ 
Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & 0.014 (0.064) & 0.047 (0.079) \\
  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.025 (0.077) & 0.043 (0.158) \\ 
Bias $STG.i$ & 0.8 & 0.065 (0.079)& 0.021 (0.081)& 0.023 (0.051)& 0.003 (0.067) & 0.009 (0.075) & 0.041 (0.074) \\
 & 0.5 & 0.059 (0.099) & 0.052 (0.109) & 0.006 (0.053)& 0.007 (0.076)& 0.076 (0.224) & 0.024 (0.153) \\
\% Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
  &0.5& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
   \hline
\multicolumn{8}{p{9in}}{Bias entries for the $TPF(3|c)$ and $FPF(3|c)$ curves are average bias over the simulations and 2000 points on each simulated curve (empirical standard deviation of the estimated statistic over the simulations). H0$_1$ is the null hypothesis of no surrogate value (i.e., no variation in risk difference over $S_j(1)$ for any time point), based on the joint Wald test of $\gamma_{0j}=\gamma_{1j}=0$ $\forall j\in\{1\ldots,J\}$. *Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,x)$.}
\end{tabular}
\end{center}
\end{sidewaystable}

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power of difference in $STG(t)$ tests for $t=3$ years \label{Power}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$& $STG(3)^{*}=0.23$  & $STG(3)^{*}=0.51$\\
  \hline
$STG(3)=0.1$ &0.5 & 4.80 & 15.80 & 80.40 & 16.90 & 58.20 \\ 
   &0.8 &7.20 & 31.20 & 90.00 & 32.80 & 96.40 \\ 
$STG(3)=0.24$ &0.5&  & 4.40 & 41.80 & 8.00 & 39.00 \\ 
   &0.8 &  & 5.20 & 49.60 & 10.00 & 69.60 \\ 
$STG(3)=0.48$&0.5 &  &  & 6.00 & 44.30 & 13.60 \\ 
   &0.8&  &  & 5.20 & 47.60 & 17.80 \\ 
$STG(3)^{*}=0.23$&(0.5, 0.5) &  &  &  & 3.80 & 21.60 \\ 
   &(0.8, 0.8) &  &  &  & 5.00 & 13.80 \\ 
$STG(3)^{*}=0.51$&(0.5, 0.5)&   &  &  &  & 2.40 \\ 
   &(0.8, 0.8)&  &  &  &  & 6.00 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $STG_1(3)-STG_2(3)=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
\end{tabular}
\end{center}
\end{sidewaystable}

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power of difference in $\widetilde{STG}$ tests for $t \in (\tau, 3.5)$ years \label{IPow}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}=0.47$& $\widetilde{STG}^*=0.23$  & $\widetilde{STG}^*=0.52$\\
  \hline
$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 75.6 & 24.6 & 56 \\ 
&  0.5  & 7.2 & 37.6 & 91.8 & 93.9 & 97.4 \\ 
 $\widetilde{STG}=0.25$ & 0.8 &  & 4 & 63.6 & 47.2 & 37.6 \\ 
&  0.5 &  & 4 & 78.8 & 80 & 94.8 \\ 
$\widetilde{STG}=0.47$ & 0.8 &  &  & 6 & 7.5 & 8.8 \\ 
 & 0.5 &  &  & 6 & 2.8 & 16.2 \\ 
$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 24.2 \\ 
&  (0.5, 0.5) &  &  &  & 6.3 & 17.2 \\ 
$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 1.6 \\ 
 & (0.5, 0.5) &  &  &  &  & 6 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $\widetilde{STG}_1-\widetilde{STG}_2=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
\end{tabular}
\end{center}
\end{sidewaystable}

\section{Example}
The Diabetes Control and Complication Trial (DCCT) enrolled 1,441 persons with type 1 diabetes from 1983 to 1989 to determine the effects of intensive diabetes therapy on long-term complications of diabetes. Participants in the DCCT were randomly assigned to intensive diabetes therapy aimed at lowering glucose concentrations as close as safely possible to the normal range or to conventional therapy aimed at preventing hyperglycemic symptoms. One of the outcomes of the DCCT, nephropathy (damage to the kidneys), is the leading cause of death and dialysis in the young with type 1 diabetes, particularly those with poorly controlled glucose levels. Nephropathy is often defined by a high albumin excretion rate, as micro-albuminuria (defined as albumin excretion rate $>$ 30 mg/24 hr) is the least invasive gold standard for kidney damage. The trial ended early in 1993 due to strong evidence of treatment efficacy, with a longest follow-up time of 7.5 years after the biomarker measurments of interest; estimated adjusted mean risk of micro-albuminuria was reduced by 56\% (P-value 0.01)\citep{DCCT}.

Although the trial could not be blinded to subjects or clinicians, it was blinded to investigators until the close of the trial. It is unlikely that the outcome would have been materially affected by subject or clinician knowledge of treatment status, as the only known risk factor in this population is poor diabetes management. Subjects that had dramatic behavioral changes while on trial were considered non-compliant and were censored at that point. This accounts for less than 3\% of person years in the trial and is balanced between treatment arms. The current analysis includes all participants who were free from micro-albuminuria at baseline (n = 1035); baseline micro-albuminuria is balanced over the arms of the trial.

We consider four candidate surrogates: the difference in log-transformed hemoglobin A1C (HBA1C) measurements from baseline to year one, the difference in log-transformed estimated glomerular filtration rate (EGFR) measurements from baseline to year one, their linear combination into a univariate surrogate (HBA1C-EGFR) and the multivariate combination of the linear combination HBA1C-EGFR and HBA1C. Although there are some subjects for whom these values increased over the year, expert opinion is that increases were most likely due to measurement error or random fluctuation and are not indications of clinically relevant increases. We therefore truncate both candidate surrogates at zero for our primary analysis. EGFR values increased in only 1.2\% of subjects, 6.1\% for HBA1C and both increased in only one subject. This is consistent with the error for both being random. As a sensitivity analysis, we repeat the analysis with the non-truncated measurements and found the the results were similar, no reported.

The event of interest is the onset of persistent micro-albuminuria, which is defined as having two consecutive albumin excretion rate measurements $>$ 30 mg/g. No subjects had an event prior to 1 year post-randomization. All subjects had the BIPs measured, which is defined as a linear combination of baseline laboratory measurements and demographic variables. The estimated linear correlation between the joint linear combination BIP and the candidate PS are 0.57 and 0.76 for HBA1C and univariate linear combination HBA1C-EGFR, respectively. All three univariate candidate PS also have respective individual BIPs. The correlation between the individual BIP and their PS are 0.7, for HBA1C, 0.88 for EGFR and 0.76 for the univariate linear combination HBA1C-EGFR. The individual BIPs are used for investigation of each biomarkers' univariate PS value, as well as comparison between the three univariate PS candidates. The joint linear combination BIP is used for the combined model. For all analyses we fit the conditional Weibull model proposed in A5 via the semi-parametric EML method and perform 1000 bootstraps.

The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001. The P-value for testing H0$_1$ for EGFR is 0.19. The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.055. The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has an estimated P-value for testing H0$_1$ of 0.018. Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone. The estimated $\widehat{\widetilde{STG}}$ is 0.2 for HBA1C alone and 0.15 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.202. Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing of any of the suggested summary statistics.

The P-value for the test that $\widehat{\widetilde{STG}}$ difference between HBA1C alone and the univariate the linear combination of HBA1C-EGFR is zero, is 0.057. This is evidence that HBA1C alone is superior to the univariate the linear combination of HBA1C-EGFR. As the difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS is so small, it is not surprising that the P-value for the null that their difference is equal to zero is 0.99.

Figure \ref{exp} Panel A depicts the $\widehat{CDF}^{8.5}(c)$ versus $c$ curves for all four candidate surrogates, HBA1C, EGFR the univariate linear combination HBA1C-EGFR and the multivariate combination. The EGFR curve is most similar to the marginal risk difference line, supporting the finding that EGFR has little value as a PS. The multivariate combination surrogate which is flatter than the EGFR alone curve is very similar to HBA1C curve. This provides further evidence that the combination candidate does not improve upon HBA1C alone. We use time-point year 7.5 for our main analysis because it is the longest follow-up time in the trial. As there is no evidence of violation of A6, Figure \ref{exp} Panel B displays the $\widehat{ROC}^{8.5}(q)$ versus $q$ curves for the HBA1C and the two combination candidates. Figure \ref{exp} Panel C depicts the $STG(t)$ versus $t$ for $t$ greater than one year for HBA1C and the two combination candidates. Both Panels B and C support that there is no significant difference between HBA1C alone and the multivariate combination, but that HBA1C alone is significantly better than the univariate linear combination candidate.

A naive evaluation of the association of the two univariate candidate surrogates and outcome separately in each arm
of the trial, using a Cox proportional hazards model and the hazard ratio magnitude for ranking, gives the same ranking of candidate surrogates in the treatment arm as do our methods. The findings support our conclusion that EGFR has no association with outcome under treatment. However, in the control arm EGFR is significantly associated with outcome and HBA1C is not. Although the naive analyses finds some evidence that HBA1C is a better candidate surrogate than EGFR under treatment, it will not always be the case that rank is preserved among the correlates of risk in the treatment arm alone. The surrogate evaluation literature has many examples of biomarkers that are highly correlated with outcome under treatment alone, but are not associated with treatment efficacy \citep{flemming96}.

\begin{sidewaysfigure}[H]
\begin{center}
\includegraphics[scale=0.75]{dcct-example-figure-2014-6-2.pdf}
\end{center}
\caption{Panel A depicts the estimated $CDF^{8.5}(c)$ versus $c$ for the truncated univariate candidate principal surrogates HBA1C and EGFR as well as for the univariate linear combination HBA1C-EGFR candidate and the multivariate candidate PS. Panel B depicts the estimated $ROC^{8.5}(q)$ versus $q$ for HBA1C and both of the combination candidate. Both panel suggest that HBA1C is better than EGFR alone or their univariate combination and that there is no significant difference between multivariate combination and HBA1C alone. Panel C depicts the $STG(t)$ versus $t$ for each of truncated candidate PS that showed some evidence of surrogate value, HBA1C and the univariate combination HBA1C-EGFR and the multivariate combination, for $t$ greater than one year and less than the trial close of nine years. \label{exp}}
\end{sidewaysfigure}

\section{Discussion}
Principal surrogates are important endpoints for Phase I and II trials. Without reliable surrogate endpoints for these trials, treatment development is slowed, allowing poor treatments to be advanced to Phase III and abandoning potentially effective formulations prior to Phase III. Few PS evaluation methods allow for comparison of candidate PSs or for the evaluation of combinations of biomarkers as candidate PS and the few existing methods for comparison or combination of biomarkers as candidate PS have not been explored for time-to-event data. Our proposed Weibull model extension of the \citet{Huang11} semi-parametric EML method as well as our proposed classification accuracy and summary measures accommodate evaluation of multivariate PS and comparison of candidate PS using time-to-event data.

Although our proposed estimands and estimators for the classification accuracy curves depend on assumption A6 to be well defined, we show that when there are small undetectable violations of A6, the estimates approximate the desirable modified estimands of no-benefit or unaffected ROC curves (Appendix A supplementary materials). Even when A6 is clearly violated, the suggested $CDF^{t}_{\Delta}(c)$ and $STG(t)$ versus $t$ curves allow for visual comparison of candidate surrogates and the $STG(t)$ or $\widetilde{STG}$ difference tests perform well for discriminating between two candidate PS that both have some value as a PS. While only the risk estimands themselves and the proposed $CDF^{t}(c)$ curve preform well for biomarkers with no surrogate value, our proposed coefficient-based test of any surrogate value has satisfactory power and correct Type 1 error rate. Using the proposed methods we were able to show that the DCCT candidate surrogate HBA1C has more value as a PS than EGFR, and that the combination of HBA1C and EGFR in the same risk model was not an improvement over HBA1C alone.

Our proposed semi-parametric EML estimators seem to perform well when there is a predictive BIP with correlation with the candidate PS of at least 0.5. The performance decreases with the correlation of the BIP and $S_j(1)$, as has been illustrated for previous EML methods. This is a limitation of EML based methods, however, for the motivating example the BIP correlation exceeds $0.5$. The need for a joint BIP that is well correlated with all biomarkers included in a multivariate PS model is also a limitation of the methods. Methods allowing for the use of individual BIP for each biomarker included in a multivariate PS model would be a useful extension to our methods. Although it is not the focus of this work, our model allows for time-variation in the hazard, an extension to the existing time-to-event PS evaluation methods \citep{Qin07, Miao13}. This extension is discussed in greater detail in the supplementary materials where we consider a more complex Weibull model allowing for time-variation in the treatment effect and the surrogate quality following \citet{Gabriel13}.

\bibliographystyle{statmed}
%\bibliographystyle{unsrtnat}
\bibliography{bib_dissertation}
\end{document}

### 2014-06-03 power_table.R

@@ -1,0 +1,27 @@
+install.packages('xtable')
+library(xtable)
+
+ISTG<-read.csv('power-table-integrated-STG-erin.csv')
+
+
+ISTG$X<-c(rep(c(0.5, 0.8), 3), 0,0,0,0)
+ISTG$X[7]<-ISTG$X[9]<-paste('(0.5,0.5)')
+ISTG$X[8]<-ISTG$X[10]<-paste('(0.8,0.8)')
+
+
+
+xtable(ISTG)
+
+
+rownames(ISTG)<-c('$\\widetilde{STG}=0.10$', '2', '$\\widetilde{STG}=0.25$', '4',
+'$\\widetilde{STG}=0.47$', '6', '$\\widetilde{STG}^{*}=0.23$', '8',  '$\\widetilde{STG}^{*}=0.52$', '10')
+
+
+
+
+
+
+
+
+
+ 

### 2014-06-03 Compare_053014_statsmed_revisions.tex
Proofread and fixed typos

@@
 Often evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality.
-We develop two extensions of the standard total gain \citet{Huang11}, both of which provide a single values measure of surrogate quality that can be used to compare candidate surrogates for the same clinical output in the same trial.
-The time-dependent standard total gain, ($STG(t)$), allows for a single value comparison of marginal CEP models as well as inference for comparison of candidate PS that summaries all information and properly accounts for censoring at and given time point $t$.
+We develop two extensions of the standardized total gain \citet{Huang11}, both of which provide a single values measure of surrogate quality that can be used to compare candidate surrogates for the same clinical output in the same trial.
+The time-dependent standardized total gain, ($STG(t)$), allows for a single value comparison of marginal CEP models as well as inference for comparison of candidate PS that summaries all information and properly accounts for censoring at and given time point $t$.
 Over a range of time points, a display of $STG(t)$ versus $t$ curve provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time.
-The time integrated  standard total gain, ($widetilde{STG}$), which is the time-dependent standard total gain integrated over the estimated survive times, provides a single value summary of surrogate quality independent of time.
+The time integrated  standardized total gain, ($widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survive times, provides a single value summary of surrogate quality independent of time.
 
@@
 We also suggest a summary statistic standardized total gain that quantifies both our suggested accuracy concept and ACS in a single measurement.
-Details for standard total gain are given in Section \ref{STG}.
+Details for standardized total gain are given in Section \ref{STG}.
 As well, if one assumes a parametric a model for risk conditional on a set of biomarkers a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.
@@
 Assumption A2, is generally true in randomized, blinded trials.
-Assumption A3 aides in the identification of subjects with min$(T,C) < \tau$ to be excluded from the evaluation cohort without causing bias.
+Assumption A3 aids in the identification of subjects with min$(T,C) < \tau$ to be excluded from the evaluation cohort without causing bias.
 Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced events or drop-out over the trial arms prior to $\tau$.
@@
 We can then use these imputed values $S_{j,i,k}^*(1)$ to estimate the EML contribution for subject $i$ by $\int g(t_i|\gamma, \beta, z_i, s_{1,1}, \ldots, s_{1,J}, y_i)dF_{S_{1}(1),\ldots,S_{J}(1)|W}(s_{1,1},\ldots,s_{1,J}|W)$, by the empirical integral ${\left(1/n_{V}\right)}\sum_k^{n_V}{g_z(t_i|\gamma, \beta,z_i, S_{1,k}^*(1), \ldots,S_{J,k}^*(1),y_i)}$.
-missingness in the treatment arm, such as subsampling, case-cohort sampling and unplanned missing at random variables, can be inputed in the same way.
-In cases were the validation set is not a random sample of the treatment subjects, the location-scale models can be weighted to accommodate the bias sampling.
+missingness in the treatment arm, such as subsampling, case-cohort sampling and unplanned missing at random variables, can be imputed in the same way.
+In cases where the validation set is not a random sample of the treatment subjects, the location-scale models can be weighted to accommodate the biased sampling.
 
@@
 Here, $TPF(t|c)$ and $FPF(t|c)$ are the time-dependent true and false positive fractions.
-Sensitivity and true positive fraction represent the probability that we predict a subject will have benefit using the model and some threshold $c$, $\Delta(t|s_{1,1}, \dots, s_{1,J}) > c$, given that we observe that they did have benefit, $D(t)=1$.
+Sensitivity, or true positive fraction, represents the probability that we predict a subject will have benefit using the model and some threshold $c$, $\Delta(t|s_{1,1}, \dots, s_{1,J}) > c$, given that we observe that they did have benefit, $D(t)=1$.
 Similarly, specificity represents the probability that we predict a subject will have no benefit based on our model and using some threshold $c$, $\Delta(t|s_{1,1}, \dots, s_{1,J}) < c$, given that we observe that they did not have benefit, $D(t)=0$.
 This is one minus the false positive fraction, which is the probability that we falsely predict that a subject will have benefit, when they truly did not.
-These probabilities over all the thresholds, demonstrate how well our surrogate based model predicts our outcome of interest, benefit from treatment, and are therefore a useful illustration of the usefulness of the surrogate to predict outcome given the assumed model.
-Figure 1 of Appendix A of supplementary materials illustrate the meaning of these classification accuracy measures.
+These probabilities demonstrate how well our surrogate based model predicts benefit from treatment, and are therefore a useful illustration of the usefulness of the surrogate to predict outcome given the assumed model.
+Figure 1 of Appendix A of supplementary materials illustrates the meaning of these classification accuracy measures.
 
@@
 Greater discussion of this point is given the Appendix A of the supplementary materials.
-When there are obvious violations of A6 the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued.
+When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued.
 When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation or candidate surrogates.
@@
 If this candidate is found to be statistically significantly superior to some or all the other candidates this provides greater evidence of superiority.
-However, as with almost all statistical tests, lack of significant reject of the null does not prove the null; lack of significant  $\widetilde{STG}$ difference does not prove lack of superiority.
+However, as with almost all statistical tests, a failure to reject the null does not prove the null; lack of significant  $\widetilde{STG}$ difference does not prove lack of superiority.
 
@@
 Results for the 0.3 correlation are displayed.
-500 simulation were run for each scenario; some simulations did not converge and were removed from the simulations used for bias and power calculations, there was a 2.5\% non-convergence rate  All candidate PS in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.
+500 replicates of the simulation were run for each scenario.
+Some simulations did not converge and were removed from the simulations used for bias and power calculations; there was a 2.5\% non-convergence rate.
+All candidate PS in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.
 
@@
 
-Figure \ref{exp} Panel A depicts the $\widehat{CDF}^{8.5}(c)$ versus $c$ curves for all four candidate surrogates, HBA1C, EGFR the univariate linear combination HBA1C-EGFR and the multivariate combination.
+Figure \ref{exp} Panel A depicts the $\widehat{CDF}^{7.5}(c)$ versus $c$ curves for all four candidate surrogates, HBA1C, EGFR the univariate linear combination HBA1C-EGFR and the multivariate combination.
 The EGFR curve is most similar to the marginal risk difference line, supporting the finding that EGFR has little value as a PS.
@@
 We use time-point year 7.5 for our main analysis because it is the longest follow-up time in the trial.
-As there is no evidence of violation of A6, Figure \ref{exp} Panel B displays the $\widehat{ROC}^{8.5}(q)$ versus $q$ curves for the HBA1C and the two combination candidates.
+As there is no evidence of violation of A6, Figure \ref{exp} Panel B displays the $\widehat{ROC}^{7.5}(q)$ versus $q$ curves for the HBA1C and the two combination candidates.
 Figure \ref{exp} Panel C depicts the $STG(t)$ versus $t$ for $t$ greater than one year for HBA1C and the two combination candidates.
@@
 \begin{center}
-\includegraphics[scale=0.75]{dcct-example-figure-2014-6-2.pdf}
+\includegraphics[scale=0.65]{dcct-example-figure-2014-6-2.pdf}
 \end{center}
@@
 Panel B depicts the estimated $ROC^{8.5}(q)$ versus $q$ for HBA1C and both of the combination candidate.
-Both panel suggest that HBA1C is better than EGFR alone or their univariate combination and that there is no significant difference between multivariate combination and HBA1C alone.
+Both panels suggest that HBA1C is better than EGFR alone or their univariate combination and that there is no significant difference between multivariate combination and HBA1C alone.
 Panel C depicts the $STG(t)$ versus $t$ for each of truncated candidate PS that showed some evidence of surrogate value, HBA1C and the univariate combination HBA1C-EGFR and the multivariate combination, for $t$ greater than one year and less than the trial close of nine years.
@@
 
-Although our proposed estimands and estimators for the classification accuracy curves depend on assumption A6 to be well defined, we show that when there are small undetectable violations of A6, the estimates approximate the desirable modified estimands of no-benefit or unaffected ROC curves (Appendix A supplementary materials).
+Although our proposed estimands and estimators for the classification accuracy curves depend on assumption A6 to be well-defined, we show that when there are small undetectable violations of A6, the estimates approximate the desirable modified estimands of no-benefit or unaffected ROC curves (Appendix A supplementary materials).
 Even when A6 is clearly violated, the suggested $CDF^{t}_{\Delta}(c)$ and $STG(t)$ versus $t$ curves allow for visual comparison of candidate surrogates and the $STG(t)$ or $\widetilde{STG}$ difference tests perform well for discriminating between two candidate PS that both have some value as a PS.
-While only the risk estimands themselves and the proposed $CDF^{t}(c)$ curve preform well for biomarkers with no surrogate value, our proposed coefficient-based test of any surrogate value has satisfactory power and correct Type 1 error rate.
+While only the risk estimands themselves and the proposed $CDF^{t}(c)$ curve perform well for biomarkers with no surrogate value, our proposed coefficient-based test of any surrogate value has satisfactory power and correct Type 1 error rate.
 Using the proposed methods we were able to show that the DCCT candidate surrogate HBA1C has more value as a PS than EGFR, and that the combination of HBA1C and EGFR in the same risk model was not an improvement over HBA1C alone.

### 2014-06-03 Compare_053014_statsmed_revisions.tex
edit

--- 2014-06-03 Compare_053014_statsmed_revisions.tex
@@
 \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.
-Our method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.
+Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.
 
-We propose the use of several curves for displaying surrogate quality which consider treatment efficacy within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves.
+We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves.
+A CEP is any function of the treatment arm specific biomarker conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero.
+A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement of $S$.
+Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect and the strength of this association can be displayed by the marginal-CEP versus candidate PS under active treatment figure \citep{Gilbert08}.
 Evaluation of multivariate principal surrogates is complicated by the increased dimension.
-Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect.
-The strength of this association can be displayed by the CEP curve \citep{Gilbert08}.
-The CEP is a contrast of risk over the treatment arms within subgroups defined by the candidate surrogate under active treatment and marginalizing over the candidate surrogate under control.
-The dimension of the marginal CEP function increases with the dimension of the candidate surrogate, making the marginal CEP difficult to display and interpret for multivariate candidate PS.
-This makes the curves two-dimensional regardless of the number of biomarkers included in the risk model.
+
+The dimension of the marginal-CEP versus candidate PS under active treatment figure increases with the dimension of the candidate surrogate, making the marginal-CEP difficult to display and interpret for multivariate candidate PS.
+If instead we consider the clinical outcome within level of the marginal-CEP we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model.
 Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define.
@@
 Time-dependent TPF and FPF curves based on the levels of diagnostic tests and time-independent curves based on levels of causal contrasts, \citep{Huang12b}, have been proposed in previous works.
-However, no version of the TPF or FPF curves have been proposed for the evaluation or comparison of candidate PSs, nor has the causal-contrast-based and time-dependent versions of the curves been defined.
+However, no version of the TPF or FPF curves have been proposed for the evaluation or comparison of candidate PS, nor has the CEP-based and time-dependent versions of the curves been defined.
 
-We develop summary statistics for the comparison of candidate PS for a continuous time-to-event end point.
+We develop summary statistics for the comparison of candidate PS for a continuous time-to-event endpoint.
 Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS .
-Often evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality.
-We develop two extensions of the standard total gain \citet{Huang11}, both of which provide a single values measure of surrogate quality that can be used to compare candidate surrogates for the same clinical output in the same trial.
-The time-dependent standard total gain, ($STG(t)$), allows for a single value comparison of marginal CEP models as well as inference for comparison of candidate PS that summaries all information and properly accounts for censoring at and given time point $t$.
-Over a range of time points, a display of $STG(t)$ versus $t$ curve provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time.
-The time integrated  standard total gain, ($widetilde{STG}$), which is the time-dependent standard total gain integrated over the estimated survive times, provides a single value summary of surrogate quality independent of time.
+Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality.
+We develop two extensions of the standardized total gain \citet{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial.
+The time-dependent standardized total gain, ($STG(t)$), allows for the comparison of candidate PS at a given time point $t$ that properly accounts for censoring.
+Over a range of time points, a display of $STG(t)$ versus $t$ provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time.
+The time integrated  standardized total gain, ($\widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survival times, provides a single value summary of surrogate quality independent of time.
 
-In Section 2, we outline our notation.
-In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PSs or the comparison of candidate PSs.
+In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PS or the comparison of candidate PS.
 In Section 3.1, we describe the time-dependent risk estimands and outline the assumptions that are helpful for identifying these estimands.
@@
 In Section 3.7 we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}.
-In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PSs.
+In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PS.
 In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT).
@@
 
-All $S_j$ are measured at the same fixed time point $\tau$ after randomization, and this time point, post-randomization but prior to clinical-event, and z define the surrogate measurement.
+All $S_j$ are measured at the same fixed post-randomization time point $\tau$.
 If $T_{j,i}(z)$ is less than $\tau$ for a given subject $i$, all $S_{j,i}(z)$ are undefined for that subject, as observation of the post-randomization pre-clinical-event measures $S_{j}(z)$ are no longer possible under either treatment condition.
-For this reason, subjects with min$(T,C) < \tau$ are excluded from the evaluation cohort.
+For this reason, subjects with $T< \tau$ are excluded from the evaluation cohort.
 Let $M_j$ be the indicator that $S_{j}(1)$ is observed, and let $s_{1,j,i}$ be the realization of $S_{j}(1)$ for subject $i$.
@@
 This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.
-ACN is a strict but sensible necessary condition, which guarantees that there is no effect on the clinical out when there is no effect on the PS.
+ACN is a strict but sensible necessary condition, which guarantees that there is no effect on the clinical outcomewhen there is no effect on the PS.
 However, it is clearly not a sufficient condition, as any biomarker in a trial with no clinical treatment effect is likely to satisfy ACN.
-The combination of ACS and ACN are often looked evaluated in the current literature, with some works relaxing strict ACN.
+The combination of ACS and ACN are often used for evaluation in the current literature, with some works relaxing strict ACN.
 ACS also suggests a possibly ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.
@@
 Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically.
-As well, the definition does not seem as sensible in the multi-dimensional case.
+The definition does not seem as sensible in the multi-dimensional case.
 Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates.
@@
 We also suggest a summary statistic standardized total gain that quantifies both our suggested accuracy concept and ACS in a single measurement.
-Details for standard total gain are given in Section \ref{STG}.
-As well, if one assumes a parametric a model for risk conditional on a set of biomarkers a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.
+Details for standardized total gain are given in Section \ref{STG}.
+As well, if one assumes a parametric a model for risk conditional on a set of biomarkers, a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.
 
@@
 The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF.
-The conditional CDF-based risk estimands can by defined as:
+The conditional CDF-based risk estimands can be defined as:
 \begin{eqnarray*}
@@
 \end{itemize} 
-Assumption A1 implies that subject's potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency.
+Assumption A1 implies that subjects' potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency.
 For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$.
@@
 One can test the null hypothesis of no surrogate value or no ACS, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$.
-If any of the $S_{J}(1)$ associated terms significantly differ from zero there is evidence of variation in the risk difference over the candidate PS.
+If any of the $S_{J}(1)$ associated terms significantly differ from zero then there is evidence of variation in the risk difference over the candidate PS.
 
@@
 \subsection{Cumulative Distribution Function of $\Delta$} \label{CDF}
- Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PSs and for the comparison of candidate PS.
+ Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PS and for the comparison of candidate PS.
 An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$.
@@
 
-We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$.
+We investigate four different univariate candidate PS with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$.
 We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $STG(3)$ of 0.23 and 0.51, respectively.
@@
 Without reliable surrogate endpoints for these trials, treatment development is slowed, allowing poor treatments to be advanced to Phase III and abandoning potentially effective formulations prior to Phase III.
-Few PS evaluation methods allow for comparison of candidate PSs or for the evaluation of combinations of biomarkers as candidate PS and the few existing methods for comparison or combination of biomarkers as candidate PS have not been explored for time-to-event data.
+Few PS evaluation methods allow for comparison of candidate PS or for the evaluation of combinations of biomarkers as candidate PS and the few existing methods for comparison or combination of biomarkers as candidate PS have not been explored for time-to-event data.
 Our proposed Weibull model extension of the \citet{Huang11} semi-parametric EML method as well as our proposed classification accuracy and summary measures accommodate evaluation of multivariate PS and comparison of candidate PS using time-to-event data.

### 2014-06-03 Compare_053014_statsmed_revisions.tex
One more typo found

--- 2014-06-03 Compare_053014_statsmed_revisions.tex
@@
 Over a range of time points, a display of $STG(t)$ versus $t$ curve provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time.
-The time integrated  standardized total gain, ($widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survive times, provides a single value summary of surrogate quality independent of time.
+The time integrated  standardized total gain, ($\widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated distribution of survival times, provides a single value summary of surrogate quality independent of time.
 

### 2014-06-03 Compare_053014_statsmed_revisions.tex

--- 2014-06-03 Compare_053014_statsmed_revisions.tex
@@
 We propose the use of functions based on the CEP rather than the set of biomarker values for display of ACS.
-The propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS.
+We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS.
 We also suggest a summary statistic standardized total gain that quantifies both our suggested accuracy concept and ACS in a single measurement.
 Details for standardized total gain are given in Section \ref{STG}.
-As well, if one assumes a parametric a model for risk conditional on a set of biomarkers, a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.
+As well, if one assumes a parametric a model for risk conditional on a set of biomarkers a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.
 
@@
 
-We investigate four different univariate candidate PS with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$.
+We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$.
 We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $STG(3)$ of 0.23 and 0.51, respectively.

### 2014-06-04 Compare_053014_statsmed_revisions.tex
updated background

@@
 Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z=\{0,1\}$.
-Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time.
+Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=min(T(z),C(z))$.
 Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z=\{0,1\}$.
@@
 \item A2: Ignorable Treatment Assignment
-\item A3: Equal individual clinical risk up to time $\tau$: $T(1)\leq\tau$ if and only if $T(0)\leq\tau$
+\item A3: Equal drop our and individual clinical risk up to time $\tau$: $X(1)\leq\tau$ if and only if $X(0)\leq\tau$
 \item A4: Random censoring: $T(z) \perp C(z)$ for $z=\{0,1\}$.

### 2014-06-05 Compare_053014_statsmed_revisions.tex
current

@@
 We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves.
-A CEP is any function of the treatment arm specific biomarker conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero.
+A CEP is a function of the treatment arm specific biomarker conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero.
 A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement of $S$.

### 2014-06-05 Compare_053014_statsmed_revisions.tex
Another round of editing and fixed example figure to be 1 by 3

@@
 We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves.
-A CEP is a function of the treatment arm specific biomarker conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero.
-A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement of $S$.
-Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect and the strength of this association can be displayed by the marginal-CEP versus candidate PS under active treatment figure \citep{Gilbert08}.
+A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero.
+A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement.
+Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect.
+This association can be displayed by the marginal-CEP versus candidate PS curve \citep{Gilbert08}.
 Evaluation of multivariate principal surrogates is complicated by the increased dimension.
 
-The dimension of the marginal-CEP versus candidate PS under active treatment figure increases with the dimension of the candidate surrogate, making the marginal-CEP difficult to display and interpret for multivariate candidate PS.
-If instead we consider the clinical outcome within level of the marginal-CEP we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model.
+The dimension of the marginal-CEP versus candidate PS curve increases with the dimension of the candidate surrogate, making the marginal-CEP difficult to display and interpret for multivariate candidate PS.
+If instead we consider the clinical outcome within levels of the marginal-CEP, we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model.
 Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define.
 We also consider the cumulative distribution function (CDF) of the marginal CEP function.
-Time-dependent TPF and FPF curves based on the levels of diagnostic tests and time-independent curves based on levels of causal contrasts, \citep{Huang12b}, have been proposed in previous works.
-However, no version of the TPF or FPF curves have been proposed for the evaluation or comparison of candidate PS, nor has the CEP-based and time-dependent versions of the curves been defined.
 
@@
 Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality.
-We develop two extensions of the standardized total gain \citet{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial.
+We develop two extensions of the standardized total gain \citep{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial.
 The time-dependent standardized total gain, ($STG(t)$), allows for the comparison of candidate PS at a given time point $t$ that properly accounts for censoring.
@@
 In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PS or the comparison of candidate PS.
-In Section 3.1, we describe the time-dependent risk estimands and outline the assumptions that are helpful for identifying these estimands.
+In Section 3.1, we describe the time-dependent risk estimands and outline some assumptions that are helpful for identifying these estimands.
 We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.
 We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison.
-In Section 3.7 we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}.
+In Section 3.7, we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}.
 In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PS.
 In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT).
-We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combination of the two candidates is not an improvement over the higher quality surrogate alone.
+We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combinations of the two candidates are not improvements over the higher quality surrogate alone.
 In Section 6, we discuss potential limitations of our methods and future avenues for research.
@@
 In the principal stratification framework of \citet{Frangakis02} we use potential outcomes, where all post-randomization measures are considered under assignment to either treatment arm for each individual.
-Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z=\{0,1\}$.
+Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z\in \{0,1\}$.
 Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=min(T(z),C(z))$.
-Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z=\{0,1\}$.
+Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z\in \{0,1\}$.
 Let $Y_{i}^{t}(z)$ be the indicator that $T_i(z)\leq t$.
@@
 This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.
-ACN is a strict but sensible necessary condition, which guarantees that there is no effect on the clinical outcomewhen there is no effect on the PS.
+ACN is a strict but sensible and necessary condition, which guarantees that there is no effect on the clinical outcome when there is no effect on the PS.
 However, it is clearly not a sufficient condition, as any biomarker in a trial with no clinical treatment effect is likely to satisfy ACN.
 The combination of ACS and ACN are often used for evaluation in the current literature, with some works relaxing strict ACN.
-ACS also suggests a possibly ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.
+ACS also suggests a possible ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.
 Strict ACN alone classify candidates as either PS or not PS, with no other indication of quality.
 
-In \citet{Gilbert08}  and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate; marginal risk estimands condition on $S(1)$ and marginalizing over the distribution of $S(0)$.
+In \citet{Gilbert08}  and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate; marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$.
 The concept of ACS is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$.
 This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect.
-Although in some special cases when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN can not be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}.
+Although in some special cases when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}.
 Often the concept of ACN is relaxed in practice, such that if the CEP of interest is equal to zero when $s_1=0$, ACN is said to be likely.
-Although this is similar to ANC, this is not the same as ANC unless CB holds and $S(0)=0$.
-\citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in general clinical trial where biomarkers under placebo will not tend to be constant.
+Although this is similar to ACN, this is not the same as ACN unless CB holds and $S(0)=0$.
+\citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under placebo will tend not to be constant.
 
@@
 We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional of a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$.
-For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections 3.3 and 3.4.
-Like ACN, the classification accuracy measures allow for the quantification of the agreement between treatment effect on the clinical outcome and predictive model when there is no treatment effect on any of the biomarkers in the model.
-Unlike ANC, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.
-These accuracy measures can also be displayed as curves vs.
-the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model.
-We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary for a set of biomarkers to have value as a PS, but that this is not sufficient.
+For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections \ref{CAM} and \ref{CDF}.
+Like ACN, the classification accuracy measures allow for the quantification of the agreement between the treatment effect on the clinical outcome and the predictive model when there is no treatment effect on any of the biomarkers in the model.
+Unlike ACN, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.
+These accuracy measures can also be displayed as curves versus the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model.
+We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary but not sufficient for a set of biomarkers to have value as a PS.
 Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.
@@
 We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS.
-We also suggest a summary statistic standardized total gain that quantifies both our suggested accuracy concept and ACS in a single measurement.
+We also suggest a summary statistic, the standardized total gain, that quantifies both our suggested accuracy concept and ACS in a single measurement.
 Details for standardized total gain are given in Section \ref{STG}.
-As well, if one assumes a parametric a model for risk conditional on a set of biomarkers a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model, the specifics of this test for our assumed model are given in Section \ref{CRE}.
+As well, if one assumes a parametric model for risk conditional on a set of biomarkers then a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model.
+The specifics of this test for our assumed model are given in Section \ref{CRE}.
 
@@
 Our suggested accuracy measures and the CDF of the risk model CEP can help more finely group candidates by predictive quality and amount of variation in CEP, but they do not provide inference for comparing candidates.
-We propose the use of standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for comparison inference.
+We propose the use of the standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for comparison inference.
 These are outlined in Sections \ref{STG} and \ref{ISTG}.
@@
 \item A2: Ignorable Treatment Assignment
-\item A3: Equal drop our and individual clinical risk up to time $\tau$: $X(1)\leq\tau$ if and only if $X(0)\leq\tau$
-\item A4: Random censoring: $T(z) \perp C(z)$ for $z=\{0,1\}$.
+\item A3: Equal drop out and individual clinical risk up to time $\tau$: $\min\{T(1), C(1)\} \leq\tau$ if and only if $\min\{T(0), C(0)\} \leq\tau$
+\item A4: Random censoring: $T(z) \perp C(z)$ for $z\in \{0,1\}$.
 \end{itemize} 
@@
 Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used.
-Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z=\{0,1\}$.
+Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z\in \{0,1\}$.
 
@@
 \end{itemize}
-for $z=\{0,1\}$.
+for $z\in \{0,1\}$.
 We focus on this classical model in the main text for clarity; more complex time-dependent models are outlined in Appendix B of the supplementary materials.
-In order for both assumptions A3 and A5 to hold simultaneously, we can only considered censoring or event times greater than $\tau$.
+In order for both assumptions A3 and A5 to hold simultaneously, we can only consider censoring or event times greater than $\tau$.
 This can be accomplished in practice by rescaling all follow-up times from $\tau$ rather than from enrollment.
@@
 One can test the null hypothesis of no surrogate value or no ACS, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$.
-If any of the $S_{J}(1)$ associated terms significantly differ from zero then there is evidence of variation in the risk difference over the candidate PS.
+If any of the $S_{J}(1)$ associated terms significantly different from zero then there is evidence of variation in the risk difference over the candidate PS.
 
@@
 We can then use these imputed values $S_{j,i,k}^*(1)$ to estimate the EML contribution for subject $i$ by $\int g(t_i|\gamma, \beta, z_i, s_{1,1}, \ldots, s_{1,J}, y_i)dF_{S_{1}(1),\ldots,S_{J}(1)|W}(s_{1,1},\ldots,s_{1,J}|W)$, by the empirical integral ${\left(1/n_{V}\right)}\sum_k^{n_V}{g_z(t_i|\gamma, \beta,z_i, S_{1,k}^*(1), \ldots,S_{J,k}^*(1),y_i)}$.
-missingness in the treatment arm, such as subsampling, case-cohort sampling and unplanned missing at random variables, can be imputed in the same way.
+Missingness in the treatment arm, such as subsampling, case-cohort sampling and unplanned missing at random variables, can be imputed in the same way.
 In cases where the validation set is not a random sample of the treatment subjects, the location-scale models can be weighted to accommodate the biased sampling.
 
-To accommodate multivariate surrogates, the same location-scale model is fit for each of the univariate candidate PS separately in the set of the validation subjects.
-Imputed values for each missing biomarker are produced separately, but provided that the same baseline variables are used in each model empirical integration over the set of all imputations is approximately equivalent to integrating over the joint distribution of biomarkers given the baseline variables used for imputation.
+To accommodate multivariate surrogates, the same location-scale model is fit for each of the univariate candidate PS separately in the set of validation subjects.
+Imputed values for each missing biomarker are produced separately, but provided that the same baseline variables are used in each model, empirical integration over the set of all imputations is approximately equivalent to integrating over the joint distribution of biomarkers given the baseline variables used for imputation.
 This is an important limitation of this method.
@@
 \begin{eqnarray*}
-\mbox{Sensitivity}(t|c)&=&P(\Delta(t|s_{1,1}, \dots, s_{1,J}) > c|D(t)=1)=TPF(t|c),\\
-\mbox{Specificity}(t|c)&=&P(\Delta(t|s_{1,1}, \dots, s_{1,J}) \leq c|D(t)=0)=1-FPF(t|c).
+\mbox{Sensitivity}(t|c)&=&P\{\Delta(t|s_{1,1}, \dots, s_{1,J}) > c|D(t)=1\}=TPF(t|c),\\
+\mbox{Specificity}(t|c)&=&P\{\Delta(t|s_{1,1}, \dots, s_{1,J}) \leq c|D(t)=0\}=1-FPF(t|c).
 \end{eqnarray*}
@@
 In this case, sensitivity and specificity are based on unobservable data and cannot be estimated empirically.
-However, under assumptions A1-A6, we can estimate them.
+However, under assumptions A1-A5 plus the additional assumption of monotonicity of treatment effect, we can estimate them.
+Monotonicity of treatment effect is formally stated as
+\begin{itemize}
+\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z\in \{0,1\}$.
+\end{itemize}
 Assumption A6 is needed to identify the classification accuracy measures, but monotonicity does not have to imply that the treatment is never harmful and in some cases the alternative assumption that the treatment never helps is more likely.
@@
 \begin{eqnarray*}
-TPF(t|c)&=&{E[\Delta(t|S_{1,1}, \dots, S_{1,J})\}\mathrm{I}_{\{\Delta(t|S_{1,1}, \dots, S_{1,J})>c\}}] \over E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\}}.
+TPF(t|c)&=& E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\mathrm{I}_{[\Delta(t|S_{1,1}, \dots, S_{1,J})>c]}\} \over E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\}.
 \end{eqnarray*}
@@
 The time-dependent ROC curve gives an illustration of the $TPF(t|c)$ and $FPF(t|c)$ relationship at given levels of $c$.
-This curve is defined by $ROC^{t}(q)=TPF(t|\{(FPF)^{-1}(t|q)\})$ for a given time point $t$, adapting the binary-clinical endpoint $ROC(q)$ definition of \citet{Huang12b}.
+This curve is defined by $ROC^{t}(q)=TPF\{t|FPF^{-1}(t|q)\}$ for a given time point $t$, adapting the binary-clinical endpoint $ROC(q)$ definition of \citet{Huang12b}.
 Figure \ref{4panel} panels B and D displays the $ROC^{t}(q)$ curves for several different surrogate quality levels.
@@
 When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued.
-When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation or candidate surrogates.
+When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation of candidate surrogates.
 
 \subsection{Cumulative Distribution Function of $\Delta$} \label{CDF}
+
  Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PS and for the comparison of candidate PS.
 An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$.
-If we assume monotonicity of treatment effect, formally stated as
-\begin{itemize}
-\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z=\{0,1\}$,
-\end{itemize}
-then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$).
+If we assume monotonicity of treatment effect (assumption A6) then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$).
 If we alternatively assume no benefit, then the line will be at the proportion of subjects with $D(t)=-1$.
@@
 \citet{Huang11} suggest a standardized version of TG for the comparison of PS for binary clinical endpoints.
-We extend the concept of total gain to the time-to-event setting and define the time-dependent total gain as $TG(t) = \int_0^{1}{|R^t(v) -(\rho_0(t) -\rho_1(t))|dv}$ and the time-dependent standardized total gain as $STG(t) \equiv TG(t)/[2(\rho_0(t) - \rho_1(t))\{1-\rho_0(t) + \rho_1(t)\}]$.
+We extend the concept of total gain to the time-to-event setting and define the time-dependent total gain as $TG(t) = \int_0^{1}{|R^t(v) -\{\rho_0(t) -\rho_1(t)\}|dv}$ and the time-dependent standardized total gain as $STG(t) \equiv TG(t)/\{2[\rho_0(t) - \rho_1(t)][1-\rho_0(t) + \rho_1(t)]\}$.
 
@@
 \widehat{TG}(t)&=&\int{|\widehat{\Delta}(t|s_{1,1}, \dots, s_{1,J}) -\{\hat{\rho}_0(t) -\hat{\rho}_1(t)\}|d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})} \hspace{4mm} \mbox{and} \label{TG} \\
-\widehat{STG}(t)&=&\widehat{TG}(t)/[2(\hat{\rho}_0(t) - \hat{\rho}_1(t)]\{1-\hat{\rho}_0(t) + \hat{\rho}_1(t)\}.
+\widehat{STG}(t)&=&\widehat{TG}(t)/\{2[\hat{\rho}_0(t) - \hat{\rho}_1(t)][1-\hat{\rho}_0(t) + \hat{\rho}_1(t)]\}.
 \end{eqnarray}
@@
 The coefficient test of  null hypothesis $H0_1$ can again be used to reduce the dimension of multivariate surrogates, as a biomarker $j$ for which the coefficients $\gamma_{0j}=\gamma_{1j}=0$ in a multivariate surrogate can be considered to not add value to surrogate.
-Once the field of consideration has been narrowed to considered only candidates with at least some surrogate quality, PS can be further evaluated based on their $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ and the $STG(t)$ v $t$  curves.
+Once the field of consideration has been narrowed to considered only candidates with at least some surrogate quality, PS can be further evaluated based on their $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ and the $STG(t)$ versus $t$  curves.
 These curves can also help to give a visualization of a ranking of the candidates.
@@
 
-Table \ref{Ipow} displays the power to discriminate between two surrogate quality levels based on the difference in integrated STG, $\widetilde{STG}$.
+Table \ref{IPow} displays the power to discriminate between two surrogate quality levels based on the difference in integrated STG, $\widetilde{STG}$.
 We again find that for all null hypothesis scenarios the test has nearly a nominal 5\% Type 1 error rate with similar to slightly higher power than the $STG(3)$ difference test for the same scenarios.
@@
 \begin{center}
-\includegraphics[scale=0.65]{dcct-example-figure-2014-6-2.pdf}
+\includegraphics[width=.9\textwidth]{dcct-example-figure-2014-6-2.pdf}
 \end{center}

### 2014-06-16 Compare_053014_statsmed_revisions.tex
revisions

@@
 However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information.
-We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate principal surrogates.
+We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate candidate principal surrogates.
 We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates.
@@
 \end{abstract}
-\keywords{Causal inference; Comparison of candidate principal stratification; Survival analysis; Accuracy measures; Multivariate principal stratification}
+\keywords{Causal inference; Surrogate endpoint evaluation; Survival analysis; Accuracy measures; Multivariate principal stratification}
 \maketitle
@@
 
-We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate principal surrogates of continuous time-to-event clinical endpoints.
+We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate candidate principal surrogates of continuous time-to-event clinical endpoints.
 Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint.
-\citet{Gabriel13} extended methods for evaluating univariate biomarker as a PS of continuous time-to-event clinical endpoints.
+\citet{Gabriel13} extended methods for evaluating univariate a biomarker as a PS of a continuous time-to-event clinical endpoints.
 \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.
@@
 
-The dimension of the marginal-CEP versus candidate PS curve increases with the dimension of the candidate surrogate, making the marginal-CEP difficult to display and interpret for multivariate candidate PS.
+The dimension of the marginal-CEP versus candidate PS curve increases with the dimension of the candidate surrogate, making the marginal-CEP curve difficult to display and interpret for multivariate candidate PS.
 If instead we consider the clinical outcome within levels of the marginal-CEP, we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model.
@@
 Let $D_i(t)=Y_{i}^{t}(0)-Y_{i}^{t}(1)$ be the individual treatment effect for subject $i$ on the clinical endpoint at or before time $t$.
-Let $\{S_{1,i}\ldots,S_{J,i}, T_i, C_i, Y_i, Y_{i}^{t} \}$ denote the observed values of the potential measures for subject $i$.
+Let $\{S_{1,i}\ldots,S_{J,i}, T_i, C_i, Y_i, Y_{i}^{t} \}$ denote the observed values of the potential outcomes for subject $i$.
 
@@
 If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN).
-Subsequent work added a second criterion that a good PS should modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, what \citet{Gilbert08} called average causal sufficiency ACS.
+Subsequent work added a second criterion that a good PS should strongly modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, the wide variability criterion.
 This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.
-ACN is a strict but sensible and necessary condition, which guarantees that there is no effect on the clinical outcome when there is no effect on the PS.
-However, it is clearly not a sufficient condition, as any biomarker in a trial with no clinical treatment effect is likely to satisfy ACN.
-The combination of ACS and ACN are often used for evaluation in the current literature, with some works relaxing strict ACN.
-ACS also suggests a possible ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.
-Strict ACN alone classify candidates as either PS or not PS, with no other indication of quality.
+ACN is a sensible condition that can help eliminate candidate PS, but is not a sufficient condition, as was shown in \citet{Vanderweele11}.
+The combination of wide effect modification and ACN are often used for evaluation in the current literature.
+The wide effect modification criterion may also suggest a possible ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.
 
-In \citet{Gilbert08}  and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate; marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$.
-The concept of ACS is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$.
+In \citet{Gilbert08} and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate.
+marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$.
+The concept of wide variability is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$.
 This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect.
-Although in some special cases when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}.
-Often the concept of ACN is relaxed in practice, such that if the CEP of interest is equal to zero when $s_1=0$, ACN is said to be likely.
-Although this is similar to ACN, this is not the same as ACN unless CB holds and $S(0)=0$.
-\citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under placebo will tend not to be constant.
+Other than the special case when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}.
+\citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under control will tend not to be constant.
 
-For a univariate candidate PS strict or relaxed ACN can be displayed using the joint CEP or the marginal CEP.
+For a univariate candidate PS ACN can be displayed using the joint CEP, or marginal CEP under case CB.
 Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically.
-The definition does not seem as sensible in the multi-dimensional case.
+As well, the definition does not seem as sensible in the multi-dimensional case.
 Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates.
-However, it seems like you would want less clinical treatment effect if there was no treatment effect on any of the biomarkers.
-We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional of a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$.
+However, ACN may not guarantee this is the case \citep{Vanderweele11}.
+We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional on a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$.
 For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections \ref{CAM} and \ref{CDF}.
@@
 
-ACS for a multivariate PS is wide variation in the CEP over subgroups defined by the set of biomarkers.
-This is difficult to display graphically due to the increased dimension of the surrogate.
-We propose the use of functions based on the CEP rather than the set of biomarker values for display of ACS.
-We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate ACS.
-We also suggest a summary statistic, the standardized total gain, that quantifies both our suggested accuracy concept and ACS in a single measurement.
+ For a multivariate PS the concept of wide variation in the CEP over subgroups defined by the set of biomarkers is clear.
+However, this is again difficult to display graphically due to the increased dimension of the surrogate.
+We propose the use of functions based on the CEP rather than the set of biomarker values for display of the wide variation criterion.
+We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate wide variation.
+We also suggest a summary statistic, the standardized total gain, that quantifies both our suggested accuracy concept and wide variation in a single measurement.
 Details for standardized total gain are given in Section \ref{STG}.
@@
 \item A2: Ignorable Treatment Assignment
-\item A3: Equal drop out and individual clinical risk up to time $\tau$: $\min\{T(1), C(1)\} \leq\tau$ if and only if $\min\{T(0), C(0)\} \leq\tau$
+\item A3: Equal clinical risk up to time $\tau$: $T(1) \leq\tau$ if and only if $T(0) \leq\tau$
 \item A4: Random censoring: $T(z) \perp C(z)$ for $z\in \{0,1\}$.
@@
 Assumption A2, is generally true in randomized, blinded trials.
-Assumption A3 aides in the identification of subjects with min$(T,C) < \tau$ to be excluded from the evaluation cohort without causing bias.
-Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced events or drop-out over the trial arms prior to $\tau$.
+Assumption A3 aides in the identification of subjects with $T < \tau$ to be excluded from the evaluation cohort without causing bias.
+Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced event rates over the trial arms prior to $\tau$.
 Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used.
@@
 
-One can test the null hypothesis of no surrogate value or no ACS, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$.
-If any of the $S_{J}(1)$ associated terms significantly different from zero then there is evidence of variation in the risk difference over the candidate PS.
+One can test the null hypothesis of no surrogate value or no variation in the marginal-CEP, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$.
+If any of the $S_{j}(1)$ associated terms significantly differ from zero then there is evidence of variation in the risk difference over the candidate PS.
 
@@
 \end{eqnarray}
-among those treated subjects with all $S(1)_{(j,k)}$ and $W$ measured. We denote the number of subjects in this validation group by $n_V$.
+among those treated subjects with all $S_{(j,k)}(1)$ and $W$ measured. We denote the number of subjects in this validation group by $n_V$.
 
@@
 \end{itemize}
-Assumption A6 is needed to identify the classification accuracy measures, but monotonicity does not have to imply that the treatment is never harmful and in some cases the alternative assumption that the treatment never helps is more likely.
+Assumption A6 is needed to identify the classification accuracy measures.
 Assumption A6 is not fully testable, although it can be rejected.
-Given Assumptions A1-A5, there are clear testable implications of Assumption A6, such as $P(T\leq t|Z=1)\geq P(T\leq t|Z=0)$ for all $t$ if no harm of treatment is assumed, $P(T \leq t|Z=0) \geq P(T\leq t|Z=1)$ for all $t$ if no benefit of treatment is assumed, and no switch in sign of the model-based estimate of risk difference over the range of the observed values of the $S_j(1)$s.
+Given Assumptions A1-A5, there are clear testable implications of Assumption A6, such as $P(T\leq t|Z=1)\geq P(T\leq t|Z=0)$, and no switch in sign of the model-based estimate of risk difference over the range of the observed values of the $S_j(1)$s.
 
 Assumption A6 may be more plausible in settings where the outcome is a specific illness or infection rather than death from any cause.
-We assume without loss of generality the no-harm of active treatment form of monotonicity heretofore, $T(0) \leq T(1)$, as this is the form of the assumption that is more likely to apply to our motivating example.
-Assuming A1-A6, $TPF(t|c)$ can be defined as:
+Assuming A1-A6, $TPF(t|c)$ can be identified and estimated as we define it by:
 \begin{eqnarray*}
@@
 However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities.
-Greater discussion of this point is given the Appendix A of the supplementary materials.
+Greater discussion of this point is given in Appendix A of the supplementary materials.
 When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued.
@@
 As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times.
-We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the distribution function of the event times.
-The estimated marginal distribution of event times $\hat{F}(T)$ is calculated by integrating over $Z$ and $S$ in the conditional CDF defined by the assumption A5.
+We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T.
+The estimated marginal distribution of event times $\hat{F}(T)$ can be calculated in a number of ways, by integrating over $Z$ and $S$ in the conditional CDF defined by the assumption A5 or Kaplan-Meier estimator.
 Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
@@
 \subsection{Evaluation and Comparison of Candidate PS} \label{DEF}
-Testing null hypothesis $H0_1$,  no surrogate value, using the biomarker associated coefficients from the risk estimation model as outlined in section \ref{CRE} is a first step in evaluation or comparing biomarkers as PS.
-This test can also be used to help determine what biomarkers to include in a multivariate surrogate.
-One should first consider each biomarker as a univariate surrogate, eliminating those from consideration that are found to have no evidence of variation in the CEP-curve by $H0_1$.
+Testing the null hypothesis $H0_1$, no surrogate value, using the biomarker associated coefficients from the risk estimation model as outlined in section \ref{CRE} is a first step in evaluation or comparing biomarkers as PS.
+This test can also be used to help determine which biomarkers to include in a multivariate surrogate.
+One approach is to consider each biomarker as a univariate surrogate, eliminating those from consideration that are found to have no evidence of variation in the CEP-curve by $H0_1$.
 Then, combinations of the biomarkers can be considered as multivariate candidate surrogates, giving greater consideration to combinations of univariate candidates that were found to have some evidence of surrogate quality.
-The coefficient test of  null hypothesis $H0_1$ can again be used to reduce the dimension of multivariate surrogates, as a biomarker $j$ for which the coefficients $\gamma_{0j}=\gamma_{1j}=0$ in a multivariate surrogate can be considered to not add value to surrogate.
-Once the field of consideration has been narrowed to considered only candidates with at least some surrogate quality, PS can be further evaluated based on their $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ and the $STG(t)$ versus $t$  curves.
+The coefficient test of  null hypothesis $H0_1$ can again be used to reduce the dimension of multivariate surrogates, as a biomarker $j$ for which the coefficients $\gamma_{0j}=\gamma_{1j}=0$ in a multivariate surrogate can be considered to not add value to the surrogate.
+Once the field of consideration has been narrowed to consider only candidates with at least some surrogate quality, PS can be further evaluated based on their $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ and the $STG(t)$ versus $t$ curves.
 These curves can also help to give a visualization of a ranking of the candidates.
 
-The $STG(t)$ or $\widetilde{STG}$ difference between each of the candidates that are evaluated to have similarly high value as PS can be used to determine if there is statistical evidence to support the superiority of one of the candidates.
+The $STG(t)$ or the $\widetilde{STG}$ difference between each of the candidates that are evaluated to have similarly high value as PS can be used to determine if there is statistical evidence to support the superiority of one of the candidates.
 The choice of time-point for comparison should be motivated by the scientific question of interest.
-For example, If one year treatment efficacy is of primary interest, a $t$ of one year less $\tau$ would be a logical choice from comparison.
-If there is no time point of particular scientific interest, inference should be based on $\widetilde{STG}$ difference and $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ should be considered at the longest follow-up time less $\tau$.
+If there is no time point of particular scientific interest, inference should be based on the $\widetilde{STG}$ difference and $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ should be considered at the longest follow-up time less $\tau$.
 As the CDF based risk is cumulative, a $t$ that is equal to the longest follow-up less $\tau$ includes all the available outcome information and allows for the best investigation of durability both of surrogacy and treatment effect.
@@
 When A6 does not hold, the candidate surrogate with the highest $\widetilde{STG}$ and a $CDF^{t}_{\Delta}(c)$ curve that is superior to all other candidates for at least some $c$, has the most value as a PS among the candidates considered.
-If this candidate is found to be statistically significantly superior to some or all the other candidates this provides greater evidence of superiority.
+If this candidate is found to be statistically significantly superior to some or all of the other candidates this provides greater evidence of superiority.
 However, as with almost all statistical tests, a failure to reject the null does not prove the null; lack of significant  $\widetilde{STG}$ difference does not prove lack of superiority.
@@
 500 replicates of the simulation were run for each scenario.
-Some simulations did not converge and were removed from the simulations used for bias and power calculations; there was a 2.5\% non-convergence rate.
-All candidate PS in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.
+All candidate PSs in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.
 
@@
  & 0.5 & 0.059 (0.099) & 0.052 (0.109) & 0.006 (0.053)& 0.007 (0.076)& 0.076 (0.224) & 0.024 (0.153) \\
-\% Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
+Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
   &0.5& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
@@
 Nephropathy is often defined by a high albumin excretion rate, as micro-albuminuria (defined as albumin excretion rate $>$ 30 mg/24 hr) is the least invasive gold standard for kidney damage.
-The trial ended early in 1993 due to strong evidence of treatment efficacy, with a longest follow-up time of 7.5 years after the biomarker measurments of interest; estimated adjusted mean risk of micro-albuminuria was reduced by 56\% (P-value 0.01)\citep{DCCT}.
+The trial ended early in 1993 due to strong evidence of treatment efficacy, with a longest follow-up time of 7.5 years after the measurement of the biomarkers of interest; estimated adjusted mean risk of micro-albuminuria was reduced by 56\% (P-value 0.01)\citep{DCCT}.
 
@@
 The current analysis includes all participants who were free from micro-albuminuria at baseline (n = 1035); baseline micro-albuminuria is balanced over the arms of the trial.
+We use the time-point year 7.5 for our main analysis because it is the longest follow-up time in the trial.
 
@@
 This is consistent with the error for both being random.
-As a sensitivity analysis, we repeat the analysis with the non-truncated measurements and found the the results were similar, no reported.
+As a sensitivity analysis, we repeat the analysis with the non-truncated measurements and found the the results were similar (not reported).
 
@@
 The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.055.
-The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has an estimated P-value for testing H0$_1$ of 0.018.
+The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of 0.018.
 Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone.
 The estimated $\widehat{\widetilde{STG}}$ is 0.2 for HBA1C alone and 0.15 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.202.
-Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing of any of the suggested summary statistics.
+Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.
 
-The P-value for the test that $\widehat{\widetilde{STG}}$ difference between HBA1C alone and the univariate the linear combination of HBA1C-EGFR is zero, is 0.057.
+The P-value is 0.057 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR.
 This is evidence that HBA1C alone is superior to the univariate the linear combination of HBA1C-EGFR.
@@
 The EGFR curve is most similar to the marginal risk difference line, supporting the finding that EGFR has little value as a PS.
-The multivariate combination surrogate which is flatter than the EGFR alone curve is very similar to HBA1C curve.
+The multivariate combination surrogate which is flatter than the EGFR alone curve is very similar to the HBA1C curve.
 This provides further evidence that the combination candidate does not improve upon HBA1C alone.
-We use time-point year 7.5 for our main analysis because it is the longest follow-up time in the trial.
 As there is no evidence of violation of A6, Figure \ref{exp} Panel B displays the $\widehat{ROC}^{7.5}(q)$ versus $q$ curves for the HBA1C and the two combination candidates.
@@
 
-A naive evaluation of the association of the two univariate candidate surrogates and outcome separately in each arm
-of the trial, using a Cox proportional hazards model and the hazard ratio magnitude for ranking, gives the same ranking of candidate surrogates in the treatment arm as do our methods.
-The findings support our conclusion that EGFR has no association with outcome under treatment.
+A naive evaluation of the association of the two univariate candidate surrogates and outcome separately in each arm of the trial, using a Cox proportional hazards model and the hazard ratio magnitude for ranking, gives the same ranking of candidate surrogates in the treatment arm as do our methods.
+The findings support our conclusion that EGFR has no association with outcome in the treatment arm.
 However, in the control arm EGFR is significantly associated with outcome and HBA1C is not.

### 2014-06-18 Compare_053014_statsmed_revisions.tex
Updated tables and analysis using KM estimate of integrated STG and new simulation results

@@
 We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T.
-The estimated marginal distribution of event times $\hat{F}(T)$ can be calculated in a number of ways, by integrating over $Z$ and $S$ in the conditional CDF defined by the assumption A5 or Kaplan-Meier estimator.
+The estimated marginal distribution of event times $\hat{F}(T)$ can be reliably calculated by the Kaplan-Meier estimator.
 Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
@@
   &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.025 (0.077) & 0.043 (0.158) \\ 
-Bias $STG.i$ & 0.8 & 0.065 (0.079)& 0.021 (0.081)& 0.023 (0.051)& 0.003 (0.067) & 0.009 (0.075) & 0.041 (0.074) \\
- & 0.5 & 0.059 (0.099) & 0.052 (0.109) & 0.006 (0.053)& 0.007 (0.076)& 0.076 (0.224) & 0.024 (0.153) \\
+Bias $\widetilde{STG}$ & 0.8 & 0.063 (0.050)& 0.013 (0.077)& 0.019 (0.101)& 0.050 (0.028) & 0.023 (0.093) & 0.032 (0.054) \\
+ & 0.5 & 0.092 (0.073) & 0.041 (0.102) & 0.034 (0.053)& 0.055 (0.028)& 0.022 (0.095) & 0.069 (0.092) \\
 Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
@@
 \begin{center}
-\caption{Power of difference in $\widetilde{STG}$ tests for $t \in (\tau, 3.5)$ years \label{IPow}}
+\caption{Power of difference in $\widetilde{STG}$ tests for $t \in (\tau, \infty)$ years \label{IPow}}
 \begin{tabular}{rrccccc}
@@
   \hline
-$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 75.6 & 24.6 & 56 \\ 
-&  0.5  & 7.2 & 37.6 & 91.8 & 93.9 & 97.4 \\ 
- $\widetilde{STG}=0.25$ & 0.8 &  & 4 & 63.6 & 47.2 & 37.6 \\ 
-&  0.5 &  & 4 & 78.8 & 80 & 94.8 \\ 
-$\widetilde{STG}=0.47$ & 0.8 &  &  & 6 & 7.5 & 8.8 \\ 
- & 0.5 &  &  & 6 & 2.8 & 16.2 \\ 
-$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 24.2 \\ 
-&  (0.5, 0.5) &  &  &  & 6.3 & 17.2 \\ 
-$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 1.6 \\ 
- & (0.5, 0.5) &  &  &  &  & 6 \\ 
+$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 99.6 & 7.8 & 99.2 \\ 
+&  0.5  & 5.7 & 17.8 & 99.8 & 9.2 & 99.4 \\ 
+ $\widetilde{STG}=0.25$ & 0.8 &  & 5.4 & 91.6 & 6.6 & 83.6 \\ 
+&  0.5 &  & 5.5 & 95.2 & 7.1 & 89.8 \\ 
+$\widetilde{STG}=0.47$ & 0.8 &  &  & 4.9 & 82.6 & 8.8 \\ 
+ & 0.5 &  &  & 5.2 & 81.9 & 16.2 \\ 
+$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 75.8 \\ 
+&  (0.5, 0.5) &  &  &  & 6.3 & 78.0 \\ 
+$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 5.6 \\ 
+ & (0.5, 0.5) &  &  &  &  & 5.9 \\ 
    \hline
@@
 Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone.
-The estimated $\widehat{\widetilde{STG}}$ is 0.2 for HBA1C alone and 0.15 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.202.
+The estimated $\widehat{\widetilde{STG}}$ is 0.31 for HBA1C alone and 0.21 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.35.
 Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.
 
-The P-value is 0.057 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR.
+The P-value is 0.058 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR.
 This is evidence that HBA1C alone is superior to the univariate the linear combination of HBA1C-EGFR.
-As the difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS is so small, it is not surprising that the P-value for the null that their difference is equal to zero is 0.99.
+The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS gives a p-value of 0.18.
 

### 2014-07-17 response-to-review-new.tex
Revisions, vector W, still needs new simulation tables

% !TEX TS-program = pdflatex
% !TEX encoding = UTF-8 Unicode

% This is a simple template for a LaTeX document using the "article" class.
% See "book", "report", "letter" for other types of document.

\documentclass[11pt]{article} % use larger type; default would be 10pt

\usepackage[utf8]{inputenc} % set input encoding (not needed with XeLaTeX)

%%% Examples of Article customizations
% These packages are optional, depending whether you want the features they provide.
% See the LaTeX Companion or other references for full information.

%%% PAGE DIMENSIONS
\usepackage{geometry} % to change the page dimensions
\geometry{a4paper} % or letterpaper (US) or a5paper or....
% \geometry{margin=2in} % for example, change the margins to 2 inches all round
% \geometry{landscape} % set up the page for landscape
%   read geometry.pdf for detailed page layout information

\usepackage{graphicx} % support the \includegraphics command and options

% \usepackage[parfill]{parskip} % Activate to begin paragraphs with an empty line rather than an indent

%%% PACKAGES
\usepackage{booktabs} % for much better looking tables
\usepackage{array} % for better arrays (eg matrices) in maths
\usepackage{paralist} % very flexible & customisable lists (eg. enumerate/itemize, etc.)
\usepackage{verbatim} % adds environment for commenting out blocks of text & for better verbatim
\usepackage{subfig} % make it possible to include more than one captioned figure/table in a single float
% These packages are all incorporated in the memoir class to one degree or another...

%%% HEADERS & FOOTERS
\usepackage{fancyhdr} % This should be set AFTER setting up the page geometry
\pagestyle{fancy} % options: empty , plain , fancy
\renewcommand{\headrulewidth}{0pt} % customise the layout...
\lhead{}\chead{}\rhead{}
\lfoot{}\cfoot{\thepage}\rfoot{}

%%% SECTION TITLE APPEARANCE
\usepackage{sectsty}
\allsectionsfont{\sffamily\mdseries\upshape} % (See the fntguide.pdf for font help)
% (This matches ConTeXt defaults)

%%% ToC (table of contents) APPEARANCE
\usepackage[nottoc,notlof,notlot]{tocbibind} % Put the bibliography in the ToC
\usepackage[titles,subfigure]{tocloft} % Alter the style of the Table of Contents
\renewcommand{\cftsecfont}{\rmfamily\mdseries\upshape}
\renewcommand{\cftsecpagefont}{\rmfamily\mdseries\upshape} % No bold!

%%% END Article customizations

%%% The "real" document content comes below...

\title{Response to Referee’s report of SIM submission 13-0382.R1 entitled "Comparing and
combing biomarkers as principle surrogates for time-to-event clinical
endpoints"}
\author{Erin E. Gabriel, Michael C. Sachs, Peter B. Gilbert}
%\date{} % Activate to display a given date or no date (if empty),
         % otherwise the current date is printed

\begin{document}
\maketitle
We are happy to have the opportunity to revise our manuscript. We have put forth considerable effort to address reviewer 2's comments and we feel that the paper has improved greatly.

We have made some critical revisions to the methods section, and have modified the real-data example to better illustrate our proposed methods.  We have developed a new summary statistic that does not require the selection of a specific time. During the development of this new summary statistic, we noticed a convergence issue for the multivariate PS estimands, that can lead to poor efficiency in practice. This issue was previously masked by our truncation of the STG for a given time point. We now consider using a multivariate BIP, one independent BIP for each candidate PS, that appears to solve the convergence issues.

We have spent several months attempting to locate an adequate new example data set. Although we were unable to do so, we tried several, including the Mr. FIT data, and several unpublished small immune marker trial of NIAID. In each case there was inadequate baseline information to support the estimation of the counterfactual estimands. However, in running each of these examples, we determined that the use of truncated normal in the simulations did not fit what we were suggesting in practice for the example. We have modified these simulations to use censored normal generation and censored imputation. Below we outline the specific changes.

\section{Comments from Reviewer 2}
\begin{quotation}
I have to admit that I am somewhat disappointed with this revision.

Only minor changes in the essential section 2 have been made.

The paper is still not sufficiently self-contained and one has to read at least [1] and [10] to have a chance at understanding what’s going on in the paper. Both [1] and [10] are very nice and interesting papers by the way.

The extension from [10] in terms of modelling is straightforward and minor.

The tools developed to assess and select one or more surrogates are problematic at least to me as they include the subjective choice of one or more time points at which performance is evaluated. Also, the actual selection of an optimal subset of surrogates is not well described.

It would be ideal with an example where these extensions really mattered.
\end{quotation}

\section{Response to Reviewer 2}
\begin{enumerate}
\item We reorganized the methods into sections 2 and 3. Section 2 now contains the notation and background information on the concept of principal surrogacy. This section provides a more thorough introduction to the methods so that uninitiated readers may better understand without referring to [1] or [10]. 
\item The remaining subsections of the methods have been reorganized so that the concepts flow more naturally from defining estimands of interest, to assumptions, to modeling, to estimation, to summary statistics, and finally to PS comparison. 
\item We have added an additional subsection where we introduce the integrated standardized total gain to address reviewer 2's concerns about the subjective choice of time points.  The integrated STG does not require the analyst to specify a time point, but rather averages the time dependent STG with respect to the distribution of event times. We suggest estimating the marginal distribution of event times with the Kaplan-Meier estimator. We evaluate the integrated STG in the simulations and demonstrate its operating characteristics. The integrated STG tends to have slightly higher power for comparison of candidate PS as compared to the STG at a single time point. 
\item In regard to the reviewer's comment about selecting the optimal subset of surrogates, we have revised section 3.7 to more clearly describe the hypothesis tests that are used to rule out useless PS from consideration and for comparing and ranking candidates with some value. 
\item We have modified the example to better illustrate our extensions. We created a third candidate PS from a linear combination of the two real candidate PS such that the PS has some value as a surrogate, but with a significantly smaller STG. In this case the test for the difference in STG yields evidence of a difference between the two surrogates where standard hypothesis testing could not. This makes the example more like a simulation, in that we modified the data to generate the desired results, but we feel that it does provide a better illustration of where our proposed methods actually matter. 
\item We have added to the multivariate simulations the investigation of the use of a multivariate BIP. 
\item We have added updated the simulations to reflect our suggested censoring of a candidate surrogate, as done in the example, rather than truncation. 
\end{enumerate}

We thank you again for the thoughtful review; the paper has improved greatly thanks to your efforts.

\vspace{2em}

\hfill{Sincerely,}
\vspace{2em}

\hfill{Erin E. Gabriel}

\end{document}

### 2014-07-17 supp-power-table.tex
Revisions, vector W, still needs new simulation tables

--- 2014-06-18 Compare_053014_statsmed_revisions.tex
@@
-\documentclass[times, doublespace]{simauth}
-
-\usepackage{moreverb}
-\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
- \usepackage[numbers, square, sort&compress]{natbib}
-
-\begin{document}
-% The following line would print the thesis in 
-%opening
-
-\runninghead{EE Gabriel and MC Sachs and PB Gilbert}
-\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints}
-\author{Erin E. Gabriel\affil{a}\corrauth\
-Michael C. Sachs\affil{b}, Peter B. Gilbert\affil{a,c}}
-\address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
-\affilnum{b}Kidney Research Institute, University of Washington\\
-\affilnum{c}Department of Biostatistics, University of Washington}
-\corraddr{egabriel@fhcrc.org}
-\begin{abstract}
-Principal surrogate endpoints are useful as targets for Phase I and II trials.
-In many recent trials, multiple post-randomization biomarkers are measured.
-However, few statistical methods exist for comparison of or combination of biomarkers as principal surrogates and none of these methods to our knowledge utilize time-to-event clinical endpoint information.
-We propose a Weibull model extension of the semi-parametric estimated maximum likelihood method of \citet{Huang11} that allows for the inclusion of multiple biomarkers in the same risk model as multivariate candidate principal surrogates.
-We propose several methods for comparing candidate principal surrogates and evaluating multivariate principal surrogates.
-These include the time-dependent and surrogate-dependent true and false positive fraction, the time-dependent and the integrated standardized total gain and the cumulative distribution function of the risk difference.
-We illustrate the operating characteristics of our proposed methods in simulations and outline how these statistics can be used to evaluate and compare candidate principal surrogates.
-We use these methods to investigate candidate surrogates in the Diabetes Control and Complications Trial.
-\end{abstract}
-\keywords{Causal inference; Surrogate endpoint evaluation; Survival analysis; Accuracy measures; Multivariate principal stratification}
-\maketitle
-
-\section{Introduction} 
-A valid surrogate endpoint can be used as a primary outcome for evaluating treatments in phase I-II trials, by virtue of providing reliable prediction of treatment effects on a clinical endpoint of interest.
-In 2002, \citet{Frangakis02} introduced the principal stratification framework and a definition of a principal surrogate (PS).
-Several methods for evaluation of principal surrogates have been developed under this framework, eg.
-\citep{Taylor05, Follmann06, Gilbert08, Wolfson10, Li10, Huang11, Huang12, Zigler12, Gabriel13}.
-We extend this literature in several ways.
-
-We combine the estimation methods of \citet{Gabriel13} and  \citet{Huang11}  to allow for the evaluation of multivariate candidate principal surrogates of continuous time-to-event clinical endpoints.
-Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint.
-\citet{Gabriel13} extended methods for evaluating univariate a biomarker as a PS of a continuous time-to-event clinical endpoints.
-\citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.
-Our proposed method uses a semi-parametric estimated maximum likelihood fit for a Weibull time-to-event model.
-
-We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves.
-A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero.
-A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement.
-Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect.
-This association can be displayed by the marginal-CEP versus candidate PS curve \citep{Gilbert08}.
-Evaluation of multivariate principal surrogates is complicated by the increased dimension.
-
-The dimension of the marginal-CEP versus candidate PS curve increases with the dimension of the candidate surrogate, making the marginal-CEP curve difficult to display and interpret for multivariate candidate PS.
-If instead we consider the clinical outcome within levels of the marginal-CEP, we can define useful curves that are two-dimensional regardless of the number of biomarkers included in the risk model.
-Our proposed curves include the marginal-CEP-based and time-dependent true positive fraction ($TPF(t|c)$) and false positive fraction ($FPF(t|c)$) as well as the receiver operating characteristic curve ($ROC^{t}(q)$) they define.
-We also consider the cumulative distribution function (CDF) of the marginal CEP function.
-
-We develop summary statistics for the comparison of candidate PS for a continuous time-to-event endpoint.
-Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS .
-Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality.
-We develop two extensions of the standardized total gain \citep{Huang11}, both of which provide a single value measure of surrogate quality that can be used to compare candidate surrogates for the same clinical outcome in the same trial.
-The time-dependent standardized total gain, ($STG(t)$), allows for the comparison of candidate PS at a given time point $t$ that properly accounts for censoring.
-Over a range of time points, a display of $STG(t)$ versus $t$ provides information about surrogate quality waning, as well as a means of comparison of candidate surrogates over time.
-The time integrated  standardized total gain, ($\widetilde{STG}$), which is the time-dependent standardized total gain integrated over the estimated survival times, provides a single value summary of surrogate quality independent of time.
-
-In Section 2, we give some necessary notation and discuss in greater detail why the existing criteria for principal surrogate evaluation are not ideal for the evaluation of multivariate candidate PS or the comparison of candidate PS.
-In Section 3.1, we describe the time-dependent risk estimands and outline some assumptions that are helpful for identifying these estimands.
-We outline our proposed extension to the estimation methods of \citet{Huang11} and \citet{Gabriel13} in Section 3.2.
-We describe our proposed evaluation methods in Sections \ref{CAM}-\ref{CDF} and in Sections \ref{STG}-\ref{ISTG} we detail our proposed summary statistics for comparison.
-In Section 3.7, we outline our suggested procedure for biomarker comparison using the methods developed in Sections \ref{CAM}-\ref{ISTG}.
-In Section 4 we evaluate our methods in a simulation study of both univariate and multivariate candidate PS.
-In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT).
-We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combinations of the two candidates are not improvements over the higher quality surrogate alone.
-In Section 6, we discuss potential limitations of our methods and future avenues for research.
-
-\section{Notation and Criterion for Principal Surrogacy} \label{DEF}
-Let $Z$ be the treatment indicator, 0 for control/non-active treatment and 1 for treatment.
-Let $W$ be a vector of baseline measurements taken prior to randomization.
-In the principal stratification framework of \citet{Frangakis02} we use potential outcomes, where all post-randomization measures are considered under assignment to either treatment arm for each individual.
-Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z\in \{0,1\}$.
-Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=min(T(z),C(z))$.
-Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z\in \{0,1\}$.
-Let $Y_{i}^{t}(z)$ be the indicator that $T_i(z)\leq t$.
-Let $D_i(t)=Y_{i}^{t}(0)-Y_{i}^{t}(1)$ be the individual treatment effect for subject $i$ on the clinical endpoint at or before time $t$.
-Let $\{S_{1,i}\ldots,S_{J,i}, T_i, C_i, Y_i, Y_{i}^{t} \}$ denote the observed values of the potential outcomes for subject $i$.
-
-All $S_j$ are measured at the same fixed post-randomization time point $\tau$.
-If $T_{j,i}(z)$ is less than $\tau$ for a given subject $i$, all $S_{j,i}(z)$ are undefined for that subject, as observation of the post-randomization pre-clinical-event measures $S_{j}(z)$ are no longer possible under either treatment condition.
-For this reason, subjects with $T< \tau$ are excluded from the evaluation cohort.
-Let $M_j$ be the indicator that $S_{j}(1)$ is observed, and let $s_{1,j,i}$ be the realization of $S_{j}(1)$ for subject $i$.
-We assume the full set of observed and counterfactual outcomes are independently and identically distributed over the trial subjects $i= 1, \ldots n$.
-Let $F_{S_1(1), \ldots, S_J(1)}(\cdot,\dots,\cdot)$ and $F_{S_1(1), \ldots, S_J(1)|W}(\cdot|\cdot)$ be the joint CDF of the $S_1(1), \ldots, S_J(1)$ and joint CDF of the $S_1(1), \ldots, S_J(1)$ conditional on $W$, respectively.
-Define $\rho_z(t)=Pr(Y^t(z)=1)=E(Y^t(z))=Pr(T(z)\leq t)$ to be the marginal prevalence of the potential clinical outcome at or before time $t$ for treatment arm $z$.
-Our risk estimands of interest, $risk_z(s)$, are functions of the candidate surrogate that give some measure of risk under the treatment $z$.
-
-\citet{Frangakis02} developed the concept of principal surrogacy and defined a PS by a criterion based on joint risk estimands, which conditions on the candidate surrogate under active treatment and under control, $risk_z(s_1,s_0)$.
-If $risk_1(s_1,s_0)=risk_0(s_1,s_0)$ for all $s_1 = s_0$, S is a principal surrogate; \citet{Gilbert08} called this average causal necessity (ACN).
-Subsequent work added a second criterion that a good PS should strongly modify the clinical treatment effects over the $(s_1, s_0)$ subgroups, the wide variability criterion.
-This can be demonstrated by investigating contrasts of $risk_1(s_1,s_0)$ and $risk_0(s_1,s_0)$, such as the ratio, over the levels of $(s_1, s_0)$.
-ACN is a sensible condition that can help eliminate candidate PS, but is not a sufficient condition, as was shown in \citet{Vanderweele11}.
-The combination of wide effect modification and ACN are often used for evaluation in the current literature.
-The wide effect modification criterion may also suggest a possible ranking of candidate surrogates based on the level of treatment efficacy modification over the subgroups defined by the surrogate.
-
-In \citet{Gilbert08} and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate.
-marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$.
-The concept of wide variability is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$.
-This can simply be viewed as a subgroup analysis, where subgroups defined by levels of $S(1)$, should strongly modify the clinical treatment effect.
-Other than the special case when the candidate surrogate is constant in the control arm, $S(0)=C$ called Constant Biomarker (CB), ACN cannot be verified based on the marginal risks alone \citep{Gilbert08, Gabriel13}.
-\citet{Gabriel13} outlines some PS candidate constructions that can make CB more likely to hold in a general clinical trial where biomarkers under control will tend not to be constant.
-
-For a univariate candidate PS ACN can be displayed using the joint CEP, or marginal CEP under case CB.
-Although the concept of ACN is easily extended to a multivariate PS, $risk_1(s_{1,1}, \dots, s_{1,J} ,s_{0,1}, \dots, s_{0,J})=risk_0(s_{1,1}, \dots, s_{1,J} ,s_{0,1} \dots s_{0,J})$ for all $s_{1,j} = s_{0,j}$ with $j \in \{1, \dots, J\}$, it is a more difficult concept to display graphically.
-As well, the definition does not seem as sensible in the multi-dimensional case.
-Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates.
-However, ACN may not guarantee this is the case \citep{Vanderweele11}.
-We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional on a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$.
-For this purpose, we propose the use of a set of classification accuracy measures, outlined in Sections \ref{CAM} and \ref{CDF}.
-Like ACN, the classification accuracy measures allow for the quantification of the agreement between the treatment effect on the clinical outcome and the predictive model when there is no treatment effect on any of the biomarkers in the model.
-Unlike ACN, the measures allow for quantification of this agreement for all other levels of the biomarkers as well.
-These accuracy measures can also be displayed as curves versus the levels of the predictive model to allow for 2-dimensional  illustration of the predictive accuracy of the risk model.
-We suggest that good agreement between the predictive model containing the set of biomarkers and the clinical treatment efficacy is necessary but not sufficient for a set of biomarkers to have value as a PS.
-Much like ACN, it is possible to have good agreement between the model prediction and the treatment efficacy simply because there is no efficacy.
-
- For a multivariate PS the concept of wide variation in the CEP over subgroups defined by the set of biomarkers is clear.
-However, this is again difficult to display graphically due to the increased dimension of the surrogate.
-We propose the use of functions based on the CEP rather than the set of biomarker values for display of the wide variation criterion.
-We propose the CDF of the CEP function described in Section \ref{CDF} for graphical display of multivariate wide variation.
-We also suggest a summary statistic, the standardized total gain, that quantifies both our suggested accuracy concept and wide variation in a single measurement.
-Details for standardized total gain are given in Section \ref{STG}.
-As well, if one assumes a parametric model for risk conditional on a set of biomarkers then a test for the minimum amount variation in the CEP over subgroups defined by the biomarkers can be based on the coefficients of the biomarkers in the model.
-The specifics of this test for our assumed model are given in Section \ref{CRE}.
-
-After a set of candidate multivariate or univariate candidate surrogates is determine to have some value as PS, it is of interest to rank them and test to determine if any one candidate is significantly better than others.
-ANC can only classify candidates into two groups.
-Our suggested accuracy measures and the CDF of the risk model CEP can help more finely group candidates by predictive quality and amount of variation in CEP, but they do not provide inference for comparing candidates.
-We propose the use of the standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for comparison inference.
-These are outlined in Sections \ref{STG} and \ref{ISTG}.
-
-\section{Methods}
-\subsection{Time-Dependent Causal Risk Estimands} \label{CRE}
-In the time-to-event setting there are many ways to define risk.
-The risks may be based on the hazard function, $\lambda_z$, namely, $risk_{z}(t|s_{1,1},\ldots,s_{1,J})\equiv\lambda_z(t|s_{1,1},\ldots,s_{1,J})$, or the conditional CDF.
-The conditional CDF-based risk estimands can be defined as:
-\begin{eqnarray*}
-risk_{1}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_1(t|s_{1,1}, \ldots, s_{1,J})\\
-&\equiv& P\{T(1) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{1}(1)=s_{1,J},T(1)> \tau, T(0)> \tau\},\\
-risk_{0}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) &\equiv& F_0(t|s_{1,1}, \ldots, s_{1,J}) \\
-&\equiv& P\{T(0) \leq t|S_{1}(1)=s_{1,1},\ldots, S_{J}(1)=s_{1,J},T(1)>\tau, T(0)>\tau\}.
-\end{eqnarray*}
-The subscript of a function indicates the value of $z$.
-The comparison between the risk estimands over the treatment arms is the causal comparison of interest for principal surrogate evaluation.
-Our contrast of interest is defined as $\Delta(t|s_{1,1}, \dots, s_{1,J}) \equiv risk_0(t|s_{1,1}, \ldots, s_{1,J})-risk_1(t|s_{1,1}, \ldots, s_{1,J})$, the risk difference.
-Using the CDF based risk estimands, this contrast is causal because both risks condition on the same sets.
-This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}.
-For this reason, we focus on the CDF based marginal risks, which provide well defined  interpretations for our suggested summary statistics.
-
-In order to link these risk estimands to the observable data we must make some assumptions; Assumptions A1-A4 reduce the number of missing potential outcomes and help identify the risk estimands.
-\begin{itemize}
-\item A1: Stable Unit Treatment Value Assumption (SUTVA) and Consistency 
-\item A2: Ignorable Treatment Assignment
-\item A3: Equal clinical risk up to time $\tau$: $T(1) \leq\tau$ if and only if $T(0) \leq\tau$
-\item A4: Random censoring: $T(z) \perp C(z)$ for $z\in \{0,1\}$.
-\end{itemize} 
-Assumption A1 implies that subjects' potential outcomes are independent of other subjects' outcomes and treatment, SUTVA, and observation of the potential outcome does not change its value, consistency.
-For example, $(S_i|z_i=0) \equiv S_i(0)$ and this is independent of $(S_j|z_j=1) \equiv S_j(1)$, for $i \neq j$.
-SUTVA may not hold is some situations, such as cluster randomized vaccine trials where herd immunity may change the treatment effect on all outcomes in the cluster.
-Assumption A2 implies that the distribution of the potential outcomes S(z) and Y(z) are independent of the observed treatment assignment.
-Assumption A2, is generally true in randomized, blinded trials.
-Assumption A3 aides in the identification of subjects with $T < \tau$ to be excluded from the evaluation cohort without causing bias.
-Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced event rates over the trial arms prior to $\tau$.
-Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used.
-Assumptions A1-A4 imply that the conditional distribution of $T(z)$ given $\{S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J}, T(1)\geq\tau, T(0)\geq\tau\}$, equals that for the observed outcome T given the observed treatment and potential candidate surrogate measures $\{Z=z, S_{1}(1)=s_{1,1},\ldots,S_{J}(1)=s_{1,J},T\geq\tau\}$ for $z\in \{0,1\}$.
-
-Given assumptions A1-A4 we can link the time-dependent potential outcome risk estimands of interest to observed data clinical risk estimands and the partially observed candidate surrogates by assuming a parametric survival model.
-We will assume a Weibull model, the probability density function (pdf) for which we will denote as $g(\cdot|\cdot,\cdot)$.
-We parameterize the scale term of this model, $\gamma=(\gamma_{00}, \gamma_{10},\gamma_{0j}, \gamma_{1j})$, and allow for a constant shape term, $\beta_{0}$; this parameterization is equivalent to a classical Weibull survival model.
-The density is given by:
-\begin{eqnarray*}
-g(t|\gamma, \beta, z, s_{1,1}, \ldots,s_{1,J}, y)=\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})^{y} Q_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J}),
-\end{eqnarray*}
-where $Q_z(t|\gamma, \beta, s_{1,j})$ is the conditional Weibull survivor function for treatment arm $Z=z$ and $\lambda_z(t|\gamma, \beta, s_{1,1}, \ldots, s_{1,J})$ is the conditional Weibull hazard function.
-We assume the risk estimands are of the form:
-\begin{itemize}
-\item A5: $risk_{z}^{CDF}(t|s_{1,1}, \ldots, s_{1,J}) \equiv 1-\exp(-(t/\exp(\gamma_{z0}+\sum_j{\gamma_{zj}s_{1,j})})^{(\exp(\beta_{0}))}$
-\end{itemize}
-for $z\in \{0,1\}$.
-We focus on this classical model in the main text for clarity; more complex time-dependent models are outlined in Appendix B of the supplementary materials.
-In order for both assumptions A3 and A5 to hold simultaneously, we can only consider censoring or event times greater than $\tau$.
-This can be accomplished in practice by rescaling all follow-up times from $\tau$ rather than from enrollment.
-
-One can test the null hypothesis of no surrogate value or no variation in the marginal-CEP, $H0_1$, via the estimated model coefficients by a Wald test of the null hypothesis $\gamma_{0j}=\gamma_{1j}=0$ $\forall j \in \{1,\ldots J\}$.
-If any of the $S_{j}(1)$ associated terms significantly differ from zero then there is evidence of variation in the risk difference over the candidate PS.
-
-\subsection{Trial Augmentation and Estimation} \label{EST}
-The assumed parametric form of the risk estimands, A5, condition on the $S_{j}(1)$, which are missing for all subjects assigned non-active treatment $z=0$.
-Several papers have suggested the use of baseline covariates to account for the missing $S_{j}(1)$ values, a strategy called the ``baseline immunogenicity predictor'' (BIP) augmented trial design \citep{Follmann06, Gilbert08}.
-A useful BIP, $W$, is highly correlated with the $S_{j}(1)$ which allows for good prediction of the missing $S_i(1)$.
-Although other trial augmentations have been proposed in the PS literature, the only augmentation available in our motivating data set is BIP, so it will be our focus.
-Using the BIP trial augmentation the observed likelihood given Assumption A5 is:
-\begin{eqnarray*}
-&&\mbox{L}(\beta,\gamma,\nu) \equiv \prod^{n}_i\left\{g(T_i|\gamma, \beta, Z_i, s_{1,1,i},\ldots,s_{1,J,i}, Y_i)\right\}^{\prod_{j=1}^{J}{M_{j,i}}}\\ 
-&&\left\{\int g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, s_{1,l+1},\ldots, s_{1,J}, Y_i)dF_{S_{l+1}(1),\ldots,S_{J}(1)|W}(s_{1,l+1},\ldots,s_{1,J})\right\}^{\{\prod_{j=1}^{l}{M_{j,i}}*\prod_{j=l+1}^{J}{(1-M_{j,i})}\}}\\
-&&\left\{\int g(T_i|\gamma, \beta, Z_i,s_{1,1}, \ldots, s_{1,J}, Y_i)dF_{S_{1}(1),\ldots, S_{J}(1)|W}(s_{1,1},\ldots, s_{1,J})\right\}^{\{\prod_{j=1}^{J}{(1-M_{j,i})}\}}.
-\end{eqnarray*}
-
-This likelihood could be solved by direct optimization if we assumed a parametric model for $S_j(1)|W$.
-However, as proposed in \citet{Pepe91}, consistent estimation of outcome model parameters can be achieved using an estimated distribution of $S_j(1)|W$; this is called estimated maximum likelihood.
-Following \citet{Huang11}, we assume the semi-parametric location-scale model of \citet{Heagerty99} for $F_{S_{j}(1)|W}$.
-Considering first a univariate candidate surrogate, the semi-parametric model is defined as $F_{S_{j}(1)|W} \sim F_j[\{s_{1,j} -\mu_j(w)\} / \sigma_j(w)]= F_j(\varsigma_j)$, where $F_j$ is the CDF of the univariate residuals $\varsigma_j$ from fitting the location and scale parameters $\mu_j(w)$ and $\sigma_j(w)$, respectively.
-We define $\mu_j(w)$ and $\sigma_j(w)$ as parametric functions of the baseline variable(s) $W$, $\mu_j(w) = \gamma_j'w$ and $\log\{\sigma_j(w)\} = \eta_j'w$.
-We can estimate these parameters by solving the equations given in \citet{Heagerty99}:
-\begin{eqnarray}
-&&\sum_{k=1}^{n_V} {w_{k} (s_{(1,j,k)} - \gamma_j' w_{k} )\over \{\exp(\eta_j'w_k)\}^2} =0\\ \label{stuff}
-&&\sum_{k=1}^{n_V} {w_{k}\{(s_{(1,j,k)} - \gamma_j' w_{k})^2 - \{\exp(\eta_j'w_k)\}^2\} \over \{\exp(\eta_j'w_k)\}^2}=0,\label{stuff2}
-\end{eqnarray}
-among those treated subjects with all $S_{(j,k)}(1)$ and $W$ measured. We denote the number of subjects in this validation group by $n_V$.
-
-Fitting the set of parameter equations for the validation subjects yields $n_V$ residuals, $\varsigma_{j,k} \equiv S_{j,k}(1) - \hat{\gamma_j}'w_{k} + \exp(\hat{\eta_j}'w_{k})$.
-Missing counterfactual $S_{j}(1)$ values can be imputed using the parameter estimates and the $n_V$ residuals by, $S_{j,i,k}^*(1) = \hat{\gamma_j}'w_{i} + \exp(\hat{\eta_j}'w_{i}) \varsigma_{j,k}$ for $n_V$ total imputations for each missing $S_{j}(1)$ value.
-We can then use these imputed values $S_{j,i,k}^*(1)$ to estimate the EML contribution for subject $i$ by $\int g(t_i|\gamma, \beta, z_i, s_{1,1}, \ldots, s_{1,J}, y_i)dF_{S_{1}(1),\ldots,S_{J}(1)|W}(s_{1,1},\ldots,s_{1,J}|W)$, by the empirical integral ${\left(1/n_{V}\right)}\sum_k^{n_V}{g_z(t_i|\gamma, \beta,z_i, S_{1,k}^*(1), \ldots,S_{J,k}^*(1),y_i)}$.
-Missingness in the treatment arm, such as subsampling, case-cohort sampling and unplanned missing at random variables, can be imputed in the same way.
-In cases where the validation set is not a random sample of the treatment subjects, the location-scale models can be weighted to accommodate the biased sampling.
-
-To accommodate multivariate surrogates, the same location-scale model is fit for each of the univariate candidate PS separately in the set of validation subjects.
-Imputed values for each missing biomarker are produced separately, but provided that the same baseline variables are used in each model, empirical integration over the set of all imputations is approximately equivalent to integrating over the joint distribution of biomarkers given the baseline variables used for imputation.
-This is an important limitation of this method.
-The BIP need not be based on a single baseline measurement or have the same correlation with each candidate, but the BIP must be the same in each location-scale model for the estimated likelihood to be proper using our procedure.
-This is not a requirement to compare univariate candidate surrogates not included in the same model.
-This gives us an estimated log likelihood of:
-\begin{eqnarray*}
-&&\mathit{l}(\beta,\gamma, \hat{\nu})=\sum_{i} log(g_z(T_i|\gamma, \beta,Z_i, s_{1,1,i}, \ldots,s_{1,j,i},Y_i))*\left\{\prod_{j=1}^{J}{M
-_{j,i}}\right\}+\\
-&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, S^{*}_{l+1,i,k}(1),\ldots, S^{*}_{J,i,k}(1), Y_i))}\right\}*\\
-&&\left\{\prod_{j=1}^{l}{M_{j,i}}*\prod_{j=l+1}^{J}{(1-M_{j,i})}\right\}+\\
-&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g_z(T_i|\gamma, \beta,Z_i, S_{1,i,k}^*(1), \ldots,S_{j,i,k}^*(1),Y_i))}\right\}*\left\{\prod_{j=1}^{J}{(1-M
-_{j,i})}\right\}\cdot
-\end{eqnarray*}
-
-The asymptotic distributional results of \citet{Pepe91} for general EML estimators do not carry over to our setting, due to the zero probability of observing $S_{j}(1)$ in placebo recipients.
-We suggest the bootstrap for inference.
-The assumed form of the risk estimates, A5, can be estimated using the parameters from solving the EML and the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, can be estimated directly from the risk estimates.
-
-\subsection{Classification Accuracy Measures} \label{CAM}
-The time-dependent sensitivity and specificity for the risk difference conditional on $S_{1}(1), \dots, S_{J}(1)$, $\Delta(t|s_{1,1}, \dots, s_{1,J})$, can be defined by adapting \citet{Heagerty00} as:
-\begin{eqnarray*}
-\mbox{Sensitivity}(t|c)&=&P\{\Delta(t|s_{1,1}, \dots, s_{1,J}) > c|D(t)=1\}=TPF(t|c),\\
-\mbox{Specificity}(t|c)&=&P\{\Delta(t|s_{1,1}, \dots, s_{1,J}) \leq c|D(t)=0\}=1-FPF(t|c).
-\end{eqnarray*}
-Here, $TPF(t|c)$ and $FPF(t|c)$ are the time-dependent true and false positive fractions.
-Sensitivity, or true positive fraction, represents the probability that we predict a subject will have benefit using the model and some threshold $c$, $\Delta(t|s_{1,1}, \dots, s_{1,J}) > c$, given that we observe that they did have benefit, $D(t)=1$.
-Similarly, specificity represents the probability that we predict a subject will have no benefit based on our model and using some threshold $c$, $\Delta(t|s_{1,1}, \dots, s_{1,J}) < c$, given that we observe that they did not have benefit, $D(t)=0$.
-This is one minus the false positive fraction, which is the probability that we falsely predict that a subject will have benefit, when they truly did not.
-These probabilities demonstrate how well our surrogate based model predicts benefit from treatment, and are therefore a useful illustration of the usefulness of the surrogate to predict outcome given the assumed model.
-Figure 1 of Appendix A of supplementary materials illustrates the meaning of these classification accuracy measures.
-
-In this case, sensitivity and specificity are based on unobservable data and cannot be estimated empirically.
-However, under assumptions A1-A5 plus the additional assumption of monotonicity of treatment effect, we can estimate them.
-Monotonicity of treatment effect is formally stated as
-\begin{itemize}
-\item A6: $T(z) \leq T(1-z)$ for one of the treatment arms $z$, $z\in \{0,1\}$.
-\end{itemize}
-Assumption A6 is needed to identify the classification accuracy measures.
-Assumption A6 is not fully testable, although it can be rejected.
-Given Assumptions A1-A5, there are clear testable implications of Assumption A6, such as $P(T\leq t|Z=1)\geq P(T\leq t|Z=0)$, and no switch in sign of the model-based estimate of risk difference over the range of the observed values of the $S_j(1)$s.
-
-Assumption A6 may be more plausible in settings where the outcome is a specific illness or infection rather than death from any cause.
-Assuming A1-A6, $TPF(t|c)$ can be identified and estimated as we define it by:
-\begin{eqnarray*}
-TPF(t|c)&=& E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\mathrm{I}_{[\Delta(t|S_{1,1}, \dots, S_{1,J})>c]}\} \over E\{\Delta(t|S_{1,1}, \dots, S_{1,J})\}.
-\end{eqnarray*}
-The derivation of this form is given in Appendix A of the supplementary materials; both the definition and derivation are adaptations of time-independent curves given in \citet{Huang12b}.
-The function $FPR(t|c)$ can be defined similarly based on the expectation of $1-\Delta(t|s_{1,1}, \dots, s_{1,J})$.
-One can estimate these accuracy measures using the plug-in estimators:
-\begin{eqnarray*}
-\widehat{TPF}(t|c)&=& {\int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int \hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})} \hspace{2mm}\mbox{and},\\
-\widehat{FPF}(t|c)&=& {\int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})\mathrm{I}_{\{\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})>c\}}d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J}) \over \int 1-\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}.\\
-\end{eqnarray*}
-Here $\hat{\Delta}(t|s_{1,1}, \dots, s_{1,J})$ is a maximum likelihood estimate under A1-A6 and using the estimated coefficients from the assumed model A5.
-
-If the model for risk difference that conditions on $\{s_{1,1}, \ldots, s_{1,J}\}$, indicates there is a high probability of having a risk difference above a given threshold for all subjects that failed on non-active treatment but not on active treatment at or before time $t$, this is evidence of the predictive power of the model.
-The summary $TPF(t|c)$ is therefore a good measure for the comparison or evaluation of candidate PS.
-Similarly, $\mbox{Specificity}(t|c)$ or $1-FPF(t|c)$, is of interest as those subjects not helped by treatment should have a high probability of having a small risk difference based on a predictive model.
-As these measures are threshold dependent, they can be used to compare not just candidate surrogates but thresholds for a given surrogate.
-Thresholds $c$ found to have high $TPF(t|c)$ and low $FPF(t|c)$ are ideal.
-
-It is of interest to determine the relationship between $TPF(t|c)$ and $FPF(t|c)$.
-The time-dependent ROC curve gives an illustration of the $TPF(t|c)$ and $FPF(t|c)$ relationship at given levels of $c$.
-This curve is defined by $ROC^{t}(q)=TPF\{t|FPF^{-1}(t|q)\}$ for a given time point $t$, adapting the binary-clinical endpoint $ROC(q)$ definition of \citet{Huang12b}.
-Figure \ref{4panel} panels B and D displays the $ROC^{t}(q)$ curves for several different surrogate quality levels.
-The most left-upper point on the $ROC^{t}(q)$ curve is of interest for comparison of candidate surrogates, as candidates with left-upper points closer to $(0,1)$ will tend to be better PS, as elaborated in Section \ref{STG}.
-\begin{sidewaysfigure}[H]
-\begin{center}
-%\includegraphics[scale=1.75, angle=270, clip=TRUE, trim=0 0 20mm 0]{4Panel_trueTE.pdf}
-\includegraphics[scale=0.80]{Figure1.jpeg}
-\end{center}
-\caption{Panel A displays the CDF of the risk difference based on a useless and an ideal univariate surrogate.
-Panel B displays the ROC curves for the same ideal and useless surrogate candidates.
-Panel C displays the CDF of risk difference based on a medium quality univariate partial surrogate, and the CDF of risk difference based on a multivariate surrogate, which is the combination of two medium quality univariate partial surrogates.
-Panel D displays the ROC curves for the same medium quality univariate partial surrogate and multivariate surrogate.
-\label{4panel}}
-\end{sidewaysfigure}
-
-When A6 is observably violated the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased.
-However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities.
-Greater discussion of this point is given in Appendix A of the supplementary materials.
-When there are obvious violations of A6, the estimation of $FPF(t|c)$ and $TPF(t|c)$ should not be pursued.
-When Assumptions A6 is clearly violated we suggest the use of the $\widetilde{STG}$ and $CDF^{t}_{\Delta}(c)$ for comparison and evaluation of candidate surrogates.
-
-\subsection{Cumulative Distribution Function of $\Delta$} \label{CDF}
-
- Let $CDF^{t}_{\Delta}(c)$ be the CDF of the risk difference $\Delta(t|S_{1,1}, \dots, S_{1,J})$, $CDF^{t}_{\Delta}(c)\equiv P\{\Delta(t|S_{1,1}, \dots, S_{1,J})\leq c\}.$ The $CDF^{t}_{\Delta}(c)$ versus $c$ curve is of interest for the evaluation of multivariate candidate PS and for the comparison of candidate PS.
-An ideal PS will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot that is a step function with a horizontal line at the proportion of subjects with $D(t)=-1$ and a horizontal line at the proportion of subjects with $D(t)=0$; where as defined previously $D(t)=Y^{t}(0)-Y^{t}(1)$.
-If we assume monotonicity of treatment effect (assumption A6) then the ideal PS candidate will have a $CDF^{t}_{\Delta}(c)$ versus $c$ plot with a single horizontal line at the proportion of subjects with $D(t)=0$ (in the case that A6 is $T(0) \leq T(1)$).
-If we alternatively assume no benefit, then the line will be at the proportion of subjects with $D(t)=-1$.
-Assumption A6 is not needed for the function $CDF^{t}_{\Delta}(c)$ to be well-defined.
-For a useless surrogate, the plot of $CDF^{t}_{\Delta}(c)$ versus $c$ will be a vertical line at $\rho_0(t)-\rho_1(t)$, the difference in prevalence of the potential clinical outcomes between the trial arms.
-Partially useful surrogates will have $CDF^{t}_{\Delta}(c)$ versus $c$ curves between these two curves with better surrogates having flatter curves.
-Figure \ref{4panel} displays the CDF of the risk difference for several different surrogate quality levels.
-We suggest the plug-in estimator of $\widehat{CDF}^{t}_{\Delta}(c)$, $\widehat{CDF}^{t}_{\Delta}(c)=\int{\mathrm{I}_{[\hat{\Delta}(t|s_{1,1,i}, \dots, s_{1,j,i})\leq c]} d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}$.
-Although we display a useless surrogate in Figure \ref{4panel}, we suggest testing for any surrogate value prior to estimation of $\widehat{CDF}^{t}_{\Delta}(c)$ as it was found in simulations that there is evidence of bias when there is no variation in risk difference; a suggested test is given in Section \ref{CRE}.
-
-Let $R^t(v)$ be the $v$th quantile of $\Delta(t|s_{1,1}, \dots, s_{1,J})$, where $R^t(v)=c$ for $v\in(0,1)$ such that $CDF^{t}_{\Delta}(c)=\{R^t\}^{-1}(v)$.
-The $R^t(v)$ versus $v$ curve has been called the predictiveness curve for the risk difference model.
-This curve has been previously suggested as a means of candidate PS comparison and evaluation \citep{Gilbert08}.
-The range of the curve is restricted by the range of risk difference, which is determined by the prevalence of the potential clinical events of interest in the population.
-Therefore, we prefer the $CDF^{t}_{\Delta}(c)$ versus $c$ curve for displaying the proportion of those subjects for which the treatment was ineffective at a given level of estimated risk difference.
-For characterization of the predictiveness of the model we prefer the classification accuracy measures.
-
-\subsection{Time-dependent Standardized Total Gain} \label{STG}
-It is easily shown that the area under the predictiveness curve, $R^t(v)$ versus $v$, is equal to $\rho_0(t)-\rho_1(t)$.
-For this reason, the area between the quantile curve $R^t(v)$ and the line $\rho_0(t)-\rho_1(t)$ on a plot against $v$ is of interest for evaluation and comparison of candidate PS.
-This area has been called the total gain (TG).
-\citet{Huang11} suggest a standardized version of TG for the comparison of PS for binary clinical endpoints.
-We extend the concept of total gain to the time-to-event setting and define the time-dependent total gain as $TG(t) = \int_0^{1}{|R^t(v) -\{\rho_0(t) -\rho_1(t)\}|dv}$ and the time-dependent standardized total gain as $STG(t) \equiv TG(t)/\{2[\rho_0(t) - \rho_1(t)][1-\rho_0(t) + \rho_1(t)]\}$.
-
-The $STG(t)$ is of interest for comparison of candidate surrogates regardless of the validity of assumption A6 as steeper quantile curves, and thus greater area between $R^t(v)$ and $(\rho_0(t) -\rho_1(t))$, correspond to better prediction of clinical treatment effect and a better candidate surrogate.
-Under A6, $STG(t)$ also has a useful interpretation based on the classification accuracy measures, given by $STG(t)=max_c\{\mbox{Sensitivity}(t|c) + \mbox{Specificity}(t|c)\} -1$.
-This interpretation follows directly from the proof given in \citet{Huang11} for the binary clinical endpoint setting.
-Given this interpretation, $STG(t)$ at a fixed time point is a useful single measurement for evaluation of univariate and multivariate candidate surrogates, as this is equivalent to the upper-most left point on the $ROC^{t}(q)$ curve for the same time point $t$.
-
-We suggest plug-in estimators for $TG(t)$ and $STG(t)$ given by, 
-\begin{eqnarray}
-\widehat{TG}(t)&=&\int{|\widehat{\Delta}(t|s_{1,1}, \dots, s_{1,J}) -\{\hat{\rho}_0(t) -\hat{\rho}_1(t)\}|d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})} \hspace{4mm} \mbox{and} \label{TG} \\
-\widehat{STG}(t)&=&\widehat{TG}(t)/\{2[\hat{\rho}_0(t) - \hat{\rho}_1(t)][1-\hat{\rho}_0(t) + \hat{\rho}_1(t)]\}.
-\end{eqnarray}
-Here $\{\hat{\rho}_0(t) -\hat{\rho}_1(t)\}$ is a maximum likelihood estimate.
-Even when assumption A6 does not hold, the difference in estimated $STG(t)$ can be used for inference to compare candidate PS in the same trial at a particular time point $t$, by Wald test of the null hypothesis $STG_1(t)-STG_2(t)=0$.
-
-As well, the same estimator we use at a fixed time point $t$ can be used to estimate the $STG(t)$ versus $t$ curve, for all $t$ greater than $\tau$ and less than the time of trial closure.
-This curve can be used to investigate time-variation in surrogate quality for a particular surrogate or for comparison.
-Curves, $STG(t)$ versus $t$, that have noticeable downward slopes may indicate that a candidate surrogate modifies the clinical treatment effect less over time, declining surrogate quality, or that the marginal risk difference is declining.
-To better understand the context of declining or increasing $STG(t)$ over time, the marginal risk differences over time should be graphed simultaneously.
-Figure \ref{STG} Panel A depicts a time-independent high quality surrogate and useless surrogate in a trial with declining marginal risk difference, Panel B depicts the same surrogates for a trial with increasing marginal risk difference.
-
-\subsection{Integrated Standardized Total Gain} \label{ISTG}
-As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times.
-We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T.
-The estimated marginal distribution of event times $\hat{F}(T)$ can be reliably calculated by the Kaplan-Meier estimator.
-Then the plug-in estimator of the integrated $STG$ is the expectation with respect to that distribution:
-\[
- \widetilde{STG} = \int \widehat{STG}(t) \, d \hat{F}(t). 
-\]
-The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point.
-By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates.
-The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.
-
-\subsection{Evaluation and Comparison of Candidate PS} \label{DEF}
-Testing the null hypothesis $H0_1$, no surrogate value, using the biomarker associated coefficients from the risk estimation model as outlined in section \ref{CRE} is a first step in evaluation or comparing biomarkers as PS.
-This test can also be used to help determine which biomarkers to include in a multivariate surrogate.
-One approach is to consider each biomarker as a univariate surrogate, eliminating those from consideration that are found to have no evidence of variation in the CEP-curve by $H0_1$.
-Then, combinations of the biomarkers can be considered as multivariate candidate surrogates, giving greater consideration to combinations of univariate candidates that were found to have some evidence of surrogate quality.
-The coefficient test of  null hypothesis $H0_1$ can again be used to reduce the dimension of multivariate surrogates, as a biomarker $j$ for which the coefficients $\gamma_{0j}=\gamma_{1j}=0$ in a multivariate surrogate can be considered to not add value to the surrogate.
-Once the field of consideration has been narrowed to consider only candidates with at least some surrogate quality, PS can be further evaluated based on their $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ and the $STG(t)$ versus $t$ curves.
-These curves can also help to give a visualization of a ranking of the candidates.
-
-The $STG(t)$ or the $\widetilde{STG}$ difference between each of the candidates that are evaluated to have similarly high value as PS can be used to determine if there is statistical evidence to support the superiority of one of the candidates.
-The choice of time-point for comparison should be motivated by the scientific question of interest.
-If there is no time point of particular scientific interest, inference should be based on the $\widetilde{STG}$ difference and $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ should be considered at the longest follow-up time less $\tau$.
-As the CDF based risk is cumulative, a $t$ that is equal to the longest follow-up less $\tau$ includes all the available outcome information and allows for the best investigation of durability both of surrogacy and treatment effect.
-Comparisons over all observed follow-up times greater than $\tau$ can be investigated using the $STG(t)$ versus $t$ plots.
-
-Ideally there will be a single candidate surrogate that has visually superior $CDF^{t}_{\Delta}(c)$, $ROC^{t}(q)$ that also has statistically significantly higher $\widetilde{STG}$ than all other candidates.
-This may not always be the case.
-When assumption A6 holds the candidate surrogate with the highest left-most point of the $ROC^{t}(q)$ curve should also be the candidate surrogate with the highest $STG(t)$.
-When A6 is violated this may not be true.
-When A6 does not hold, the candidate surrogate with the highest $\widetilde{STG}$ and a $CDF^{t}_{\Delta}(c)$ curve that is superior to all other candidates for at least some $c$, has the most value as a PS among the candidates considered.
-If this candidate is found to be statistically significantly superior to some or all of the other candidates this provides greater evidence of superiority.
-However, as with almost all statistical tests, a failure to reject the null does not prove the null; lack of significant  $\widetilde{STG}$ difference does not prove lack of superiority.
-
-\section{Simulation Studies}
-Suppose the conditional CDF of $T$ given $S_j(1)$ and $Z$ follows a Weibull model and $\{S_j(1),W\}$ follows a multivariate truncated normal model with correlation R$_{S_jW}$ between each of the univariate candidates and the BIP and correlation R$_{S_kS_j}$ between each pair of candidates.
-Information lost to drop out occurs completely at random and at a rate of approximately 5\% per year.
-Event times are censored at 3 years after $\tau$, at which time the trials have 50\% cumulative treatment efficacy on average.
-An average of 103 treatment arm infections are observed by 3 years post $\tau$, corresponding to an annual incidence rate of 0.06, whereas an average of 206 control arm infections are observed in the same period, for an annual incidence rate of 0.12 over the 500 simulations.
-
-We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$.
-We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $STG(3)$ of 0.23 and 0.51, respectively.
-We investigate these different levels of candidate surrogates for two different BIP correlations, $R_{s_{1j}w}$, of 0.8 and 0.5.
-For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0.8), (0.5, 0.5)\}$ with a correlation between the candidate surrogates ranging from 0 to 0.3.
-Results for the 0.3 correlation are displayed.
-500 replicates of the simulation were run for each scenario.
-All candidate PSs in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.
-
-Bias results are displayed in Table \ref{Bias} and suggest that the semi-parametric EML method for the Weibull risk difference model is adequately unbiased and that the suggested estimators for $CDF^{t}_{\Delta}(c)$, $TPF(t|c)$, $FPF(t|c)$, $STG(t)$ and $\widetilde{STG}$ have adequate precision and accuracy in all cases for which they are applicable.
-As we have suggested above, there is some bias when there is no surrogate value.
-When there is no surrogate value the estimated model coefficients are still unbiased, with average bias less than 3\% of one Monte Carlo SD (results not displayed).
-The large standard deviation of the $STG(t)$ estimates over all the simulations may imply that this estimator is not as efficient as possible.
-A correlation of 0.5 or higher between a candidate PS and a BIP seems adequate for unbiased estimation of the suggested summary statistics and curves.
-The results for the joint Wald test of any surrogate value, displayed in Table \ref{Bias}, suggest that this test has correct Type 1 error and adequate power in all scenarios.
-Monte Carlo standard errors are used in all Wald tests due to the computational burden of the bootstrap.
-
-Table \ref{Power} displays the power to discriminate between two surrogate quality levels based on the $STG(t)$ difference Wald test for fixed time point $t=3$.
-Monte Carlo standard errors are used and only those candidate surrogates with any surrogate quality are displayed.
-We find that for all null hypothesis scenarios the test has correct type 1 error.
-We also find that the test has adequate power to detect differences in $STG(3)$, particular when there is good correlation between the BIP and the candidate PS, with the power increasing with the difference in $STG(3)$, as expected.
-The test is not valid if either of the candidate surrogates has no surrogate value, as type 1 error is inflated when the true value is on the boundary of the parameter space (results not shown).
-
-Table \ref{IPow} displays the power to discriminate between two surrogate quality levels based on the difference in integrated STG, $\widetilde{STG}$.
-We again find that for all null hypothesis scenarios the test has nearly a nominal 5\% Type 1 error rate with similar to slightly higher power than the $STG(3)$ difference test for the same scenarios.
-
-\begin{sidewaystable}[ht]
-\begin{center}
-\caption{Performance of Accuracy Measures and the Joint Wald Test of the Null Hypothesis of No Surrogate Value H0$_1$ \label{Bias}}
-\begin{tabular}{lccccccc}
-  \hline
-&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$&$STG(3)^{*}=0.23$&$STG(3)^{*}=0.51$\\ 
-  \hline
-Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0 (0.009) & 0.006 (0.011) \\ 
- &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & -0.001 (0.009) & 0.025 (0.031) \\ 
-Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058) & -0.024 (0.065) & 0.013 (0.048) & 0.063 (0.139) & 0.023 (0.037) \\ 
- &0.5& 0.56 (0.048) & 0.028 (0.07) & -0.006 (0.075) & 0.011 (0.047) & 0.062 (0.128) & 0.013 (0.072) \\ 
-Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002) & 0.001 (0.003) & 0 (0.001) & -0.004 (0.01) & -0.004 (0.006) \\ 
- &0.5& 0.497 (0.002) & -0.001 (0.003) & 0 (0.003) & -0.001 (0.001) & -0.004 (0.009) & -0.012 (0.018) \\ 
-Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & 0.014 (0.064) & 0.047 (0.079) \\
-  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.025 (0.077) & 0.043 (0.158) \\ 
-Bias $\widetilde{STG}$ & 0.8 & 0.063 (0.050)& 0.013 (0.077)& 0.019 (0.101)& 0.050 (0.028) & 0.023 (0.093) & 0.032 (0.054) \\
- & 0.5 & 0.092 (0.073) & 0.041 (0.102) & 0.034 (0.053)& 0.055 (0.028)& 0.022 (0.095) & 0.069 (0.092) \\
-Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
-  &0.5& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
-   \hline
-\multicolumn{8}{p{9in}}{Bias entries for the $TPF(3|c)$ and $FPF(3|c)$ curves are average bias over the simulations and 2000 points on each simulated curve (empirical standard deviation of the estimated statistic over the simulations).
-H0$_1$ is the null hypothesis of no surrogate value (i.e., no variation in risk difference over $S_j(1)$ for any time point), based on the joint Wald test of $\gamma_{0j}=\gamma_{1j}=0$ $\forall j\in\{1\ldots,J\}$.
-*Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,x)$.}
-\end{tabular}
-\end{center}
-\end{sidewaystable}
-
 \begin{sidewaystable}[ht]
@@
   \hline
- & $R_{s_{1,j}w}$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$& $STG(3)^{*}=0.23$  & $STG(3)^{*}=0.51$\\
+ & $R_{s_{1,j}w}$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.20$& $STG(3)^{*}=0.28$  & $STG(3)^{*}=0.31$\\
   \hline
-$STG(3)=0.1$ &0.5 & 4.80 & 15.80 & 80.40 & 16.90 & 58.20 \\ 
-   &0.8 &7.20 & 31.20 & 90.00 & 32.80 & 96.40 \\ 
-$STG(3)=0.24$ &0.5&  & 4.40 & 41.80 & 8.00 & 39.00 \\ 
-   &0.8 &  & 5.20 & 49.60 & 10.00 & 69.60 \\ 
-$STG(3)=0.48$&0.5 &  &  & 6.00 & 44.30 & 13.60 \\ 
-   &0.8&  &  & 5.20 & 47.60 & 17.80 \\ 
-$STG(3)^{*}=0.23$&(0.5, 0.5) &  &  &  & 3.80 & 21.60 \\ 
-   &(0.8, 0.8) &  &  &  & 5.00 & 13.80 \\ 
-$STG(3)^{*}=0.51$&(0.5, 0.5)&   &  &  &  & 2.40 \\ 
-   &(0.8, 0.8)&  &  &  &  & 6.00 \\ 
+$STG(3)=0.1$ &0.8 & 5.5 & 17.0 & 16.8 & 28.2 & 71.0 \\ 
+   &0.5 &5.8 & 11.9 & 14.2 & 45.6 & 73.7 \\ 
+$STG(3)=0.24$ &0.8&  & 5.1 & 7.9 & 7.5 & 13.7 \\ 
+   &0.5 &  & 5.5 & 8.9 & 7.4 & 13.3 \\ 
+$STG(3)=0.20$&0.8 &  &  & 5.3 & 52.5 & 96.4 \\ 
+   &0.5&  &  & 6.3 & 45.5 & 93.9 \\ 
+$STG(3)^{*}=0.28$&(0.8, 0.8) &  &  &  & 5.4 & 7.7 \\ 
+   &(0.5, 0.5) &  &  &  & 5.7 & 8.9 \\ 
+$STG(3)^{*}=0.31$&(0.8, 0.8)&   &  &  &  & 5.1 \\ 
+   &(0.5, 0.5)&  &  &  &  & 4.2 \\ 
    \hline
@@
 \end{sidewaystable}
-
-\begin{sidewaystable}[ht]
-\begin{center}
-\caption{Power of difference in $\widetilde{STG}$ tests for $t \in (\tau, \infty)$ years \label{IPow}}
-\begin{tabular}{rrccccc}
-  \hline
- & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}=0.47$& $\widetilde{STG}^*=0.23$  & $\widetilde{STG}^*=0.52$\\
-  \hline
-$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 99.6 & 7.8 & 99.2 \\ 
-&  0.5  & 5.7 & 17.8 & 99.8 & 9.2 & 99.4 \\ 
- $\widetilde{STG}=0.25$ & 0.8 &  & 5.4 & 91.6 & 6.6 & 83.6 \\ 
-&  0.5 &  & 5.5 & 95.2 & 7.1 & 89.8 \\ 
-$\widetilde{STG}=0.47$ & 0.8 &  &  & 4.9 & 82.6 & 8.8 \\ 
- & 0.5 &  &  & 5.2 & 81.9 & 16.2 \\ 
-$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 75.8 \\ 
-&  (0.5, 0.5) &  &  &  & 6.3 & 78.0 \\ 
-$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 5.6 \\ 
- & (0.5, 0.5) &  &  &  &  & 5.9 \\ 
-   \hline
-\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $\widetilde{STG}_1-\widetilde{STG}_2=0$.
-*Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
-\end{tabular}
-\end{center}
-\end{sidewaystable}
-
-\section{Example}
-The Diabetes Control and Complication Trial (DCCT) enrolled 1,441 persons with type 1 diabetes from 1983 to 1989 to determine the effects of intensive diabetes therapy on long-term complications of diabetes.
-Participants in the DCCT were randomly assigned to intensive diabetes therapy aimed at lowering glucose concentrations as close as safely possible to the normal range or to conventional therapy aimed at preventing hyperglycemic symptoms.
-One of the outcomes of the DCCT, nephropathy (damage to the kidneys), is the leading cause of death and dialysis in the young with type 1 diabetes, particularly those with poorly controlled glucose levels.
-Nephropathy is often defined by a high albumin excretion rate, as micro-albuminuria (defined as albumin excretion rate $>$ 30 mg/24 hr) is the least invasive gold standard for kidney damage.
-The trial ended early in 1993 due to strong evidence of treatment efficacy, with a longest follow-up time of 7.5 years after the measurement of the biomarkers of interest; estimated adjusted mean risk of micro-albuminuria was reduced by 56\% (P-value 0.01)\citep{DCCT}.
-
-Although the trial could not be blinded to subjects or clinicians, it was blinded to investigators until the close of the trial.
-It is unlikely that the outcome would have been materially affected by subject or clinician knowledge of treatment status, as the only known risk factor in this population is poor diabetes management.
-Subjects that had dramatic behavioral changes while on trial were considered non-compliant and were censored at that point.
-This accounts for less than 3\% of person years in the trial and is balanced between treatment arms.
-The current analysis includes all participants who were free from micro-albuminuria at baseline (n = 1035); baseline micro-albuminuria is balanced over the arms of the trial.
-We use the time-point year 7.5 for our main analysis because it is the longest follow-up time in the trial.
-
-We consider four candidate surrogates: the difference in log-transformed hemoglobin A1C (HBA1C) measurements from baseline to year one, the difference in log-transformed estimated glomerular filtration rate (EGFR) measurements from baseline to year one, their linear combination into a univariate surrogate (HBA1C-EGFR) and the multivariate combination of the linear combination HBA1C-EGFR and HBA1C.
-Although there are some subjects for whom these values increased over the year, expert opinion is that increases were most likely due to measurement error or random fluctuation and are not indications of clinically relevant increases.
-We therefore truncate both candidate surrogates at zero for our primary analysis.
-EGFR values increased in only 1.2\% of subjects, 6.1\% for HBA1C and both increased in only one subject.
-This is consistent with the error for both being random.
-As a sensitivity analysis, we repeat the analysis with the non-truncated measurements and found the the results were similar (not reported).
-
-The event of interest is the onset of persistent micro-albuminuria, which is defined as having two consecutive albumin excretion rate measurements $>$ 30 mg/g.
-No subjects had an event prior to 1 year post-randomization.
-All subjects had the BIPs measured, which is defined as a linear combination of baseline laboratory measurements and demographic variables.
-The estimated linear correlation between the joint linear combination BIP and the candidate PS are 0.57 and 0.76 for HBA1C and univariate linear combination HBA1C-EGFR, respectively.
-All three univariate candidate PS also have respective individual BIPs.
-The correlation between the individual BIP and their PS are 0.7, for HBA1C, 0.88 for EGFR and 0.76 for the univariate linear combination HBA1C-EGFR.
-The individual BIPs are used for investigation of each biomarkers' univariate PS value, as well as comparison between the three univariate PS candidates.
-The joint linear combination BIP is used for the combined model.
-For all analyses we fit the conditional Weibull model proposed in A5 via the semi-parametric EML method and perform 1000 bootstraps.
-
-The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001.
-The P-value for testing H0$_1$ for EGFR is 0.19.
-The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.055.
-The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of 0.018.
-Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone.
-The estimated $\widehat{\widetilde{STG}}$ is 0.31 for HBA1C alone and 0.21 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.35.
-Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.
-
-The P-value is 0.058 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR.
-This is evidence that HBA1C alone is superior to the univariate the linear combination of HBA1C-EGFR.
-The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS gives a p-value of 0.18.
-
-Figure \ref{exp} Panel A depicts the $\widehat{CDF}^{7.5}(c)$ versus $c$ curves for all four candidate surrogates, HBA1C, EGFR the univariate linear combination HBA1C-EGFR and the multivariate combination.
-The EGFR curve is most similar to the marginal risk difference line, supporting the finding that EGFR has little value as a PS.
-The multivariate combination surrogate which is flatter than the EGFR alone curve is very similar to the HBA1C curve.
-This provides further evidence that the combination candidate does not improve upon HBA1C alone.
-As there is no evidence of violation of A6, Figure \ref{exp} Panel B displays the $\widehat{ROC}^{7.5}(q)$ versus $q$ curves for the HBA1C and the two combination candidates.
-Figure \ref{exp} Panel C depicts the $STG(t)$ versus $t$ for $t$ greater than one year for HBA1C and the two combination candidates.
-Both Panels B and C support that there is no significant difference between HBA1C alone and the multivariate combination, but that HBA1C alone is significantly better than the univariate linear combination candidate.
-
-A naive evaluation of the association of the two univariate candidate surrogates and outcome separately in each arm of the trial, using a Cox proportional hazards model and the hazard ratio magnitude for ranking, gives the same ranking of candidate surrogates in the treatment arm as do our methods.
-The findings support our conclusion that EGFR has no association with outcome in the treatment arm.
-However, in the control arm EGFR is significantly associated with outcome and HBA1C is not.
-Although the naive analyses finds some evidence that HBA1C is a better candidate surrogate than EGFR under treatment, it will not always be the case that rank is preserved among the correlates of risk in the treatment arm alone.
-The surrogate evaluation literature has many examples of biomarkers that are highly correlated with outcome under treatment alone, but are not associated with treatment efficacy \citep{flemming96}.
-
-\begin{sidewaysfigure}[H]
-\begin{center}
-\includegraphics[width=.9\textwidth]{dcct-example-figure-2014-6-2.pdf}
-\end{center}
-\caption{Panel A depicts the estimated $CDF^{8.5}(c)$ versus $c$ for the truncated univariate candidate principal surrogates HBA1C and EGFR as well as for the univariate linear combination HBA1C-EGFR candidate and the multivariate candidate PS.
-Panel B depicts the estimated $ROC^{8.5}(q)$ versus $q$ for HBA1C and both of the combination candidate.
-Both panels suggest that HBA1C is better than EGFR alone or their univariate combination and that there is no significant difference between multivariate combination and HBA1C alone.
-Panel C depicts the $STG(t)$ versus $t$ for each of truncated candidate PS that showed some evidence of surrogate value, HBA1C and the univariate combination HBA1C-EGFR and the multivariate combination, for $t$ greater than one year and less than the trial close of nine years.
-\label{exp}}
-\end{sidewaysfigure}
-
-\section{Discussion}
-Principal surrogates are important endpoints for Phase I and II trials.
-Without reliable surrogate endpoints for these trials, treatment development is slowed, allowing poor treatments to be advanced to Phase III and abandoning potentially effective formulations prior to Phase III.
-Few PS evaluation methods allow for comparison of candidate PS or for the evaluation of combinations of biomarkers as candidate PS and the few existing methods for comparison or combination of biomarkers as candidate PS have not been explored for time-to-event data.
-Our proposed Weibull model extension of the \citet{Huang11} semi-parametric EML method as well as our proposed classification accuracy and summary measures accommodate evaluation of multivariate PS and comparison of candidate PS using time-to-event data.
-
-Although our proposed estimands and estimators for the classification accuracy curves depend on assumption A6 to be well-defined, we show that when there are small undetectable violations of A6, the estimates approximate the desirable modified estimands of no-benefit or unaffected ROC curves (Appendix A supplementary materials).
-Even when A6 is clearly violated, the suggested $CDF^{t}_{\Delta}(c)$ and $STG(t)$ versus $t$ curves allow for visual comparison of candidate surrogates and the $STG(t)$ or $\widetilde{STG}$ difference tests perform well for discriminating between two candidate PS that both have some value as a PS.
-While only the risk estimands themselves and the proposed $CDF^{t}(c)$ curve perform well for biomarkers with no surrogate value, our proposed coefficient-based test of any surrogate value has satisfactory power and correct Type 1 error rate.
-Using the proposed methods we were able to show that the DCCT candidate surrogate HBA1C has more value as a PS than EGFR, and that the combination of HBA1C and EGFR in the same risk model was not an improvement over HBA1C alone.
-
-Our proposed semi-parametric EML estimators seem to perform well when there is a predictive BIP with correlation with the candidate PS of at least 0.5.
-The performance decreases with the correlation of the BIP and $S_j(1)$, as has been illustrated for previous EML methods.
-This is a limitation of EML based methods, however, for the motivating example the BIP correlation exceeds $0.5$.
-The need for a joint BIP that is well correlated with all biomarkers included in a multivariate PS model is also a limitation of the methods.
-Methods allowing for the use of individual BIP for each biomarker included in a multivariate PS model would be a useful extension to our methods.
-Although it is not the focus of this work, our model allows for time-variation in the hazard, an extension to the existing time-to-event PS evaluation methods \citep{Qin07, Miao13}.
-This extension is discussed in greater detail in the supplementary materials where we consider a more complex Weibull model allowing for time-variation in the treatment effect and the surrogate quality following \citet{Gabriel13}.
-
-\bibliographystyle{statmed}
-%\bibliographystyle{unsrtnat}
-\bibliography{bib_dissertation}
-\end{document}

### 2014-07-17 Compare_053014_statsmed_revisions.tex
Example update, fixed typos

@@
 We propose the use of several curves for displaying PS quality which consider the clinical outcome within levels defined by the marginal causal effect predictiveness (CEP) function rather than the levels of the surrogates themselves.
-A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is zero and when the risks are not equal the CEP is not equal to zero.
+A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is equal to zero and when the risks are not equal the CEP is not equal to zero.
 A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement.
@@
 Let $T_i(z)$ be the potential time from randomization to clinical event for individual $i$ had s/he received treatment $z\in \{0,1\}$.
-Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=min(T(z),C(z))$.
-Let there be $J$ candidate surrogates of interest which is measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z\in \{0,1\}$.
+Let $Y(z)$ be the indicator of $T(z)<C(z)$, where $C(z)$ is the potential censoring time and let $X(z)=\min\{T(z),C(z)\}$.
+Let there be $J$ candidate surrogates of interest which are measured prior to the clinical event but after randomization; with $S_{j}(z)$ being the $j$th candidate surrogate under treatment arm $z\in \{0,1\}$.
 Let $Y_{i}^{t}(z)$ be the indicator that $T_i(z)\leq t$.
@@
 Let $F_{S_1(1), \ldots, S_J(1)}(\cdot,\dots,\cdot)$ and $F_{S_1(1), \ldots, S_J(1)|W}(\cdot|\cdot)$ be the joint CDF of the $S_1(1), \ldots, S_J(1)$ and joint CDF of the $S_1(1), \ldots, S_J(1)$ conditional on $W$, respectively.
-Define $\rho_z(t)=Pr(Y^t(z)=1)=E(Y^t(z))=Pr(T(z)\leq t)$ to be the marginal prevalence of the potential clinical outcome at or before time $t$ for treatment arm $z$.
+Define $\rho_z(t)=Pr\{Y^t(z)=1\}=E\{Y^t(z)\}=Pr\{T(z)\leq t\}$ to be the marginal prevalence of the potential clinical outcome at or before time $t$ for treatment arm $z$.
 Our risk estimands of interest, $risk_z(s)$, are functions of the candidate surrogate that give some measure of risk under the treatment $z$.
@@
 Clearly, we do not want a large amount of clinical treatment effect for a group of subjects with no treatment effect on any of the surrogates.
-However, ACN may not guarantee this is the case \citep{Vanderweele11}.
+However, ACN may not guarantee this is that case \citep{Vanderweele11}.
 We propose to extend the concept of ACN to investigate the ability of the model for risk difference conditional on a set of biomarkers to correctly categorize subjects by their treatment effect at time $t$, $D(t)$.
@@
 Assumption A2, is generally true in randomized, blinded trials.
-Assumption A3 aides in the identification of subjects with $T < \tau$ to be excluded from the evaluation cohort without causing bias.
-Although, A3 is not fully testable, indications of A3 violation are significantly unbalanced event rates over the trial arms prior to $\tau$.
+Assumption A3 aids in the identification of subjects with $T < \tau$ to be excluded from the evaluation cohort without causing bias.
+Although A3 is not fully testable, indications of A3 violation are significantly unbalanced event rates over the trial arms prior to $\tau$.
 Assumption A4 is standard in much of the survival analysis literature and allows for classic censoring models to be used.
@@
 Imputed values for each missing biomarker are produced separately, but provided that the same baseline variables are used in each model, empirical integration over the set of all imputations is approximately equivalent to integrating over the joint distribution of biomarkers given the baseline variables used for imputation.
-This is an important limitation of this method.
 The BIP need not be based on a single baseline measurement or have the same correlation with each candidate, but the BIP must be the same in each location-scale model for the estimated likelihood to be proper using our procedure.
@@
 \begin{eqnarray*}
-&&\mathit{l}(\beta,\gamma, \hat{\nu})=\sum_{i} log(g_z(T_i|\gamma, \beta,Z_i, s_{1,1,i}, \ldots,s_{1,j,i},Y_i))*\left\{\prod_{j=1}^{J}{M
+&&\mathit{l}(\beta,\gamma, \hat{\nu})=\sum_{i} \log(g_z(T_i|\gamma, \beta,Z_i, s_{1,1,i}, \ldots,s_{1,j,i},Y_i))*\left\{\prod_{j=1}^{J}{M
 _{j,i}}\right\}+\\
-&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, S^{*}_{l+1,i,k}(1),\ldots, S^{*}_{J,i,k}(1), Y_i))}\right\}*\\
+&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{\log(g(T_i|\gamma, \beta, Z_i,s_{1,1,i},\ldots, s_{1,l,i}, S^{*}_{l+1,i,k}(1),\ldots, S^{*}_{J,i,k}(1), Y_i))}\right\}*\\
 &&\left\{\prod_{j=1}^{l}{M_{j,i}}*\prod_{j=l+1}^{J}{(1-M_{j,i})}\right\}+\\
-&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{log(g_z(T_i|\gamma, \beta,Z_i, S_{1,i,k}^*(1), \ldots,S_{j,i,k}^*(1),Y_i))}\right\}*\left\{\prod_{j=1}^{J}{(1-M
+&&\sum_{i}\left\{\left({1 \over n_{V}}\right)\sum_k^{n_V}{\log(g_z(T_i|\gamma, \beta,Z_i, S_{1,i,k}^*(1), \ldots,S_{j,i,k}^*(1),Y_i))}\right\}*\left\{\prod_{j=1}^{J}{(1-M
 _{j,i})}\right\}\cdot
@@
 The $STG(t)$ is of interest for comparison of candidate surrogates regardless of the validity of assumption A6 as steeper quantile curves, and thus greater area between $R^t(v)$ and $(\rho_0(t) -\rho_1(t))$, correspond to better prediction of clinical treatment effect and a better candidate surrogate.
-Under A6, $STG(t)$ also has a useful interpretation based on the classification accuracy measures, given by $STG(t)=max_c\{\mbox{Sensitivity}(t|c) + \mbox{Specificity}(t|c)\} -1$.
+Under A6, $STG(t)$ also has a useful interpretation based on the classification accuracy measures, given by $STG(t)=\max_c\{\mbox{Sensitivity}(t|c) + \mbox{Specificity}(t|c)\} -1$.
 This interpretation follows directly from the proof given in \citet{Huang11} for the binary clinical endpoint setting.
@@
 As an alternative to estimating the $STG$ at a fixed time $t$, we can summarize the $STG(t)$ by averaging over the estimated distribution of event times.
-We define the integrated $STG$ ($ISTG$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T.
+We define the integrated $STG$ ($\widetilde{STG}$) as $\int STG(t) \, dF(t)$, where $F(t)$ is the marginal CDF of T.
 The estimated marginal distribution of event times $\hat{F}(T)$ can be reliably calculated by the Kaplan-Meier estimator.
@@
 This may not always be the case.
-When assumption A6 holds the candidate surrogate with the highest left-most point of the $ROC^{t}(q)$ curve should also be the candidate surrogate with the highest $STG(t)$.
+When assumption A6 holds the candidate surrogate with the highest left-most point of the $ROC^{t}(q)$ curve should also be the candidate surrogate with the highest $\widetilde{STG}$.
 When A6 is violated this may not be true.
@@
 However, as with almost all statistical tests, a failure to reject the null does not prove the null; lack of significant  $\widetilde{STG}$ difference does not prove lack of superiority.
+Multivariate candidates will tend to have higher $\widetilde{STG}$ than the univariate components simply because of the greater variation over multiple surrogates.
+Due to the increased burden of collecting multiple biomarker measurements, we suggest that when ranking candidate PS, univariate candidates are given greater priority unless a multivariate PS is found to be statistically significantly superior.
 
 \section{Simulation Studies}
-Suppose the conditional CDF of $T$ given $S_j(1)$ and $Z$ follows a Weibull model and $\{S_j(1),W\}$ follows a multivariate truncated normal model with correlation R$_{S_jW}$ between each of the univariate candidates and the BIP and correlation R$_{S_kS_j}$ between each pair of candidates.
+Suppose the conditional CDF of $T$ given $S_j(1)$ and $Z$ follows a Weibull model and $\{S_j(1),W\}$ follows a multivariate normal or censored normal model with correlation R$_{S_jW}$ between each of the univariate biomarkers and the potentially multivariate BIP.
 Information lost to drop out occurs completely at random and at a rate of approximately 5\% per year.
@@
 
-We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with a $STG$ at year three, $STG(3)$, of 0.48, two marginal quality surrogates with $STG(3)$ of 0.1 and 0.24, and a null surrogate with $STG(t)=0$ for all time points $t$.
-We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $STG(3)$ of 0.23 and 0.51, respectively.
-We investigate these different levels of candidate surrogates for two different BIP correlations, $R_{s_{1j}w}$, of 0.8 and 0.5.
-For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0.8), (0.5, 0.5)\}$ with a correlation between the candidate surrogates ranging from 0 to 0.3.
-Results for the 0.3 correlation are displayed.
+We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with an integrated $STG$, $\widetilde{STG}$, of 0.37, two marginal quality surrogates with $\widetilde{STG}$ of 0.1 and 0.25, and a null surrogate with $\widetilde{STG} = 0$.
+We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $\widetilde{STG}$ of 0.46 and 0.81, respectively.
+We investigate these different levels of candidate surrogates for two different BIP correlations; $R_{s_{1j}w}$, of 0.8 and 0.5 between the univariate candidate and univariate BIP.
+For the multivariate scenarios, we consider the correlation value sets ranging from $(R_{s_{1,1}w},R_{s_{1,2}w})\in\{(0.8,0,0,0.8), (0.5,0,0,0.5)\}$ between the bivariate candidate PS and the bivariate BIP $w=(w_1, w_1)$.
+We investigate correlations between the biomarkers in the bivariate PS ranging from 0 to 0.3 and the correlation between the Ws in the vector BIP from 0 to 1.
+Results for $R_{S_kS_j}=0.3$ and two independent Ws for a bivariate BIP are displayed in Tables \ref{Bias} and \ref{IPow}; other scenarios can be found in Appendix B of the supplementary materials.
 500 replicates of the simulation were run for each scenario.
-All candidate PSs in the simulations are time invariant and therefore the characteristics of the $STG(t)$ estimator can be determined based on a single time-point $t$, for this we use the simulated trial terminal follow-up time of  3 years; results for time-dependent models can be found in Table 4, Appendix B of the supplementary materials.
+All candidate PSs in the simulations are time invariant; results for time-dependent models can be found in Table 6, Appendix B of the supplementary materials.
 
-Bias results are displayed in Table \ref{Bias} and suggest that the semi-parametric EML method for the Weibull risk difference model is adequately unbiased and that the suggested estimators for $CDF^{t}_{\Delta}(c)$, $TPF(t|c)$, $FPF(t|c)$, $STG(t)$ and $\widetilde{STG}$ have adequate precision and accuracy in all cases for which they are applicable.
+We fit the Weibull risk in each scenario using the proposed semi-parametric EML method, censoring the imputations when the data generation was censored and using the imputations as generated if normal generation was used; scenarios indicated by a $\#$ in Tables \ref{Bias} and \ref{IPow} use censored generation and imputation.
+Bias results are displayed in Table \ref{Bias} and suggest that the method is adequately unbiased and that the suggested estimators for $CDF^{t}_{\Delta}(c)$, $TPF(t|c)$, $FPF(t|c)$, $STG(t)$ and $\widetilde{STG}$ have adequate precision and accuracy in all cases for which they are applicable.
 As we have suggested above, there is some bias when there is no surrogate value.
 When there is no surrogate value the estimated model coefficients are still unbiased, with average bias less than 3\% of one Monte Carlo SD (results not displayed).
-The large standard deviation of the $STG(t)$ estimates over all the simulations may imply that this estimator is not as efficient as possible.
 A correlation of 0.5 or higher between a candidate PS and a BIP seems adequate for unbiased estimation of the suggested summary statistics and curves.
+Although unbiased, the results from using a univariate BIP for a multivariate candidate PS indicate that convergence issues can occur in some scenarios and that vector BIPs improve performance; results can be seen in Table 1 of Appendix B of the supplementary materials.
+
 The results for the joint Wald test of any surrogate value, displayed in Table \ref{Bias}, suggest that this test has correct Type 1 error and adequate power in all scenarios.
 Monte Carlo standard errors are used in all Wald tests due to the computational burden of the bootstrap.
-
-Table \ref{Power} displays the power to discriminate between two surrogate quality levels based on the $STG(t)$ difference Wald test for fixed time point $t=3$.
+Table \ref{IPow} displays the power to discriminate between two surrogate quality levels based on the difference in integrated STG, $\widetilde{STG}$.
 Monte Carlo standard errors are used and only those candidate surrogates with any surrogate quality are displayed.
 We find that for all null hypothesis scenarios the test has correct type 1 error.
-We also find that the test has adequate power to detect differences in $STG(3)$, particular when there is good correlation between the BIP and the candidate PS, with the power increasing with the difference in $STG(3)$, as expected.
+We also find that the test has adequate power to detect differences in $\widetilde{STG}$, particular when there is good correlation between the BIP and the candidate PS, with the power increasing with the difference in $\widetilde{STG}$, as expected.
 The test is not valid if either of the candidate surrogates has no surrogate value, as type 1 error is inflated when the true value is on the boundary of the parameter space (results not shown).
-
-Table \ref{IPow} displays the power to discriminate between two surrogate quality levels based on the difference in integrated STG, $\widetilde{STG}$.
-We again find that for all null hypothesis scenarios the test has nearly a nominal 5\% Type 1 error rate with similar to slightly higher power than the $STG(3)$ difference test for the same scenarios.
+The power to discriminate based on the difference in STG at a fixed time point shows similar patterns; results can be seen in Table 2 of Appendix B of the supplementary materials.
 
@@
   \hline
-&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$&$STG(3)^{*}=0.23$&$STG(3)^{*}=0.51$\\ 
+&$R_{s_{1,j}w}^{**}$&$STG(3)=0$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG^{\#}(3)=0.48$&$STG(3)^{\#*}=0.23$&$STG(3)^{\#*}=0.51$\\ 
   \hline
@@
 H0$_1$ is the null hypothesis of no surrogate value (i.e., no variation in risk difference over $S_j(1)$ for any time point), based on the joint Wald test of $\gamma_{0j}=\gamma_{1j}=0$ $\forall j\in\{1\ldots,J\}$.
-*Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,x)$.}
+*Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,0,0,x)$ for bivariate candidate PS and the bivariate BIP.
+\# indicates that the candidate surrogate was generated using a censored normal distribution.}
 \end{tabular}
@@
 \begin{center}
-\caption{Power of difference in $STG(t)$ tests for $t=3$ years \label{Power}}
+\caption{Power to detect difference in $\widetilde{STG}$ \label{IPow}}
 \begin{tabular}{rrccccc}
   \hline
- & $R_{s_{1,j}w}$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)=0.48$& $STG(3)^{*}=0.23$  & $STG(3)^{*}=0.51$\\
-  \hline
-$STG(3)=0.1$ &0.5 & 4.80 & 15.80 & 80.40 & 16.90 & 58.20 \\ 
-   &0.8 &7.20 & 31.20 & 90.00 & 32.80 & 96.40 \\ 
-$STG(3)=0.24$ &0.5&  & 4.40 & 41.80 & 8.00 & 39.00 \\ 
-   &0.8 &  & 5.20 & 49.60 & 10.00 & 69.60 \\ 
-$STG(3)=0.48$&0.5 &  &  & 6.00 & 44.30 & 13.60 \\ 
-   &0.8&  &  & 5.20 & 47.60 & 17.80 \\ 
-$STG(3)^{*}=0.23$&(0.5, 0.5) &  &  &  & 3.80 & 21.60 \\ 
-   &(0.8, 0.8) &  &  &  & 5.00 & 13.80 \\ 
-$STG(3)^{*}=0.51$&(0.5, 0.5)&   &  &  &  & 2.40 \\ 
-   &(0.8, 0.8)&  &  &  &  & 6.00 \\ 
-   \hline
-\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $STG_1(3)-STG_2(3)=0$.
-*Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
-\end{tabular}
-\end{center}
-\end{sidewaystable}
-
-\begin{sidewaystable}[ht]
-\begin{center}
-\caption{Power of difference in $\widetilde{STG}$ tests for $t \in (\tau, \infty)$ years \label{IPow}}
-\begin{tabular}{rrccccc}
-  \hline
- & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}=0.47$& $\widetilde{STG}^*=0.23$  & $\widetilde{STG}^*=0.52$\\
+ & $R_{s_{1,j}w}$ & $\widetilde{STG}=0.10$  & $\widetilde{STG}=0.25$  & $\widetilde{STG}^{\#}=0.47$& $\widetilde{STG}^{\#*}=0.23$  & $\widetilde{STG}^{\#*}=0.52$\\
   \hline
@@
 \multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $\widetilde{STG}_1-\widetilde{STG}_2=0$.
-*Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.}
+*Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$.
+\# indicates that the candidate surrogate was generated using a censored normal distribution.}
 \end{tabular}
@@
 Nephropathy is often defined by a high albumin excretion rate, as micro-albuminuria (defined as albumin excretion rate $>$ 30 mg/24 hr) is the least invasive gold standard for kidney damage.
-The trial ended early in 1993 due to strong evidence of treatment efficacy, with a longest follow-up time of 7.5 years after the measurement of the biomarkers of interest; estimated adjusted mean risk of micro-albuminuria was reduced by 56\% (P-value 0.01)\citep{DCCT}.
+The trial ended early in 1993 due to strong evidence of treatment efficacy, with a longest follow-up time of 7.5 years after the measurement of the biomarkers of interest; estimated adjusted mean risk of micro-albuminuria was reduced by 56\% (P-value $=0.01$)\citep{DCCT}.
 
@@
 Although there are some subjects for whom these values increased over the year, expert opinion is that increases were most likely due to measurement error or random fluctuation and are not indications of clinically relevant increases.
-We therefore truncate both candidate surrogates at zero for our primary analysis.
+We therefore censored both candidate surrogates at zero for our primary analysis.
 EGFR values increased in only 1.2\% of subjects, 6.1\% for HBA1C and both increased in only one subject.
 This is consistent with the error for both being random.
-As a sensitivity analysis, we repeat the analysis with the non-truncated measurements and found the the results were similar (not reported).
+As a sensitivity analysis, we repeat the analysis with the non-censored measurements and found that the results were similar (not reported).
 
@@
 All subjects had the BIPs measured, which is defined as a linear combination of baseline laboratory measurements and demographic variables.
-The estimated linear correlation between the joint linear combination BIP and the candidate PS are 0.57 and 0.76 for HBA1C and univariate linear combination HBA1C-EGFR, respectively.
+The estimated linear correlation between the vector BIP and the candidate PS are 0.69 and 0.82 for HBA1C and univariate linear combination HBA1C-EGFR, respectively.
 All three univariate candidate PS also have respective individual BIPs.
@@
 The individual BIPs are used for investigation of each biomarkers' univariate PS value, as well as comparison between the three univariate PS candidates.
-The joint linear combination BIP is used for the combined model.
+The vector BIP is used for the combined model.
 For all analyses we fit the conditional Weibull model proposed in A5 via the semi-parametric EML method and perform 1000 bootstraps.
@@
 The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001.
-The P-value for testing H0$_1$ for EGFR is 0.19.
-The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.055.
-The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of 0.018.
+The P-value for testing H0$_1$ for EGFR is 0.18.
+The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.062.
+The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of $<$ 0.001.
 Based on the tests of H0$_1$ alone there is some evidence that HBA1C and both the combination candidate surrogates are better than EGFR alone.
-The estimated $\widehat{\widetilde{STG}}$ is 0.31 for HBA1C alone and 0.21 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.35.
+The estimated $\widehat{\widetilde{STG}}$ is 0.37 for HBA1C alone and 0.12 for the univariate linear combination HBA1C-EGFR; their multivariate combination has a $\widehat{\widetilde{STG}}$ of 0.46.
 Given that there is no evidence to suggest that EGFR alone has any value as a PS, we do not pursue estimation or testing via any of the suggested summary statistics.
 
-The P-value is 0.058 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR.
-This is evidence that HBA1C alone is superior to the univariate the linear combination of HBA1C-EGFR.
-The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS gives a p-value of 0.18.
+The P-value is 0.018 for testing that the estimated $\widetilde{STG}$ difference is zero between HBA1C alone and the univariate linear combination of HBA1C and EGFR.
+This is marginal evidence that HBA1C alone is superior to the univariate linear combination of HBA1C-EGFR.
+The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS provides no evidence of a significant difference, P-value of 0.2.
 
@@
 \begin{center}
-\includegraphics[width=.9\textwidth]{dcct-example-figure-2014-6-2.pdf}
+\includegraphics[width=.9\textwidth]{dcct-example-figure-2014-7-16.pdf}
 \end{center}
@@
 This is a limitation of EML based methods, however, for the motivating example the BIP correlation exceeds $0.5$.
-The need for a joint BIP that is well correlated with all biomarkers included in a multivariate PS model is also a limitation of the methods.
-Methods allowing for the use of individual BIP for each biomarker included in a multivariate PS model would be a useful extension to our methods.
+We also find that for higher dimensional candidate surrogates vector BIPs can improve convergence properties over univariate BIPs.
 Although it is not the focus of this work, our model allows for time-variation in the hazard, an extension to the existing time-to-event PS evaluation methods \citep{Qin07, Miao13}.

### 2014-07-17 Compare_053014_statsmed_revisions.tex
Final edits, still waiting on tables

@@
 \address{\affilnum{a}Vaccine and Infectious Disease Division, Fred Hutchinson Cancer Research Center\\
-\affilnum{b}Kidney Research Institute, University of Washington\\
+\affilnum{b}Biometric Research Branch, National Cancer Institute\\
 \affilnum{c}Department of Biostatistics, University of Washington}
@@
 A CEP is a function of the treatment-arm-specific biomarker-conditional risks, such that when the risks are equal the CEP is equal to zero and when the risks are not equal the CEP is not equal to zero.
-A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement.
+A marginal-CEP is simply a contrast of the clinical outcome over the trial arms within subgroups defined by the candidate surrogate under active treatment; for example, the additive difference in risk between the placebo and treatment groups in those subjects that have or would have had a biomarker measurement of $x$ under treatment.
 Univariate candidate PS can be evaluated by the strength of the association between the candidate PS under active treatment assignment and the clinical treatment effect.
@@
 We develop summary statistics for the comparison of candidate PS for a continuous time-to-event endpoint.
-Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS .
+Most existing methods of evaluating candidate PS are not helpful for ranking candidate PS.
 Often, evaluation criteria simply provide a way of testing the binary quality of being or not being a PS, rather than providing a quantitative measurement of surrogate quality.
@@
 In Section 5, we apply our methods to the Diabetes Control and Complications Trial (DCCT).
-We find evidence to support the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that combinations of the two candidates are not improvements over the higher quality surrogate alone.
+We find evidence to support that the high quality surrogate identified in \citet{Gabriel13} is better than an alternative candidate surrogate and that the combination of the two candidates is not an improvement over the higher quality surrogate alone.
 In Section 6, we discuss potential limitations of our methods and future avenues for research.
@@
 In \citet{Gilbert08} and most of the recent literature, marginal risk estimands are used instead of joint risk estimands because they are easier to identify and estimate.
-marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$.
+Marginal risk estimands condition on $S(1)$ and marginalize over the distribution of $S(0)$.
 The concept of wide variability is easily extended to marginal risk, such that contrasts of risk over the arms of the trial, such as our CEP of risk difference, vary widely in $s_1$.
@@
 Our suggested accuracy measures and the CDF of the risk model CEP can help more finely group candidates by predictive quality and amount of variation in CEP, but they do not provide inference for comparing candidates.
-We propose the use of the standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for comparison inference.
+We propose the use of the standardized total gain for candidate surrogate ranking and tests of the difference in standardized total gain between two candidates for inference.
 These are outlined in Sections \ref{STG} and \ref{ISTG}.
@@
 Given assumptions A1-A4 we can link the time-dependent potential outcome risk estimands of interest to observed data clinical risk estimands and the partially observed candidate surrogates by assuming a parametric survival model.
-We will assume a Weibull model, the probability density function (pdf) for which we will denote as $g(\cdot|\cdot,\cdot)$.
+We will assume a Weibull model, the probability density function (pdf) which we will denote as $g(\cdot|\cdot,\cdot)$.
 We parameterize the scale term of this model, $\gamma=(\gamma_{00}, \gamma_{10},\gamma_{0j}, \gamma_{1j})$, and allow for a constant shape term, $\beta_{0}$; this parameterization is equivalent to a classical Weibull survival model.
@@
 Similarly, $\mbox{Specificity}(t|c)$ or $1-FPF(t|c)$, is of interest as those subjects not helped by treatment should have a high probability of having a small risk difference based on a predictive model.
-As these measures are threshold dependent, they can be used to compare not just candidate surrogates but thresholds for a given surrogate.
+As these measures are threshold dependent, they can be used to compare not just candidate surrogates but thresholds for a given risk model.
 Thresholds $c$ found to have high $TPF(t|c)$ and low $FPF(t|c)$ are ideal.
@@
 It is of interest to determine the relationship between $TPF(t|c)$ and $FPF(t|c)$.
-The time-dependent ROC curve gives an illustration of the $TPF(t|c)$ and $FPF(t|c)$ relationship at given levels of $c$.
+The time-dependent ROC curve gives an illustration of the $TPF(t|c)$ and $FPF(t|c)$ relationship at given levels of the risk difference.
 This curve is defined by $ROC^{t}(q)=TPF\{t|FPF^{-1}(t|q)\}$ for a given time point $t$, adapting the binary-clinical endpoint $ROC(q)$ definition of \citet{Huang12b}.
@@
 We suggest the plug-in estimator of $\widehat{CDF}^{t}_{\Delta}(c)$, $\widehat{CDF}^{t}_{\Delta}(c)=\int{\mathrm{I}_{[\hat{\Delta}(t|s_{1,1,i}, \dots, s_{1,j,i})\leq c]} d\hat{F}_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})}$.
-Although we display a useless surrogate in Figure \ref{4panel}, we suggest testing for any surrogate value prior to estimation of $\widehat{CDF}^{t}_{\Delta}(c)$ as it was found in simulations that there is evidence of bias when there is no variation in risk difference; a suggested test is given in Section \ref{CRE}.
 
@@
 The test for difference between $\widehat{\widetilde{STG}}$ for HBA1C alone and the multivariate candidate PS provides no evidence of a significant difference, P-value of 0.2.
+Although the multivariate surrogate has a notably higher STG, given the lack of statistical evidence and the increased burden of measuring a second surrogate, we would still favor the use of HBA1C alone.
 
@@
 Figure \ref{exp} Panel C depicts the $STG(t)$ versus $t$ for $t$ greater than one year for HBA1C and the two combination candidates.
-Both Panels B and C support that there is no significant difference between HBA1C alone and the multivariate combination, but that HBA1C alone is significantly better than the univariate linear combination candidate.
+Both Panels A and C support that there is no significant difference between HBA1C alone and the multivariate combination, but that HBA1C alone is significantly better than the univariate linear combination candidate.
+Panel C indicates that there may be some value in the combination surrogate above the HBA1C alone, however due to the high variance, there is insufficient evidence that this difference is significant.
 
@@
 Even when A6 is clearly violated, the suggested $CDF^{t}_{\Delta}(c)$ and $STG(t)$ versus $t$ curves allow for visual comparison of candidate surrogates and the $STG(t)$ or $\widetilde{STG}$ difference tests perform well for discriminating between two candidate PS that both have some value as a PS.
-While only the risk estimands themselves and the proposed $CDF^{t}(c)$ curve perform well for biomarkers with no surrogate value, our proposed coefficient-based test of any surrogate value has satisfactory power and correct Type 1 error rate.
-Using the proposed methods we were able to show that the DCCT candidate surrogate HBA1C has more value as a PS than EGFR, and that the combination of HBA1C and EGFR in the same risk model was not an improvement over HBA1C alone.
+While only the risk estimands themselves perform well for biomarkers with no surrogate value, our proposed coefficient-based test for discrimination between potentially useful and useless surrogates has good power and correct Type 1 error rate.
+Using the proposed methods we were able to show that the DCCT candidate surrogate HBA1C has more value as a PS than EGFR, and that the combination candidates were not a significant improvement over HBA1C alone.
 
 Our proposed semi-parametric EML estimators seem to perform well when there is a predictive BIP with correlation with the candidate PS of at least 0.5.
-The performance decreases with the correlation of the BIP and $S_j(1)$, as has been illustrated for previous EML methods.
-This is a limitation of EML based methods, however, for the motivating example the BIP correlation exceeds $0.5$.
-We also find that for higher dimensional candidate surrogates vector BIPs can improve convergence properties over univariate BIPs.
+The performance decreases with the correlation of the BIP and $S_j(1)$, as has been illustrated for previous EML methods; this is a limitation of EML based methods.
+We find that for higher dimensional candidate surrogates vector BIPs can improve convergence properties over univariate BIPs for our method.
+Further investigation of the effect of using a vector BIP on the efficiency for this and other EML based methods is of research interest.
 Although it is not the focus of this work, our model allows for time-variation in the hazard, an extension to the existing time-to-event PS evaluation methods \citep{Qin07, Miao13}.
-This extension is discussed in greater detail in the supplementary materials where we consider a more complex Weibull model allowing for time-variation in the treatment effect and the surrogate quality following \citet{Gabriel13}.
+Further extension of these methods to allow even more flexible models may be useful for some surrogate evaluation settings.
 

### 2014-07-18 Compare_053014_statsmed_revisions.tex
Submission time

@@
 
-We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with an integrated $STG$, $\widetilde{STG}$, of 0.37, two marginal quality surrogates with $\widetilde{STG}$ of 0.1 and 0.25, and a null surrogate with $\widetilde{STG} = 0$.
-We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $\widetilde{STG}$ of 0.46 and 0.81, respectively.
+We investigate four different univariate candidate PSs with varying quality levels, a high quality surrogate with an integrated $STG$, $\widetilde{STG}$, of 0.47, two marginal quality surrogates with $\widetilde{STG}$ of 0.1 and 0.25, and a null surrogate with $\widetilde{STG} = 0$.
+We also investigate two scenarios where the true model for outcome contains a multivariate surrogate, where $S_1(1)$ and $S_2(1)$ combined under model A5 have a $\widetilde{STG}$ of 0.23 and 0.52, respectively.
 We investigate these different levels of candidate surrogates for two different BIP correlations; $R_{s_{1j}w}$, of 0.8 and 0.5 between the univariate candidate and univariate BIP.
@@
 When there is no surrogate value the estimated model coefficients are still unbiased, with average bias less than 3\% of one Monte Carlo SD (results not displayed).
+For two of the univariate scenarios the models are estimated under minor violations of assumption A6 (indicated by $\dagger$).
+In this case, the TPF and FPF correspond to the modified versions defined in section \ref{CAM}.
 A correlation of 0.5 or higher between a candidate PS and a BIP seems adequate for unbiased estimation of the suggested summary statistics and curves.
@@
   \hline
-Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0 (0.009) & 0.006 (0.011) \\ 
- &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & -0.001 (0.009) & 0.025 (0.031) \\ 
-Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058) & -0.024 (0.065) & 0.013 (0.048) & 0.063 (0.139) & 0.023 (0.037) \\ 
- &0.5& 0.56 (0.048) & 0.028 (0.07) & -0.006 (0.075) & 0.011 (0.047) & 0.062 (0.128) & 0.013 (0.072) \\ 
-Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002) & 0.001 (0.003) & 0 (0.001) & -0.004 (0.01) & -0.004 (0.006) \\ 
- &0.5& 0.497 (0.002) & -0.001 (0.003) & 0 (0.003) & -0.001 (0.001) & -0.004 (0.009) & -0.012 (0.018) \\ 
-Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & 0.014 (0.064) & 0.047 (0.079) \\
-  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.025 (0.077) & 0.043 (0.158) \\ 
-Bias $\widetilde{STG}$ & 0.8 & 0.063 (0.050)& 0.013 (0.077)& 0.019 (0.101)& 0.050 (0.028) & 0.023 (0.093) & 0.032 (0.054) \\
- & 0.5 & 0.092 (0.073) & 0.041 (0.102) & 0.034 (0.053)& 0.055 (0.028)& 0.022 (0.095) & 0.069 (0.092) \\
+Bias $CDF^{3}_{\Delta}(c)$& 0.8& 0.052 (0.009) & 0 (0.008) & 0.003 (0.009) & 0 (0.006) & 0.003 (0.029) & 0.002 (0.012) \\ 
+ &0.5& 0.052 (0.009) & 0.001 (0.008) & 0.002 (0.009) & 0.002 (0.006) & 0.001 (0.024) & 0.003 (0.014) \\ 
+Bias $TPF(3|c)$& 0.8& 0.544 (0.035) & 0.014 (0.058)$^\dagger$ & -0.024 (0.065)$^\dagger$ & 0.013 (0.048) & 0.005 (0.060) & 0.003 (0.019) \\ 
+ &0.5& 0.56 (0.048) & 0.028 (0.07)$^\dagger$ & -0.006 (0.075)$^\dagger$ & 0.011 (0.047) & 0.029 (0.062) & 0.009 (0.022) \\ 
+Bias $FPF(3|c)$& 0.8& 0.497 (0.002) & 0 (0.002)$^\dagger$ & 0.001 (0.003)$^\dagger$ & 0 (0.001) & 0.001 (0.005) & -0.006 (0.019) \\ 
+ &0.5& 0.497 (0.002) & -0.001 (0.003)$^\dagger$ & 0 (0.003)$^\dagger$ & -0.001 (0.001) & -0.001 (0.024) & -0.004 (0.018) \\ 
+Bias $STG(3)$& 0.8& 0.066 (0.053) & 0.021 (0.086) & -0.043 (0.101) & -0.004 (0.06) & -0.005 (0.094) & 0.013 (0.026) \\
+  &0.5 & 0.09 (0.072) & 0.042 (0.103) & -0.016 (0.115) & -0.015 (0.059) & 0.028 (0.088) & 0.046 (0.034) \\ 
+Bias $\widetilde{STG}$ & 0.8 & 0.063 (0.050)& 0.013 (0.077)& 0.019 (0.101)& 0.050 (0.028) & -0.004 (0.096) & 0.010 (0.031) \\
+ & 0.5 & 0.092 (0.073) & 0.041 (0.102) & 0.034 (0.053)& 0.055 (0.028)& 0.032 (0.091) & 0.021 (0.04) \\
 Power H0$_1$& 0.8& 0.05 & 0.94 & 0.99 & 0.99 & 0.99 & 0.99 \\ 
@@
 *Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,0,0,x)$ for bivariate candidate PS and the bivariate BIP.
-\# indicates that the candidate surrogate was generated using a censored normal distribution.}
+\# indicates that the candidate surrogate was generated using a censored normal distribution.$^\dagger$ indicates estimation under minor violation of assumption A6, where it is for the slightly modified interpretation of TPF and FPF (see section \ref{CAM})}
 \end{tabular}
@@
   \hline
-$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 14.4 & 99.6 & 7.8 & 99.2 \\ 
-&  0.5  & 5.7 & 17.8 & 99.8 & 9.2 & 99.4 \\ 
- $\widetilde{STG}=0.25$ & 0.8 &  & 5.4 & 91.6 & 6.6 & 83.6 \\ 
-&  0.5 &  & 5.5 & 95.2 & 7.1 & 89.8 \\ 
-$\widetilde{STG}=0.47$ & 0.8 &  &  & 4.9 & 82.6 & 8.8 \\ 
- & 0.5 &  &  & 5.2 & 81.9 & 16.2 \\ 
-$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 4.6 & 75.8 \\ 
-&  (0.5, 0.5) &  &  &  & 6.3 & 78.0 \\ 
-$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 5.6 \\ 
- & (0.5, 0.5) &  &  &  &  & 5.9 \\ 
+$\widetilde{STG}=0.10$ & 0.8 & 4.8 & 15.5 & 98.7 & 14.6 & 98.2 \\ 
+&  0.5  & 5.7 & 13.7 & 93.4 & 9.7 & 88.3 \\ 
+ $\widetilde{STG}=0.25$ & 0.8 &  & 5.4 & 74.2 & 5.3 & 69.5 \\ 
+&  0.5 &  & 5.5 & 64.5 & 5.6 & 49.1 \\ 
+$\widetilde{STG}=0.47$ & 0.8 &  &  & 4.9 & 86.1 & 14.1 \\ 
+ & 0.5 &  &  & 5.2 & 80.7 & 7.5 \\ 
+$\widetilde{STG}^{*}=0.23$ & (0.8, 0.8) &  &  &  & 5.1 & 79.1 \\ 
+&  (0.5, 0.5) &  &  &  & 5.8 & 78.2 \\ 
+$\widetilde{STG}^{*}=0.52$ & (0.8, 0.8) &  &  &  &  & 5.4 \\ 
+ & (0.5, 0.5) &  &  &  &  & 5.6 \\ 
    \hline
@@
 The analysis indicates that HBA1C has clear surrogate value, with P-value for testing H0$_1$ of 0.001.
-The P-value for testing H0$_1$ for EGFR is 0.18.
-The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.062.
+The P-value for testing H0$_1$ for EGFR is 0.508.
+The univariate linear combination HBA1C-EGFR has a P-value for testing H0$_1$ of 0.06.
 The multivariate candidate PS of HBA1C combined with the univariate linear combination HBA1C-EGFR has P-value for testing H0$_1$ of $<$ 0.001.

### 2014-07-18 Web_appendix.tex
Submission time

\documentclass[times, 11pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{graphicx}
\usepackage{amsmath, amsfonts, amssymb, mathrsfs, rotating, setspace,natbib}
\usepackage[left=1in,top=1in,right=1in,bottom=1in]{geometry}
%\renewcommand{\baselinestretch}{1.0}

\def \v{\mbox{Var}}
\def \E{\mathbb{E}}
\def \c{\mbox{Cov}}

\usepackage[usenames]{color}
\usepackage[colorlinks=true]{hyperref}
\hypersetup{linkcolor=blue,citecolor=blue,filecolor=black,urlcolor=blue}

%% Please use the following statements for 
%% managing the text and math fonts for your papers:
\usepackage{times}
\usepackage{amsmath, amssymb, amsthm, mathrsfs}
\usepackage{rotating}

%\usepackage{graphicx}
%\usepackage[cmbold]{mathtime}

\usepackage{natbib}
\linespread{1.5}
%% user defined macros

\def \v{\mbox{Var}}
\def \E{\mathbb{E}}
\def \c{\mbox{Cov}}

\newcommand{\bmX}{\mbox{\boldmath $X$}}
%\newcommand{\hatb0}{\mbox{$\widehat{\beta}_0$}}
\newcommand{\bmY}{\mbox{\boldmath $Y$}}
\newcommand{\bmy}{\mbox{\boldmath $y$}}
\newcommand{\bmZ}{\mbox{\boldmath $Z$}}
\newcommand{\bmz}{\mbox{\boldmath $z$}}
\newcommand{\bmx}{\mbox{\boldmath $x$}}
\newcommand{\bmQ}{\mbox{\boldmath $Q$}}
\newcommand{\bmU}{\mbox{\boldmath $U$}}
\newcommand{\bmu}{\mbox{\boldmath $u$}}
\newcommand{\bmV}{\mbox{\boldmath $V$}}
\newcommand{\bmI}{\mbox{\boldmath $I$}}
\newcommand{\bmW}{\mbox{\boldmath $W$}}
\newcommand{\bmD}{\mbox{\boldmath $D$}}
\newcommand{\bms}{\mbox{\boldmath $s$}}
\newcommand{\bma}{\mbox{\boldmath $a$}}
\newcommand{\bmb}{\mbox{\boldmath $b$}}
\newcommand{\bmq}{\mbox{\boldmath $q$}}
\newcommand{\bmf}{\mbox{\boldmath $f$}}
\newcommand{\bmA}{\mbox{\boldmath $A$}}
%\newcommand{\bmepsilon}{\mbox{\boldmath $\epsilon$}}
\newcommand{\Prob}{\mbox{Pr}}
\newcommand{\bmone}{\mbox{\bf 1}}
\newcommand{\bmzero}{\mbox{\bf 0}}
\newcommand{\bmbeta}{\mbox{\boldmath $\beta$}}
\newcommand{\bmdelta}{\mbox{\boldmath $\delta$}}
\newcommand{\bmtheta}{\mbox{\boldmath $\theta$}}
\newcommand{\bmpsi}{\mbox{\boldmath $\psi$}}
\newcommand{\bmphi}{\mbox{\boldmath $\phi$}}
\newcommand{\bmnu}{\mbox{\boldmath $\nu$}}
\newcommand{\bmgamma}{\mbox{\boldmath $\gamma$}}
\newcommand{\bmmu}{\mbox{\boldmath $\mu$}}
\newcommand{\bmepsilon}{\mbox{\boldmath $\epsilon$}}
\newcommand{\bmSigma}{\mbox{\boldmath $\Sigma$}}

%\newtheorem{proposition}[theorem]{Proposition}
\begin{document}
\author{Erin E. Gabriel, Michael C. Sachs and Peter B. Gilbert}
\title{Comparing and Combining Biomarkers as Principle Surrogates for Time-to-Event Clinical Endpoints: Supplementary Materials}
\maketitle

\section{Web Appendix A}

\subsection{Non-random Group of Subjects with Candidate Surrogate Measurements}
To account for the possible biased sampling of $S_{j}(1)$, when case-cohort or two-phase sampling is used in the treatment arm to obtain $S_j(1)$ measurements, one can use inverse sampling probability weights in both the fitting of the location-scale model and the fitting of the imputed likelihood. Those with the candidate PS measured have their likelihood contribution unweighted, and each imputed likelihood contribution is weighted by the inverse probability of measurement for the $s_{1,j,k}$(s) giving rise to that imputation. Thus, the estimated likelihood contribution for subjects missing $S_{j}(1)$ is an inverse probability weighted average of their imputed likelihood contributions. When estimating the proposed summary statistics, an unbiased sample, as we have in our simulations and motivating example, allows for empirical estimation of integrals with respect to $F_{S_1(1), \ldots, S_J(1)}(s_{1,1}, \dots, s_{1,J})$. If the group of subjects with the candidate surrogates measured is not a random sample from the treatment arm, but the sample of subjects having W measured is we suggest Monte Carlo estimation of the integrals by sampling from the weighted location-scale estimated $d\hat{F}_{S_{1}(1),\ldots,S_{J}(1)|W}(s_{1,1},\ldots,s_{1,J}|W)$ across the observed levels of W.

\subsection{Derivation of formula for $TPF(t|c)$}
\begin{eqnarray*}
TPF(t|c)&=&P[D(t)=1,\mathrm{I}_{\{\Delta(t|s_{1,1}, \dots, s_{1,J})>c\}}] \over P\{D(t)=1\}\\
&=&E[\{D(t)=1\}\mathrm{I}_{\{\Delta(t|s_{1,1}, \dots, s_{1,J})>c\}}] \over E\{D(t)=1\}\\
&=&{E[E\{D(t)=1|S_{1}(1)=s_{1,1}, \ldots, S_{j}(1)=s_{1,j}\}\mathrm{I}_{\{\Delta(t|s_{1,1}, \dots, s_{1,J})>c\}}] \over E[E\{D(t)=1|S_{1}(1)=s_{1,1}, \ldots, S_{j}(1)=s_{1,j}\}]}\\
&=&{E[\Delta(t|s_{1,1}, \dots, s_{1,J})\}\mathrm{I}_{\{\Delta(t|s_{1,1}, \dots, s_{1,J})>c\}}] \over E\{\Delta(t|s_{1,1}, \dots, s_{1,J})\}}.
\end{eqnarray*}
This deviation follows the time-independent derivation from \citet{Huang12b}.

\subsection{Violations of A6}
\begin{eqnarray*}
P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)=1)\frac{P(D(t)=1)}{P(D(t)=1)+P(D(t)=-1)}\\
-P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)=-1)\frac{P(D(t)=-1)}{P(D(t)=1)+P(D(t)=-1)}.
\end{eqnarray*}
When there are obvious violations of A6, this is clearly not a good approximation of what we hope to estimate in the non-monotone setting, the true positive fraction for benefit, as defined in \citet{Huang12b}, $TPF^{b}(t|c)=P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)=1)$. Thus, under clear violation of A6, the $TPF(t|c)$ estimate should not be calculated. Under minor potential undetectable violations of A6, $P(D(t)=-1)$ will be small and the estimator of $TPF(t|c)$ will be a reasonable approximation. For the slightly more complicated case of $FPF(t|c)$, the decomposition of the derivation under violation of A6 is given by:
\begin{eqnarray*}
&&P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)<1)\frac{P(D(t)<1)}{P(D(t)=-1)+P(D(t)<1)}\\
&-&P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)=-1)\frac{P(D(t)=-1)}{P(D(t)<1)+P(D(t)=-1)} .
\end{eqnarray*}
This will approximate one of two things under minor violations of A6 where $P(D(t)=-1)$ is small, either the false positive fraction for no-benefit $FPF^{N}(t|c)=P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)<1)$, or the false positive fraction for unaffected $FPF^{U}(t|c)=P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)=0)$, which depends on the $P(\Delta(t|s_{1,1}, \dots, s_{1,J})>c|D(t)=-1)$. This leads to two possible ROC curves, either of which are a reasonable and well defined summary statistic for comparison of candidate surrogates. This deviation follows the time-independent derivation from \citet{Huang12b}.

\subsection{Other summary measures considered}
\subsubsection{Positive Predictive Value ($PPV(t|c)$) and Negative Predictive Value ($NPV(t|c)$)}
One may wish to consider how well the risk model is predicting protection or lack of protection separately. For this, the concepts of positive predictive value and negative predictive value are useful. Time dependent PPV and NPV were defined for a risk model, $f(x)$, as $PPV(t|c)=Pr(Y^t=1|f(x) \geq c)$ and $NPV(t|c)=Pr(Y^t=0|f(x) \leq c)$ in \citep{Heagerty00}.

This concept can be extended to our setting using, a $\Delta(t|s_{1,j})$ of risk difference, where risk is based on the CDF, and $D(t)=Y^t(0)-Y^t(1)$. Defining $PPV(t|v)=Pr(D(t)=1|\Delta(t|s_{1,j})\geq c)$ and $NPV(t|v)=Pr(D(t)=0|\Delta(t|s_{1,j}) \leq c)$. If we assume no active harm, or monotonicity in treatment effect, $PPV(t|v)$ is the probability of protection given that $\Delta(t|s_{1,j})$ is greater than or equal to c. Similarly, if we assume no active harm, $NPV(t|v)$ is the probability of no effect of treatment given that $\Delta(t|s_{1,j})$ is less than or equal to c. These can also be defined and estimated via the quantile curve $R^t(v)$, as defined above, adapted from \citep{Gu10} by,
\begin{eqnarray*}
PPV(t|v)&=&\int_{v}^{1}R^t(\mu)d\mu/(1-v)\\
NPV(t|v)&=&1-\int_{0}^{v}R^t(\mu)d\mu/(v), 
\end{eqnarray*}
The values of $PPV(t|v)$ and $NPV(t|v)$ at a particular $v$ and $t$ can be used as summary statistics if one wishes, however we suggest that the full curves are of great interest. They differ from the time-dependent and covariate-specific $PPV_z^{'}$, outlined in \citep{Gabriel13}, as they are based on the quantiles of a risk difference rather than the quantiles of the surrogate itself. Because $PPV(t|v)$ and $NPV(t|v)$ are model based, they can be used to evaluate different risk models, compare potential PSs, or for PS evaluation when compared to the clinical outcome prevalence difference over the trial arms.

Positive predictive value, $PPV(t|c)$ and $NPV(t|v)$ have an appealing clinical interpretation given in \citep{Gu10} when we assume monotonicity of treatment activity. By applying Bayes theorem we have:  
\begin{eqnarray*} 
PPV(t|v) &=& {(\rho_0(t) - \rho_1(t))\over \{1-\rho_0(t) + \rho_1(t)\}}{TPR(t|v) \over FPR(t|v)} = {(\rho_0(t) - \rho_1(t))\over \{1-\rho_0(t) + \rho_1(t)\}} {
\mbox{Sensitivity}(t|v) \over 1-\mbox{Specificity}(t|v)}\\
NPV(t|v) &=& {\{1-\rho_0(t) + \rho_1(t)\} \over (\rho_0(t) - \rho_1(t))}{1-FPR(t|v) \over 1-TPR(t|v)}={\{1-\rho_0(t) + \rho_1(t)\} \over (\rho_0(t) - \rho_1(t))}{\mbox{Specificity}(t|v) \over 1-\mbox{Sensitivity}(t|v)},
\end{eqnarray*}
where $\rho_1(t)=Pr(Y^t(z)=1)$. We use the plug-in estimator for $PPV(t|c)$ and $NPV(t|c)$ given by:
\begin{eqnarray*}
\widehat{PPV}(t|v)&=&\int_{v}^{1}\hat{R}^t(\mu)d\mu/(1-v)\\
1-\widehat{NPV}(t|v)&=&\int_{0}^{v}\hat{R}^t(\mu)d\mu/(v).
\end{eqnarray*}

Particular points on the $PPV(t|v)$ and $NPV(t|v)$ curves can be compared via confidence interval to the difference in prevalence of potential clinical outcomes, $(\rho_0(t) -\rho_1(t))$ for the same time $t$. For example, if $(\rho_0(t) -\rho_1(t))$ at year $t$ is not contained in the CI for $\widehat{PPV}(t|0.8)$, this suggests that there is evidence to reject the null that there is not variation in predictive power of the model over the potential surrogate. Said another way, this is evidence to support that there is higher probability of protection when the model for risk difference is greater than the 80th percentile of predicted risk difference. Most simply, there is evidence of increased probability of protection given that the model predicts protection. A PS with causal necessity will have a $NPV(t|v)$ of one for some low $v$, but any $\widehat{NPV}(t|v)$ confidence interval that does not cover $1-(\rho_0(t) -\rho_1(t))$ provides evidence that the risk model is predictive and therefore the candidate biomarker has some value as a PS.

\subsubsection{Partial Total Gain (pTG(t))}
The definition of $PPV(t|v)$ and the suggested comparison to $(\rho_0(t) -\rho_1(t))$ leads one to the question if these two concepts can be combined into a single summary statistic similar to STG. We consider the partial total gain (pTG) of \citep{Sachs12} for this purpose. A simple and intuitive definition pTG, given $PPV(t|v)$, might be, 
\begin{eqnarray}
pTG(t|v)=\int_{v}^{1}|R^t(\mu)-(\rho_0(t) -\rho_1(t))|d\mu/(1-v) \label{orgptg}.
\end{eqnarray}
The pTG is defined more flexibly for a binary endpoint in the time-independent biomarker framework in \citep{Sachs12} over the lower tail $(0, \nu)$ and upper tail $(d, 1)$ of the quantile curve, $R(u)$, for a given risk model by, 
\[
pTG(B) = \frac{1}{v\rho + (1-d)(1-\rho)}\int_{B} |R(u) - \rho|\, d\, u, 
\]
where $B = \{(0, v) \bigcup (d, 1)\}$ and $\rho$ is prevalence. The $pTG(B)$ can be standardized so that the measure takes values between 0 and 1. This can be extended to the causal framework for a time-to-event outcome by again using the quantile curve of the risk difference, $R^t(u)$. The standardized pTG in our setting is given by,  
\[
pTG(t|B) = pTG(t|d,v)= \frac{1}{v(\rho_0(t)-\rho_1(t)) + (1-d)(1-(\rho_0(t)+\rho_1(t))}\int_{B} |R^t(u) - (\rho_0(t)-\rho_1(t))|\, d\, u \cdot
\]
The thresholds are quantiles of risk difference $v = R^{t}(c)$ and $d = R^{t}(q)$; if one sets $v=0$ we arrive at the standardized version of equation \ref{orgptg}. As given in \citep{Sachs12}, if we assume that $v<(\rho_0(t)-\rho_1(t))$, $d>(\rho_0(t)-\rho_1(t))$ and there is no individual active harm by treatment, we can define the unstandardized $pTG(t|B)$ in terms of $PPV(t|q)$ and $NPV(t|c)$.

The $pTG(t|B)$ is of interest in the surrogate evaluation framework for comparison of potential PS quality in cases where STG(t) is not useful or is too similar. Unlike the $TG(t)$ and $STG(t)$ which are summaries over the full quantile curve of the CEP, the $pTG(t|B)$ allows one to summarize a surrogate on a particular range of the risk difference. Finding a range on which the potential surrogate is of high-quality is the main goal of specific PS comparison. Thus, the $pTG(t|B)$ could be a more interesting summary statistic than those that consider the entire range of risk difference. Figure 1 of \citep{Sachs12} illustrates a scenario where pTG is beneficial over STG.

Similarly to STG(t), inference on $pTG(t|d,v)$, is most useful for comparison of candidate PS and risk models, rather than for evaluation of specific PS. Confidence intervals will almost never include zero, suggesting that any candidate PS has some partial surrogate value; CI should rather be compared to those of other candidate PS. When sets of CI within the same trial for different candidate PS do not overlap this provides some evidence that one of the PS is better at classifying treatment effect groups, with higher $pTG(t|d,v)$ implying better classification. However, just as with STG(t) P-values for the difference between two $\widehat{pTG}(t|d,v)$ being different than zero are the preferred means of inference.

We use the plug-in estimator for $pTG(t|d,v)$ when the CEP is risk difference this is given by:
\[
\widehat{pTG}(t|d,v) = \frac{1}{v(\widehat{\rho}_0(t)-\widehat{\rho}_1(t)) + (1-d)(1-(\widehat{\rho}_0(t)+\widehat{\rho}_1(t))}\int_{B} |\widehat{R}^t(u) - (\widehat{\rho}_0(t)-\widehat{\rho}_1(t))|\, d\, u.
\]

\section{Time-dependent and Surrogate-level-based PPV}
Although not useful for multivariate PS, the time-dependent covariate-specific positive predictive value (PPV) curve of \citet{Zheng08} and \citet{Zheng10} is a different way to evaluate univariate candidate PS. The PPV compares predictive value of a diagnostic test at various thresholds. As thresholds are commonly used in practice for validated surrogates of protection, a modified version of the PPV is appealing for PS quality evaluation and comparison.

To this end, we modify the meaning of the time-dependent and covariate-specific PPV of \citet{Zheng10} to be the conditional probability of survival above a given threshold of $S_j(1)$ and call it $PPV_z^{'}$. Under assumption A2, $F^{-1}_{S(1)|Z}(v) = F^{-1}_{S_j(1)}(v)\equiv c(v)$ for any $v \in [0,1]$. We therefore define our time-dependent and treatment-specific $PPV_z^{'}$ as $PPV^{'}_z(t|v)=P\{T(z) > t | S(1) \geq c(v), T(1) \geq \tau, T(0) \geq \tau\}$.
Via modification of the derivation given in \citet{Moskowitz04a} we can write, 
\[
PPV^{'}_z(t|v)=(1 - v)^{-1}  \int_{c(v)}^{\infty} {Q_z(t|s_1)dF_{S_j(1)}}(s_{1,j}).
\]
The plug-in estimator for $PPV_z^{'}(t|v)$ is given by:
\[
\widehat{PPV}^{'}_z(t|v) = (1 - v)^{-1} \int_{c(v)}^{\infty} {\widehat{Q}_z(t|s_1)d\widehat{F}_{S_j(1)}}(s_{1,j}) \hspace{2mm} \mbox{for} \hspace{1mm} t \geq \tau.
\]
Comparisons over $z$ using the $PPV^{'}_{z}(t|v)$ are causal, as they are based on the survivor function and not the hazard. A meaningful comparison of $PPV^{'}_z(t|v)$ for potential surrogate evaluation is the TE above a given threshold of $S(1)$ defined by:
\[
TE(t|v+)\equiv 1-RR(t|v+)\equiv 1-{1-PPV^{'}_1(t|v) \over 1-PPV^{'}_0(t|v)}, 
\]
where $RR$ is the conditional cumulative relative risk. This estimand measures vaccine efficacy in the subgroup of individuals who would achieve at least the threshold value of $S(1)$. Figures depicting estimated $TE(t|v+)$ versus $v$ are of interest for describing surrogate quality and determining optimal thresholds. Steps in the curve suggest points where thresholds could be considered, as they indicate rapid gains or losses in TE above a given threshold; smooth curves suggest a threshold may not be a useful view of the candidate PS.

Although A2 implies $c_z(v)=c(v)$, the $S_j(1)$ threshold may not be independent of all baseline covariates. One could easily extend these estimands and estimators to include other covariates by applying \citet{Zheng10}. When the $S(1)$ threshold is not independent of baseline covariates the plug-in estimator is given by:
\begin{singlespace}
\[
\widehat{PPV}^{'}_{z}(t|v,w) = (1 - v)^{-1} \int_{c_w(v)}^{\infty} {\widehat{S}_z(t|s_1,w)d\widehat{F}_{S(1)|W}}(s_1).
\]
\end{singlespace}
The $PPV^{'}_z$ curve is introduced in the main article via the time-dependent Weibull model; however it is easy to see how a time-independent version of the $PPV^{'}_z$ could be estimated via the Cox model or parametric exponential model. The $PPV^{'}_z$ curve could also be used for binary clinical endpoints by modification of the PPV of \citet{Moskowitz04b}, to fit the surrogate evaluation framework. For this illustration let our contrast of interest between the arms of the trial be the treatment efficacy (TE) $1 - {risk_1 / risk_0}$.

Finding a threshold of the potential surrogate above which there is high TE is of interest for surrogate evaluation and vaccine improvement in Phase I and IIa trials, but it is not necessary that this threshold be based on the quantiles of the potential surrogate. The $PPV_z^{'}$ can be determined for fixed cutoffs of $S(1)$ by changing the limits of integration from $c(v)$ to the fixed cutoff, $c$. This version of the $PPV_z^{'}$ may be advantageous for evaluation of potential surrogates with a defined protective level of immune response.

Basing the threshold on the quantiles of $S(1)$ standardizes the curves for comparison between biomarkers. For example, for a desired level of TE, $\eta$, and a candidate surrogate $S^{1}$, let $v^{1,\eta}$ be the lowest percentile of $S^{1}$ such that $TE_1(t|v^{1,X}+)=\eta$. If this is lower than $v^{2,\eta}$ for candidate surrogate $S^{2}$ it means that more of the population obtains the desired TE as measured by $S^{1}$ than $S^{2}$. This is only one way to look at the curves and comparisons based on the full $TE(t|v+)$ curve of each candidate must always be considered. Lower $\hat{v}^{k,X}$ does not imply that TE has greater variation over the candidate surrogate; lower $\hat{v}^{k,X}$ will often imply a flatter TE curve, making it a worse PS by our definition.

\section{Appendix B}

\begin{sidewaystable}[ht]
\begin{center}
\caption{Performance of Accuracy Measures Using Univariate BIP for Multivariate PS}
\begin{tabular}{lccc}
  \hline
&$R_{s_{1,j}w}^{**}$& $\widetilde{STG}^{*}=0.46$ & $\widetilde{STG}^{*}=0.81$ \\
  \hline
Bias $CDF^{3}_{\Delta}(c)$& 0.8 & -0.013 (0.035)  & 0.001 (0.028)\\ 
 &0.5& -0.01 (0.033)  & 0.01 (0.04) \\ 
Bias $TPF(3|c)$& 0.8& -0.068 (0.166)  & -0.008 (0.114)  \\ 
 &0.5& -0.072 (0.187)  & -0.063 (0.257) \\ 
Bias $FPF(3|c)$& 0.8&  0.001 (0.026) & 0.009 (0.02) \\ 
 &0.5&  -0.001 (0.004)  & 0.006 (0.028) \\ 
Bias $STG(3)$& 0.8& -0.011 (0.277) & -0.009 (0.197)  \\
  &0.5 & -0.024 (0.285)  & -0.019 (0.299) \\ 
Bias $\widetilde{STG}$ & 0.8 & -0.037 (0.221)  & -0.003 (0.058) \\
 & 0.5 & -0.037 (0.238) & -0.05 (0.151)  \\
   \hline
\multicolumn{4}{p{9in}}{Bias entries for the $TPF(3|c)$ and $FPF(3|c)$ curves are average bias over the simulations and 2000 points on each simulated curve (empirical standard deviation of the estimated statistic over the simulations). *Multivariate candidate PS, **For multivariate surrogate candidates $R_{s_{1,j}w}=x$ indicates the set $(R_{s_{1,1}w},R_{s_{1,2}w})=(x,x)$. }
\end{tabular}
\end{center}
\end{sidewaystable}

The large standard deviations of the STG, are caused by poor convergence for ~10\% of the simulations. In each case one or more the risk model parameter estimates are more than 2 standard deviations from the true value. Although this does not cause bias overall, as it is symmetric about the true coefficient values,  it does increase the standard deviation. When using bootstrap standard errors for inference, this will cause lower efficiency in practice. The use of a vector BIP, remedies this problem as can be seen in the main simulations. These convergence issues may also be alleviated by using the EM algorithm as suggested in \citet{Huang11}, even for a univariate BIP.

\begin{sidewaystable}[ht]
\begin{center}
\caption{Power of difference in $STG(t)$ tests for $t=3$ years \label{Power}}
\begin{tabular}{rrccccc}
  \hline
 & $R_{s_{1,j}w}$ & $STG(3)=0.1$  & $STG(3)=0.24$  & $STG(3)^{\#}=0.47$& $STG(3)^{\#*}=0.23$  & $STG(3)^{\#*}=0.51$\\
  \hline
$STG(3)=0.1$ &0.5 & 4.8 & 14.1 & 73.1 & 17.5 & 98.3 \\ 
   &0.8 &6.1 & 9.3 & 36.7 & 10.6 & 91.0 \\ 
$STG(3)=0.24$ &0.5&  & 4.4 & 21.5 & 5.5 & 73.6 \\ 
   &0.8 &  & 5.2 & 20.1 & 4.9 & 60.3 \\ 
$STG(3)=0.47$&0.5 &  &  & 6.0 & 23.2 & 4.8 \\ 
   &0.8&  &  & 5.2 & 19.0 & 5.2 \\ 
$STG(3)^{*}=0.23$&(0.5, 0.5) &  &  &  & 5.1 & 83.5 \\ 
   &(0.8, 0.8) &  &  &  & 5.9 & 81.5 \\ 
$STG(3)^{*}=0.51$&(0.5, 0.5)&   &  &  &  & 5.1 \\ 
   &(0.8, 0.8)&  &  &  &  & 5.5 \\ 
   \hline
\multicolumn{7}{p{9in}}{Wald test for comparing two candidate surrogates by testing $STG_1(3)-STG_2(3)=0$. *Multivariate candidate PS, **For multivariate surrogate candidates $\rho_{s_{1,j}w}$ stands for $(R_{s_{1,1}w},R_{s_{1,2}w})$. \#indicates that the candidate surrogate was generated using a censored normal distribution.}
\end{tabular}
\end{center}
\end{sidewaystable}

Let us consider a univariate candidate surrogate for which we define the risks based on the hazard function, $risk_{z}(t|s_1)\equiv\lambda_z(t|s_1)\equiv{f_z(t|s_1)/Q_z(t|s_1)}$, where $f_z(t|s_1)$ is the conditional probability density function of $T$, assuming its existence. Contrasts of risk are the causal estimands of interest for evaluation of PS. One such contrast is TE defined by $TE(t|s_1) \equiv 1 - {risk_1(t|s_1) / risk_0(t|s_1)}$, as given in \citet{Gabriel13}. Basing risk on the hazard function makes comparisons of $risk_{1}(t|s_1)$ and $risk_{0}(t|s_1)$ non-causal, as outlined in \citet{Hernan10}, because the risks condition on different sets $\{T(z)>T\}$ by arm. We consider the hazard-based TE curve as it allows for discrimination between levels of time-dependence in the data using the methods we are about to describe. As the hazard-based TE can be time-independent or time-dependent depending on the parameterization of the hazards, whereas the CDF-based TE is always time-dependent, tests can be preformed to determine if there is significant time-dependence as well as the type of time dependence present. Under the same assumptions A1-A4 as the main text, we can estimate these risks by assuming a parametric model for the clinical outcome.

We assume a Weibull model for the conditional pdf $g(\cdot|\cdot)$, of $T$ given $Z$ and $S(1)$, which parametrizes both the scale, $\gamma=(\gamma_{00}, \gamma_{10},\gamma_{01}, \gamma_{11})$, and shape, $\beta=(\beta_{00}, \beta_{10},\beta_{01}, \beta_{11})$, components of the Weibull model with treatment, $Z$, and potential surrogate, $S(1)$. Specifically, our Weibull assumption A5 states that 
\begin{itemize}
\item A5: $g(t|\gamma, \beta, z, s_1, \delta) =  \lambda_z(t|\gamma, \beta, s_1)^{\delta} Q_z(t|\gamma, \beta, s_1)$
\end{itemize}
where $Q_z(t|\gamma, \beta, s_1)$ is the parametrized conditional Weibull survivor function for the treatment arm $Z=z$, and the conditional hazard function is given by:
\begin{eqnarray} 
\lambda_z(t|\gamma, \beta, s_1)={\mbox{exp}\left(\beta_{z0} + \beta_{z1} s_1\right) \over \mbox{exp}\left(\gamma_{z0} + \gamma_{z1} s_1 \right)} \left\{{t \over \mbox{exp}\left(\gamma_{z0} +\gamma_{z1} s_1\right)} \right\}^{\{\mbox{exp}(\beta_{z0} + \beta_{z1} s_1)-1\}} \label{32}
\end{eqnarray}
The identifiability of this model as well as the meaning of the coefficients is given in \citet{Gabriel13}. We can estimate this model using the same semi-parametric EML outlined in the main text and under the BIP-only augmentation and in combination with close-out placebo vaccination augmentation \citep{Follmann06, Gilbert08, Huang11}. Based our ability to estimate this model and the meanings of the parameters we propose a three-step process for evaluating a potential PS using the above estimation method.

\begin{itemize} 
\item Step 1: Fit the time-dependent TE Weibull model via parametric EML; determine the EML estimates by maximizing $\mbox{L}(\beta, \gamma, \widehat{\nu})$, where $\mbox{L}(\beta,\gamma,\nu)$ is defined by assumption A5 of the main text.
\item Step 2: Test for time-varying conditional treatment efficacy, H0$_1$: $TE(t|s_1) = TE(s_1)$, by testing $(\beta_{11}-\beta_{01})=(\beta_{10}-\beta_{00})=0$.
\item Step 3: If the data support H0$_1$, fit the time-independent TE Weibull model, outlined below. Use estimates from this model for figures and inference on surrogate quality. If the data support rejection of H0$_1$, use the time-dependent TE model estimates for figures and inference on surrogate quality. 
\end{itemize} 
The testable null hypotheses of interest are:
\vspace{-1mm}
\begin{itemize}
\item H0$_1$: $TE(t|s_1) = TE(s_1)$, Null equivalent $(\beta_{11}-\beta_{01})=(\beta_{10}-\beta_{00})=0$,
\item H0$^{*}_2$: $TE(s_1) = TE$, Null equivalent $\beta^{*}_{1}=\gamma^{*}_{11}-\gamma^{*}_{01}=0$,
\item H0$_2$: $TE(t|s_1) = TE(t)$, Null equivalent $\beta_{11}=\beta_{01}=\{\gamma_{01}\mbox{exp}(\beta_{00})-\gamma_{11}\mbox{exp}(\beta_{10})\}=0$,
\item H0$_3$: No time-dependence that is associated with S(1), Null equivalent\\ $(\beta_{11}-\beta_{01})=0 \bigcap \{(\beta_{10}-\beta_{00}=0) \bigcup (\gamma_{11}-\gamma_{01}=0)\}$
\item H0$_4$: No time-dependence in $TE(t|0)$, Null equivalent $\beta_{10}-\beta_{00}=0$,
\end{itemize}
The justifications for these tests in given in \citet{Gabriel13}.

Let $\beta_{10}=\beta_{00}=\beta^{*}_{0}$ and $\beta_{11}=\beta_{01}=\beta^{*}_{1}$. Then the time-independent TE model is given by:
\begin{eqnarray*}
risk^{HZ}_{z}(s_1,t; \beta^{*}, \gamma^{*})&=&\lambda^{*}_z(t|s_1) \\
&=&{\mbox{exp}\left(\beta^{*}_{0}+\beta^{*}_{1} s_1\right) \over \mbox{exp}\left(\gamma^{*}_{z0} + \gamma^{*}_{z1} s_1  \right)}\left\{{t \over \mbox{exp}\left(\gamma^{*}_{z0} +\gamma^{*}_{z1} s_1 \right)}\right\}^{\left\{\mbox{exp}\left(\beta^{*}_{0} + \beta^{*}_{1} s_1\right) - 1\right\}}.
\end{eqnarray*}

We illustrate our methods in simulated data following a 1:1 randomized two-arm trial with 2000 subjects per treatment-arm using the various case-control sampling designs for CPV and BIP. Suppose the conditional cdf of $T$ given $S(1)$ and $Z$ follows a Weibull model and that $\{S(1),W\}$ follows a bivariate normal model with correlation $\rho_{WS}$. Information lost to drop out occurs completely at random, and occurs at a rate of 5\% per year. Event times are censored at 3 years post $\tau$, at which time the trials have 50\% TE on average, with an average of 104 treatment group infections and 208 placebo group infections over the 1000 simulated trials. This follows the HIV vaccine trial design proposed in \citet{Gilbert11}.

We investigate three time-independent Weibull models for $T$ given $S(1)$ and $Z$ that give three different PS quality levels, a high quality surrogate, a marginal quality surrogate and a useless surrogate. For the useless surrogate expressed by differing amounts of variation in $TE(t|s_1)$ over $s_1$ and ACN. These three scenarios are used to investigate the bias of the EML estimator, the type 1 error rate of H0$_1$ and the power and type 1 error rate of H0$_2$ and H0$^{*}_2$, in the time-independent setting. We also consider two surrogate quality levels under a time-dependent TE Weibull model for $T$ given $S(1)$ and $Z$ with time-dependence in $TE(t|0)$. We investigate a high-quality surrogate and a marginal-quality surrogate under this type of time-dependence. Finally, we consider a high-quality surrogate and a marginal-quality surrogate under a time-dependent TE Weibull model for $T$ given $S(1)$ and $Z$ where there is time dependence in TE that is both associated with the surrogate quality and in $TE(t|0)$. The four time-dependent scenarios are used to investigate the bias of the estimator and testing power to reject H0$_1$-H0$_4$ in the presence of time-varying TE.

For each of the Weibull models we consider several different types of case-control sampling of $S(1)$, $S^{CO}$ and $W$ and different levels of $\rho_{WS}$. We display results from six different types of case-control sampling of $S(1)$/$S^{CO}$ and $W$ all for $\rho_{WS}$=0.8. This is only slightly higher than our motivating scenario where there is a $\hat{\rho}_{WS}$=0.701. The six case-control sampling scenarios considered are broken into two groups of three to consider the issues of case-control sampling of $S(1)$ and $S^{CO}$ separately. First, we consider case-control sampling of $S^{CO}$, holding sampling of $S(1)$ constant at full, and varying the sampling of $S^{CO}$ between full, 1:5 case-control and no sampling of $S^{CO}$. We then investigate the effects of sub-sampling of $S(1)$, by holding case-control sampling of $S(1)$ at 1:5 and again varying sampling of $S^{CO}$ between full, 1:5 case-control and no sampling. Table 1 below displays the bias for points on the TE curve and Table 2 displays the results of null hypotheses of interest.

The findings suggest that the semi-parametric EML method works at least as well as the fully-parametric EML method of \citet{Gabriel13} for unbiased estimation of the TE curve, possibly being slightly less sensitive to sub-sampling of the CPV measure $S^{C}$, and to lower correlations of the BIP, Table 3 and Table 4. Table 4 also display the results of bias for the various summary statistics including $STG(t)$ for time-varying surrogate quality and TE waning. As suggested in the main text, these estimators are unbiased even when there is time-dependence.

\begin{sidewaystable} [H]
\caption{Percent Bias: two-arm trial for given sampling of $W$, $S^{C}$ and $S(1)$; for $W$ and $S(1)$ correlation (0.8) semi-parametric EML}
\begin{tabular}{l|ccc|cc|cc|ccc|cc|cc}
\hline
&\multicolumn{7}{c}{fill sampling $W$, full sampling $S^{C}$ and $S(1)$} &\multicolumn{7}{c}{fill sampling $W$, 5:1 $S^{C}$ and full $S(1)$}\\
Estimate&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}\\
&No Val&Some Val&High Val&Some&High&Some&High&No&Some&High&Some&High&Some&High\\
\hline
$\widehat{TE}(1|2)$& 0.40 & 0.30 & -0.10 & 0.40 & 0.50 & 1.00 & 0.50 & 0.40 & 0.30 & -0.10 & 0.50 & 0.30 & 0.90 & 0.60 \\
$\widehat{TE}(1|4)$& 1.90 & 0.60 & 0.30 & -0.10 & 0.10 & 0.50 & 0.20 & 1.60 & 0.40 & 0.30 & -0.20 & -0.00 & 0.30 & 0.30 \\
$\widehat{TE}(2.5|4)$ & 1.90 & 0.60 & 0.30 & -0.70 & 0.10 & -0.40 & 0.10 & 1.60 & 0.40 & 0.30 & -1.00 & 0.10 & -0.60 & 0.00 \\
$\widehat{TE}(2.5|0.8+)$&2.02 & 0.85 & 0.32 & -1.60 & -0.52 & -0.88 & 1.47 & 0.90 & 0.28 & -1.48 & -0.47 & -1.07 & -0.89  \\
\hline
&\multicolumn{7}{c}{fill sampling $W$, no $S^{C}$ and full $S(1)$} &\multicolumn{7}{c}{fill sampling of $W$,  full $S^{C}$ and 5:1 $S(1)$}\\
Estimate&\multicolumn{3}{|c}{Time Ind}&\multicolumn{2}{|c}{TE wane}&\multicolumn{2}{|c}{Both wane}&\multicolumn{3}{|c}{Time Ind}&\multicolumn{2}{|c}{TE wane}&\multicolumn{2}{|c}{Both wane}\\
&No Val&Some Val&High Val&Some&High&Some&High&No&Some&High&Some&High&Some&High\\
\hline
$\widehat{TE}(1|2)$&0.40 & 0.20 & -0.10 & 0.40 & 0.60 & 0.80 & 0.50 & 0.30 & 0.40 & 0.10 & 0.30 & 0.60 & 1.40 & 0.80 \\
$\widehat{TE}(1|4)$&1.70 & 0.60 & 0.10 & -0.40 & 0.30 & 0.30 & 0.30 & 1.10 & 1.00 & -1.70 & -0.20 & 0.20 & 0.10 & 0.20 \\
$\widehat{TE}(2.5|4)$&1.70 & 0.60 & 0.10 & -1.10 & 0.40 & -1.20 & 0.00 & 1.10 & 1.00 & 0.20 & -1.30 & 0.10 & -1.20 & 0.10 \\ 
$\widehat{TE}(2.5|0.8+)$& 1.10 & 0.77 & 0.36 & -1.49 & -0.86 & -0.91 & -0.66 & 1.42 & 1.07 & 0.49 & -0.95 & -0.38 & -1.15 & -0.61 \\ 
\hline
\hline
&\multicolumn{7}{c}{fill sampling $W$, 5:1 $S^{C}$ and $S(1)$} &\multicolumn{7}{c}{fill sampling $W$, no $S^{C}$ and 5:1 $S(1)$}\\
Estimate&\multicolumn{3}{|c}{Time Ind}&\multicolumn{2}{|c}{TE wane}&\multicolumn{2}{|c}{Both wane}&\multicolumn{3}{|c}{Time Ind}&\multicolumn{2}{|c}{TE wane}&\multicolumn{2}{|c}{Both wane}\\
&No Val&Some Val&High Val&Some&High&Some&High&No&Some&High&Some&High&Some&High\\
\hline
$\widehat{TE}(1|2)$& 0.30 & 0.30 & 0.20 & 0.30 & 0.40 & 1.50 & 0.80 & 0.40 & 0.40 & 0.10 & 0.30 & 0.40 & 1.00 & 0.60 \\
$\widehat{TE}(1|4)$& 1.10 & 0.70 & 0.20 & -0.40 & 0.10 & 0.10 & 0.30 & 1.40 & 0.60 & 0.10 & -0.40 & 0.20 & 0.20 & 0.30 \\  
$\widehat{TE}(2.5|4)$& 1.10 & 0.70 & 0.20 & -0.80 & -0.20 & -0.50 & 0.20 & 1.40 & 0.60 & 0.10 & -1.10 & 0.00 & -0.60 & 0.10 \\ 
$\widehat{TE}(2.5|0.8+)$& 1.02 & 0.95 & 0.36 & -1.20 & -0.39 & -1.25 & -0.79 & 0.70 & 1.09 & 0.26 & -1.54 & -0.50 & -0.96 & -0.69 \\ 
\hline
\end{tabular} 
\end{sidewaystable}

\begin{sidewaystable} [H]
\caption{Proportion of Rejections: two-arm trial for given sampling of $W$, $S^{C}$ and $S(1)$; $W$, $S(1)$ correlation (0.8) semi-parametric EML}
\begin{tabular}{l|ccc|cc|cc|ccc|cc|cc}
\hline
&\multicolumn{7}{c}{fill sampling $W$, full sampling $S^{C}$ and $S(1)$} &\multicolumn{7}{c}{fill sampling $W$, 5:1 $S^{C}$ and full $S(1)$}\\
Null&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}\\
&No Val&Some Val&High Val&Some&High&Some&High&No&Some&High&Some&High&Some&High\\
\hline
$\mbox{PH}^{\hspace{0.5mm}a}$&0.05&0.04&0.05&0.37&0.42&0.53&0.59&-&-&-&-&-&-&-\\
${\mbox{H0}_1}^{\hspace{0.5mm}b}$&0.05 & 0.05 & 0.06 & 0.36 & 0.45 & 0.81 & 0.71 &0.05 & 0.06 & 0.06 & 0.36 & 0.42 & 0.78 & 0.65\\
${\mbox{H0}_1}^{\hspace{0.5mm}c}$& 0.04 & 0.04 & 0.04 & 0.39 & 0.54 & 0.69 & 0.69 & 0.04 & 0.04 & 0.04 & 0.36 & 0.54 & 0.68 & 0.66\\
${\mbox{H0}_2}^{\hspace{0.5mm}d}$&0.05 & 0.72 & 0.99 & 0.09 & 0.25 & 0.04 & 0.10 &0.05 & 0.69 & 0.99 & 0.09 & 0.22 & 0.04 & 0.10\\
${\mbox{H0}_2}^{\hspace{0.5mm}e}$& 0.05 & 0.67 & 0.99 & 0.36 & 0.90 & 0.88 & 0.99& 0.05 & 0.66 & 0.99 & 0.35 & 0.89 & 0.87 & 0.99\\
${\mbox{H0}_3}^{\hspace{0.5mm}f}$&0.05 & 0.07 & 0.07 & 0.06 & 0.05 & 0.39 & 0.24 & 0.05 & 0.07 & 0.07 & 0.06 & 0.05 & 0.40 & 0.25 \\ 
${\mbox{H0}_4}^{\hspace{0.5mm}g}$&0.05 & 0.04 & 0.04 & 0.27 & 0.47 & 0.21 & 0.35 & 0.05 & 0.03 & 0.03 & 0.26 & 0.45 & 0.21 & 0.33 \\ 
\hline
&\multicolumn{7}{c}{fill sampling $W$, no $S^{C}$ and full $S(1)$} &\multicolumn{7}{c}{fill sampling of $W$,  full $S^{C}$ and 5:1 $S(1)$}\\
Null&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}\\
&No Val&Some Val&High Val&Some&High&Some&High&No&Some&High&Some&High&Some&High\\
\hline
${\mbox{H0}_1}^{\hspace{0.5mm}b}$& 0.04 & 0.04 & 0.04 & 0.38 & 0.54 & 0.69 & 0.68 & 0.04 & 0.04 & 0.04 & 0.36 & 0.53 & 0.66 & 0.65 \\ 
${\mbox{H0}_2}^{\hspace{0.5mm}e}$&0.05 & 0.68 & 0.99 & 0.36 & 0.90 & 0.88 & 0.99 & 0.05 & 0.60 & 0.99 & 0.32 & 0.88 & 0.84 & 0.99 \\ 
${\mbox{H0}_3}^{\hspace{0.5mm}f}$&0.06 & 0.07 &0.07 & 0.06 & 0.04 & 0.39 & 0.24 & 0.05 & 0.06 & 0.06 & 0.07 & 0.05 & 0.36 & 0.24 \\
${\mbox{H0}_4}^{\hspace{0.5mm}g}$&0.04 & 0.03 & 0.03 & 0.28 & 0.46 & 0.21 & 0.35 & 0.05 & 0.06 & 0.06 & 0.25 & 0.44 & 0.21 & 0.32 \\ 
\hline
\hline
&\multicolumn{7}{c}{fill sampling $W$, 5:1 $S^{C}$ and $S(1)$} &\multicolumn{7}{c}{fill sampling $W$, no $S^{C}$ and 5:1 $S(1)$}\\
Null&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both wane}\\
&No Val&Some Val&High Val&Some&High&Some&High&No&Some&High&Some&High&Some&High\\
\hline
${\mbox{H0}_1}^{\hspace{0.5mm}b}$& 0.04 & 0.04 & 0.07 & 0.37 & 0.51 & 0.65 & 0.64 & 0.06 & 0.04 & 0.07 & 0.37 & 0.52 & 0.66 & 0.67 \\ 
${\mbox{H0}_2}^{\hspace{0.5mm}e}$& 0.04 & 0.63 & 0.99 & 0.33 & 0.88 & 0.85 & 0.99 & 0.05 & 0.66 & 0.99 & 0.36 & 0.88 & 0.85 & 0.99 \\
${\mbox{H0}_3}^{\hspace{0.5mm}f}$ & 0.03 & 0.06 & 0.05 & 0.06 & 0.05 & 0.37 & 0.25 & 0.03 & 0.06 & 0.06 & 0.06 & 0.06 & 0.37 & 0.25 \\  
${\mbox{H0}_4}^{\hspace{0.5mm}g}$& 0.04 & 0.03 & 0.06 & 0.26 & 0.43 & 0.21 & 0.32 & 0.05 & 0.04 & 0.05 & 0.26 & 0.44 & 0.21 & 0.33\\ 
\hline
\normalsize
\end{tabular}
\end{sidewaystable}

\begin{table} [H]
\caption{Comparison Over EML Methods and BIP Correlation H0$_2$ and H0$_1$: full sampling $W$, full $S^{C}$ and $S(1)$ \label{compare_metods}}
\begin{tabular}{l|ccc|cc|cc}
\multicolumn{8}{c}{Proportion of rejection H0$_2$}\\
&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both Wane}\\
&No Val&Some Val&High Val&Some&High&Some&High\\
\hline
\multicolumn{8}{c}{$W$, $S(1)$ correlation (0.8)}\\
\hline
Parametric EML&0.5&0.68&0.99&0.49&0.93&0.90&0.99\\
Semi-parametric EML&0.5&0.67&0.99&0.36&0.91&0.88&0.99\\
\hline
\multicolumn{8}{c}{$W$, $S(1)$ correlation (0.5)}\\
\hline
Parametric EML&0.05 & 0.36 & 0.99 & 0.24 & 0.79 & 0.70 & 0.96 \\ 
Semi-parametric EML& 0.06 & 0.41 & 0.99 & 0.24 & 0.77 & 0.70 & 0.96 \\ 
\hline
\multicolumn{8}{c}{$W$, $S(1)$ correlation (0.25)}\\
\hline
Parametric EML& 0.06 & 0.16 & 0.88 & 0.14 & 0.51 & 0.46 & 0.78 \\ 
Semi-parametric EML& 0.05 & 0.17 & 0.90 & 0.16 & 0.45 & 0.42 & 0.77 \\
\hline
\hline
\multicolumn{8}{c}{Proportion of rejection H0$_1$}\\
&\multicolumn{3}{c}{Time Ind}&\multicolumn{2}{c}{TE wane}&\multicolumn{2}{c}{Both Wane}\\
&No Val&Some Val&High Val&Some&High&Some&High\\
\hline
\multicolumn{8}{c}{$W$, $S(1)$ correlation (0.8)}\\
\hline
Parametric EML&0.04&0.05&0.07&0.51&0.60&0.71&0.70\\
Semi-parametric EML&0.04&0.04&0.05&0.39&0.54&0.69&0.69\\
\hline
\multicolumn{8}{c}{$W$, $S(1)$ correlation (0.5)}\\
\hline
Parametric EML& 0.05 & 0.05 & 0.06 & 0.23 & 0.35 & 0.47 & 0.47 \\
Semi-parametric EML& 0.05 & 0.05 & 0.05 & 0.27 & 0.39 & 0.48 & 0.50 \\ 
\hline
\multicolumn{8}{c}{$W$, $S(1)$ correlation (0.25)}\\
\hline
Parametric EML& 0.05 & 0.02 & 0.03 & 0.17 & 0.25 & 0.35 & 0.35 \\ 
Semi-parametric EML& 0.05 & 0.04 & 0.04 & 0.21 & 0.28 & 0.36 & 0.37 \\ 
\hline
\end{tabular}
\end{table}

\begin{sidewaystable} [H]
\caption{Percent Bias: summary statistic comparison over methods two-arm trial; $W$, $S(1)$ correlation (0.8)}
\begin{tabular}{l|ccc|ccc|}
\hline
Estimate&\multicolumn{6}{|c|}{full sampling $W$, full sampling $S^{C}$ and $S(1)$, Highy Valuable Surrogate}\\
&\multicolumn{3}{|c|}{Parametric EML}&\multicolumn{3}{|c|}{Semi-parametric EML}\\
&Time Ind&TE wane&Both wane&Time Ind&TE wane&Both wane\\
\hline
$\widehat{STG}(2.5)$&1.76 & 0.84 & 1.73 & 1.52 & 1.06 & 1.73  \\ 
$\widehat{ptg}(2.5|0.75, 0.05)$&0.07 & 0.11 & 0.19 & 0.07 & 0.12 & 0.19 \\ 
$\widehat{PPV}(2.5|0.75)$&0.03 & 0.09 & 0.13 & 0.08 & 0.07 & 0.14 \\ 
$\widehat{NPV}(2.5|0.05)$&-0.01 & 0.14 & 0.24 & -0.04 & 0.14 & 0.23  \\ 
\hline
\hline
Estimate&\multicolumn{6}{|c|}{full sampling $W$, 5:1 $S^{C}$ and $S(1)$, Highy Valuable Surrogate}\\
&\multicolumn{3}{|c|}{Parametric EML}&\multicolumn{3}{|c|}{Semi-parametric EML}\\
&Time Ind&TE wane&Both wane&Time Ind&TE wane&Both wane\\
\hline
$\widehat{STG}(2.5)$&2.28 & 0.84 & 2.55 & 2.12 & 1.30 & 2.03  \\ 
$\widehat{ptg}(2.5|0.75, 0.05)$&  0.08 & 0.11 & 0.27 & 0.10 & 0.13 & 0.20  \\ 
$\widehat{PPV}(2.5|0.75)$&  0.02 & 0.09 & 0.20 & 0.10 & 0.08 & 0.15 \\ 
$\widehat{NPV}(2.5|005)$&  0.11 & 0.14 & 0.29 & 0.05 & 0.23 & 0.35 \\ 
\hline
\multicolumn{7}{p{9in}}{Average bias (absolute bias times 100) of summary statistic estimates over 1000 simulated experiments. All estimates in all scenarios have average bias less than 3\% of the Monte Carlo standard error.}
\end{tabular}
\end{sidewaystable}

\begin{figure}[H]
\begin{center}
%\includegraphics[scale=1.75, angle=270, clip=TRUE, trim=0 0 20mm 0]{4Panel_trueTE.pdf}
\includegraphics[scale=0.65]{dcct-figure-supplement-new-11-18-2013.pdf}
\end{center}
\caption{The A panels depict the estimated $CDF^{6.5}(c)$ versus $c$ for the truncated univariate candidate principal surrogates HBA1C and EGFR as well as for the combination candidate PS for given time points. The B panels depict the estimated $ROC^{6.5}(q)$ versus $q$ for HBA1C and the combination of HBA1C and EGFR at given time points. \label{exp}}
\end{figure}

\bibliographystyle{plainnat}
\bibliography{bib_dissertation}

\end{document}

### 2014-07-18 power_table.R
Just what you need

@@ -1,27 +1,0 @@
-install.packages('xtable')
-library(xtable)
-
-ISTG<-read.csv('power-table-integrated-STG-erin.csv')
-
-
-ISTG$X<-c(rep(c(0.5, 0.8), 3), 0,0,0,0)
-ISTG$X[7]<-ISTG$X[9]<-paste('(0.5,0.5)')
-ISTG$X[8]<-ISTG$X[10]<-paste('(0.8,0.8)')
-
-
-
-xtable(ISTG)
-
-
-rownames(ISTG)<-c('$\\widetilde{STG}=0.10$', '2', '$\\widetilde{STG}=0.25$', '4',
-'$\\widetilde{STG}=0.47$', '6', '$\\widetilde{STG}^{*}=0.23$', '8',  '$\\widetilde{STG}^{*}=0.52$', '10')
-
-
-
-
-
-
-
-
-
- 

### 2014-09-02 Compare_090214_statsmed_revisions.tex
final revisions

--- 2014-07-18 Compare_053014_statsmed_revisions.tex
@@
 Much of the literature focuses on the evaluation of univariate biomarkers as principal surrogates for a binary clinical endpoint.
-\citet{Gabriel13} extended methods for evaluating univariate a biomarker as a PS of a continuous time-to-event clinical endpoints.
+\citet{Gabriel13} extended methods for evaluating a univariate biomarker as a PS of continuous time-to-event clinical endpoints.
 \citet{Huang11} introduced a semi-parametric EML method to accommodate evaluation of multivariate candidate surrogates for binary clinical endpoints.
@@
 This would not be the case if the risks were hazard-based as the subjects still at risk at time $t$, $T(z)\geq t$, would differ between the treatment arms \citep{Hernan10, Gabriel13}.
-For this reason, we focus on the CDF based marginal risks, which provide well defined  interpretations for our suggested summary statistics.
+For this reason, we focus on the CDF based marginal risks, which provide well-defined  interpretations for our suggested summary statistics.
 
@@
 
-When A6 is observably violated the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased.
+When A6 is observably violated, the suggested estimators $FPF(t|c)$ and $TPF(t|c)$ are biased.
 However, under minor violations of A6 that might not be observable in the sample, the estimators are still reasonable estimates of useful quantities.
@@
 This interpretation follows directly from the proof given in \citet{Huang11} for the binary clinical endpoint setting.
-Given this interpretation, $STG(t)$ at a fixed time point is a useful single measurement for evaluation of univariate and multivariate candidate surrogates, as this is equivalent to the upper-most left point on the $ROC^{t}(q)$ curve for the same time point $t$.
+Given this interpretation, $STG(t)$ at a fixed time point is a useful single measurement for evaluation of univariate and multivariate candidate surrogates, as this is equivalent to the upper-left-most point on the $ROC^{t}(q)$ curve for the same time point $t$.
 
@@
 This curve can be used to investigate time-variation in surrogate quality for a particular surrogate or for comparison.
-Curves, $STG(t)$ versus $t$, that have noticeable downward slopes may indicate that a candidate surrogate modifies the clinical treatment effect less over time, declining surrogate quality, or that the marginal risk difference is declining.
+Curves, $STG(t)$ versus $t$, that have noticeable downward trends may indicate that a candidate surrogate modifies the clinical treatment effect less over time, declining surrogate quality, or that the marginal risk difference is declining.
 To better understand the context of declining or increasing $STG(t)$ over time, the marginal risk differences over time should be graphed simultaneously.
@@
 \]
-The $\widetilde{STG}$ can be used to summarize the surrogate quality without specifying a fixed time point.
-By borrowing information across times, it provides a useful, single value estimate for describing and comparing candidate surrogates.
+By integrating over the Kaplan-Meier estimate, the $\widetilde{STG}$ statistic weights time points with observed events more heavily.
+This makes $\widetilde{STG}$ a sensible and empirically based summary of surrogate quality over all time-points of interest.
+The integrated $STG$ can be used to summarize the surrogate quality without specifying a fixed time point because it borrows information across times.
+It provides a useful, single value estimate for describing and comparing candidate surrogates.
 The $\widetilde{STG}$ difference between the candidate PS in the same trial can be used in the same way as the difference in $STG(t)$ for inference on comparison.
@@
 
+\ack The author are grateful to Dean Follmann and Ying Huang for their invaluable insights and helpful discussion.
+The authors also acknowledge both anonymous referees, whose comments helped us improve this article.
+
 \bibliographystyle{statmed}
